RNA Selectively Modulates Activity of Virulent Amyloid PSMα3 and Host Defense LL-37 via Phase Separation and Aggregation Dynamics

Version 1

Abstract

Amyloids, classically linked to neurodegenerative diseases, also play critical roles in infection and immunity. Phenol-soluble modulins (PSMs) from Staphylococcus aureus are virulent amyloids that contribute to cytotoxicity, immune modulation, and biofilm stability. PSMα3 forms cross-α amyloid fibrils and shares sequence and structural features with LL-37, a human host-defense peptide that assembles into α-helical structures. Here, we uncover RNA as a potent, context-dependent modulator of their aggregation and activity. RNA consistently reduces LL-37’s cytotoxicity toward human cells without compromising its antibacterial function, suggesting a selective host-protective mechanism. In contrast, RNA preserves PSMα3’s cytotoxic and antimicrobial activity over time, likely by promoting liquid–liquid phase separation (LLPS) and stabilizing bioactive fibrillar polymorphs, enabling S. aureus to fine-tune its virulence strategies. At higher concentrations, RNA drives both peptides toward distinct aggregated states, amorphous for LL-37 and fibrillar for PSMα3, underlying their divergent functional outcomes. The amyloid inhibitor EGCG abolishes the bioactivity of both PSMα3 and LL-37, overriding RNA’s modulatory effects. Together, our findings establish RNA as a key modulator of both virulent amyloids and host-defense peptides, with broad implications for microbial immune evasion, innate immunity, and amyloid-associated diseases. Moreover, they highlight phase transitions as a tunable mechanism for regulating peptide bioactivity and a promising therapeutic target across infectious and neurodegenerative conditions.

Introduction

Bacterial amyloids are proteins capable of forming ordered fibrils, playing a crucial role in biofilm stability, surface adherence, cell toxicity and immune responses1–8. These amyloids are vital for the survival and virulence of many bacteria, enabling them to colonize and persist in diverse environments9–11. Some examples of bacterial amyloids include the Csg and Fap proteins from Enterobacteria and Pseudomonas species2,7,9,12, respectively, which play a role in biofilm stability. The phenol-soluble modulins (PSMs) produced by Staphylococci contain amyloid-forming short peptides which play multiple functional roles central to the pathogenicity and immune evasion strategies of S. aureus13–21 .

Among bacterial amyloids, PSMα3 stands out as highly cytotoxic against a broad range of microbial and human cell types22,23. It adopts a unique cross-α amyloid structure, where α-helices are stacked perpendicular to the fibril axis in paired sheets22–24. The formation of cross-α amyloids has been correlated with toxic activity, and the properties of these fibrils may regulate PSMα3’s various functions3,25,26. PSMα3 shares self-assembly of α-helices with LL-37, a human host defense peptide27. Moreover, the LL-37 antimicrobial active core (residues 17–29) shows sequence similarity to PSMα3, and the formation of fibrils composed of densely packed cationic amphipathic α-helices28. The similarity supported the hypothesis that PSMα3 might recapitulate functions of LL-37, and indeed, they share the activation of the signal inhibitory receptor on leukocytes-1 (SIRL-1) which is a negative regulator of myeloid cell function and dampens antimicrobial responses29. Such common functions might be related to molecular mimicry and the ability of bacterial peptides to hijack immunomodulation pathways30. Indeed, PSMαs interact with the formyl peptide receptor 2 (FPR2) on immune cells, initiating immune responses such as chemotaxis, phagocytosis, and the secretion of pro-inflammatory cytokines. These reactions are critical for recruiting and activating immune cells, thereby amplifying the inflammatory response15,31–36.

LL-37 has been shown to bind RNA released from dying cells, forming stable complexes that protect the RNA from enzymatic degradation and promote its uptake by dendritic cells37. Once internalized, these RNA–LL-37 complexes activate Toll-like receptors (TLRs), triggering cytokine production and dendritic cell maturation37. Beyond LL-37, other human antimicrobial peptides (AMPs), such as β-defensins, also form higher-order assemblies that organize nucleic acids into nanostructures with defined periodicity. These nanocomplexes potently stimulate TLRs, highlighting that innate immune activation is influenced not only by ligand identity, but also by the nanoscale geometry and spatial organization of the complexes38–40.

Interestingly, nucleic acid interactions also play a key role in the immune activation triggered by bacterial amyloids. For example, PSMs co-assemble with extracellular DNA (eDNA), and CpG-rich DNA specifically promotes PSMα3 fibrillation, forming co-aggregates that colocalize with DNA in biofilms. These complexes activate TLR pathways and induce anti-dsDNA autoantibody responses in mice41. Similar mechanisms are seen in E. coli and Salmonella, where curli fibers bind eDNA, enhance fibrillation, and engage TLRs signaling, also promoting autoimmunity42,43. These findings suggest that amyloid–DNA complexes in bacterial biofilms are not only structural components but also potent immunomodulators with potential links to autoimmunity.

The similarities between LL-37 and PSMα3 prompted us to investigate their interaction with RNA and the resulting structural and functional implications on cell toxicity. Our findings show that both PSMα3 and LL-37 can undergo liquid-liquid phase separation (LLPS) in the presence of RNA, and transition into solid aggregates as a function of incubation time and RNA concentration. LLPS is in general a critical process in cellular organization in which proteins or other molecules separate into distinct liquid phases, forming dynamic, membraneless compartments central to cellular processes such as metabolism, RNA regulation, and stress response44–47. PSMα3 also penetrates HeLa cells, localizes within the nucleolus, and colocalizes with nucleic acids. We observed dynamic and reversible molecular behavior within the nucleolus, mirroring in vitro observations of PSMα3’s LLPS with RNA.

Epigallocatechin gallate (EGCG), a polyphenol primarily found in green tea and recognized as an amyloid inhibitor4,48–50, disrupts the cytotoxicity and antibacterial activity of both PSMα3 and LL-37. In contrast, RNA differentially modulates their activity, suggesting distinct mechanisms linked to their self-assembly dynamics, morphologies, and toxic pathways. These findings provide new insights into RNA-mediated self-assembly in bacterial physiology and highlight potential biomedical applications for combating resistant S. aureus infections.

Results

PSMα3 interacts with RNA in vitro

The interactions between PSMα3 and RNA were assessed using an Electrophoretic Mobility Shift Assay (EMSA), comparing the binding affinity of PSMα3 to single-stranded PolyA RNA and double-stranded Poly (AU) RNA. The RNA molecules were labeled with the IR800CW fluorescent dye (as detailed in the Methods section). Freshly dissolved PSMα3, at varying concentrations, was incubated with each RNA type at a constant concentration for 30 minutes at 37°C. The EMSA results revealed that PSMα3 exhibits a stronger binding affinity to double-stranded Poly (AU) RNA, as evidenced by significant shifts at lower peptide concentrations compared to single-stranded Poly(A) (Fig. 1A).

The binding and effect of single- and double-stranded RNA on PSMα3.

(A) EMSA assay illustrating the interaction between PSMα3 and RNA. The assay compares the effects of increasing concentrations of PSMα3 (0 µM, 40 µM, 80 µM, 160 µM, and 320 µM), shown in a gradient from left to right, on the mobility shift of single-stranded Poly(A) RNA (left panel) and double-stranded Poly (AU) RNA (right panel) at ∼400ng/µL. (B) Transmission electron micrographs of 100 µM PSMα3 incubated for 2 hours in the presence of 400 ng/µL of either Poly(A) RNA (left) or Poly (AU) RNA (right). Scale bars represent 500 nm.

A high density of PSMα3 fibrils was observed after 2 hours of incubation with Poly (AU) RNA using transmission electron microscopy (TEM), while incubation with Poly(A) RNA under the same conditions resulted in amorphous aggregation (Fig. 1B). Consequently, we decided to focus our subsequent investigations on the behavior of PSMα3 in the presence of double-stranded Poly (AU) RNA.

Poly (AU) RNA concentration-dependent induced transitioning of PSMα3 condensates to solid aggregates

To better understand the nature of the PSMα3-RNA interaction, we visualized their mixtures via fluorescence microscopy, using PSMα3 labeled with Fluorescein isothiocyanate (FITC) at the C-terminus (PSMα3-FITC), which is known to form fibrils22. 100 µM PSMα3-FITC was incubated with either 50 ng/µl or 400 ng/µl propidium iodide (PI)-labeled RNA in a close to physiological buffer of 50 mM HEPES (pH 7.4) containing 150 mM sodium chloride (NaCl).

With the addition of 50 ng/µl RNA, PSMα3 formed liquid-liquid phase separated droplets that clearly colocalized with the RNA (Fig. 2A). The liquid nature of the droplets was further validated by fluorescence recovery after photobleaching (FRAP) analysis, where we observed a rapid recovery of the PSMα3-FITC signal shortly after bleaching (Fig. 2C). However, after a longer incubation of 2 hours, no fluorescence recovery was detected, suggesting that the condensates aged and transitioned to a more aggregated state (Fig. 2D). In addition, we observed the encapsulation of RNA within PSMα3-FITC “coated” droplets (Fig. S1).

Colocalization, droplet formation and texture of PSMα3 mixed with varying Poly (AU) RNA concentrations.

(A-B) Widefield fluorescence microscopy images of 100 µM PSMα3-FITC (green) in the presence of 50 ng/µL (A) or 400 ng/µL (B) Poly (AU) RNA (red), showing their individual and composite channel images. (C-E) FRAP analysis of PSMα3-FITC with 50 ng/µL Poly (AU) RNA taken after 10 min (C) or 2-hours co-incubation (D), or with 400 ng/µL Poly (AU) RNA taken after 10 min co-incubation (E). The images in panels C-D were captured before bleaching, immediately after bleaching, and 40 seconds post-bleaching. Scale bars in all images represent 20 µm.

In contrast to the effect of 50 ng/µl RNA, the addition of a higher RNA concentration of 400 ng/µl to PSMα3 resulted in immediate aggregation, with no observed phase of droplet formation (Fig. 2B). FRAP analysis indicated no recovery of signal after bleaching, likely due to the formation of solid structures rather than droplets with liquid-like properties (Fig. 2E). These aggregates also showed colocalization (Fig. 2B), indicating that the PSMα3-RNA interaction persisted in both the LLPS and solid states.

Turbidity assays showed that a 30 min co-incubation of 100 µM PSMα3 with varying concentrations of RNA displayed an increased turbidity up to 50 ng/µl, while higher RNA concentrations displayed a decrease in turbidity (Fig. S2). This suggests that 50 ng/µl can be a critical concentration between soluble species, which might be able to phase separate, and the formation of aggregates. The decrease in particle size above 50 ng/µL can be a result of either the formation of smaller particles or aggregation into larger clusters, which can settle out of suspension, removing them from the light-scattering medium and reducing turbidity.

Alternatively, it may indicate the dissolution of phase-separated droplets, as observed in systems like Ddx451. We then used transmission electron microscopy (TEM) and total internal reflection fluorescence (TIRF) microscopy to better visualize the effect of RNA on PSMα3 morphology.

RNA concentration-driven changes in PSMα3 fibrillar morphology and aggregation

The TEM micrographs showed that in the absence of RNA, 100 µM PSMα3 formed nanotube-like fibrils after 2 hours, growing wider or with some twist after 24-hours co-incubation (Fig. 3A). Incubating PSMα3 with a low concentration of 10 ng/µL RNA displayed accelerated fibril formation, with more strongly twisted, wide, sheet-like fibrils observed after both 2 hours and 24 hours (Fig. 3A). With the addition of 50 ng/µL RNA, PSMα3 fibrils have a similar twisted morphology after 2 hours, while at a longer incubation time of 24 hours, we observed a significant morphological shift into more thin amorphous aggregates (Fig. 3A). At the higher concentration of 400 ng/µL RNA, PSMα3 formed dense, thin fibrils after 2 hours, but with a possible fragmentation into smaller species or rearrangement into amorphous aggregates after 24 hours of co-incubation. This indicates that RNA concentration and time of co-incubation affect the density and morphology of PSMα3 fibrils. This corresponds to the differences observed by light microscopy of co-aggregates contexture of LLPS droplets vs solid aggregates (Fig. 2).

TEM and TIRF visualization of PSMα3 aggregation and morphology with different Poly (AU) RNA concentrations and incubation times.

(A) TEM micrographs of 100 µM PSMα3 incubated with or without Poly (AU) RNA at varying concentrations of 10 ng/µL, 50 ng/µL, and 400 ng/µL for 2 hours (top row) and 24 hours (bottom row). Scale bars represent 500 nm. (B) TIRF microscopy images showing 100 µM FITC-PSMα3 (green) co-incubated with 50 ng/µL Poly (AU) RNA and the amyloid indicator AT630 (magenta) for 30 minutes and 2 hours. Scale bars represent 20 μm. (C) TIRF microscopy images of 100 µM FITC-PSMα3 (green) co-incubated with 400 ng/µL Poly (AU) RNA and AT630 (magenta) for 30 min. Scale bars represent 20 µm.

These observations were further supported by TIRF microscopy using the amyloid-specific dye AmyTracker630 (AT630). TIRF analysis was performed at 50 ng/µL and 400 ng/µL RNA, which induced phase separation and solid aggregation of PSMα3, respectively. With the addition of 50 ng/µL RNA to PSMα3-FITC, a strong AT630 fluorescence signal was detected after 2 hours co-incubation, but not after 30 minutes. This suggests a time-dependent transition into amyloid-like species (Fig. 3B). Conversely, at 400 ng/µL RNA, a significant AT630 fluorescence was observed already after 30 minutes of co-incubation, consistent with a rapid formation of amyloid fibrils (Fig. 3C). These findings highlight a concentration- and time-dependent modulation of PSMα3 phase separation and structural transitions, where RNA promotes LLPS at lower concentrations and drives rapid amyloid formation and unique fibrillar morphologies at higher concentrations.

Solid-state circular dichroism (ssCD) spectroscopy reveals that PSMα3 lacks a defined secondary structure both immediately after preparation and following 2 hours of incubation. However, upon addition of RNA, a clear shift toward an α-helical conformation is observed (Fig. S3). This supports the notion that RNA not only accelerates aggregation but also promotes or stabilizes the α-helical fibrillar architecture, potentially consistent with cross-α amyloid structures.

RNA modulates the antibacterial activity of PSMα3

The antimicrobial activity of PSMα3 against Escherichia coli was evaluated under varying RNA concentrations and incubation times using the PrestoBlue Cell Viability Assay (Fig. 4A).

Impact of Poly (AU) RNA on PSMα3 cytotoxicity and antibacterial activity.

Antimicrobial activity of PSMα3 against E. Coli using the PrestoBlue cell viability assay (A) and its cytotoxicity against HeLa cells using the LDH colorimetric assay (B) were assessed with and without Poly (AU) RNA at varying concentrations. The experiments were performed in at least three replicates and repeated across three independent days to ensure result reliability. Cytotoxicity and bacterial cell viability percentages were calculated as the mean of all replicates, with error bars representing the standard error of the mean (SEM). Statistical significance was determined using one-way ANOVA for normally distributed data in GraphPad Prism (version 11). Significance levels are indicated as follows: *p<0.05, **p<0.01, ***p<0.001, ****p<0.0001.

Freshly dissolved 10 µM PSMα3 exhibited potent antibacterial activity, completely abolishing bacterial viability. After 2 hours of incubation, PSMα3 retained its full activity, comparable to its freshly prepared state. The presence of RNA at 50 ng/µL and 400 ng/µL had no impact on its antimicrobial function within this timeframe (Fig. 4A).

In contrast, after 24 hours of incubation, PSMα3’s antibacterial activity was significantly reduced (Fig. 4A), also at 20 µM, suggesting a decrease in its effective concentration or changes in its morphology. Notably, the addition of RNA at 50 ng/µL or 400 ng/µL prevented this loss of activity, indicating that RNA plays a stabilizing role of toxic species. These findings suggest that RNA influences PSMα3 aggregation dynamics and morphology, thereby modulating its long-term antimicrobial effectiveness.

RNA modulates PSMα3 cytotoxicity against human HeLa cells

The effect of RNA on the cytotoxicity of PSMα3 against human HeLa cells was evaluated by measuring lactate dehydrogenase (LDH) release, an indicator of cell membrane damage (Fig. 4B). Freshly dissolved PSMα3 exhibited substantial cytotoxicity, causing approximately 80% cell death at 20 µM and 50% at 10 µM. The presence of RNA did not significantly alter the toxicity of freshly dissolved PSMα3 (Fig. 4B), similar to the antibacterial activity. However, following 2 hours of incubation, PSMα3 cytotoxicity was significantly reduced, possibly due to aggregation and a loss of active peptide concentration. This reduction occurred earlier than the loss of antibacterial activity, which was only observed after 24 hours of incubation (Fig. 4A). This discrepancy may reflect differences in how membrane composition and media conditions influence PSMα3 aggregation kinetics, or it may indicate distinct mechanisms underlying cytotoxicity versus antibacterial membrane disruption.

Notably, co-incubation with RNA helped maintain PSMα3 cytotoxicity in a concentration-dependent manner. With 50 ng/µL RNA, the 2-hour incubated PSMα3 retained cytotoxicity comparable to its freshly dissolved form, suggesting that RNA prevents activity loss due to incubation. However, after 24 hours, 50 ng/µL RNA was insufficient to preserve cytotoxicity.

In contrast, at 400 ng/µL RNA, partial cytotoxicity was maintained for both 2-hour and 24-hour incubated PSMα3, compared to the freshly dissolved sample.

Overall, these findings suggest that RNA prevents the incubation-induced loss of both antibacterial activity and cytotoxicity (Fig. 4). This effect appears to be concentration- and co-incubation time-dependent, likely linked to RNA-induced morphological variations of PSMα3 species (Figs. 2&3).

PSMα3 targets nucleolar nucleic acids in HeLa cells with LLPS dynamics

The interactions and localization of PSMα3 were next explored in HeLa cells with the addition of 20µM FITC-PSMα3 just before imaging. PI was added to the cell medium as a marker for cell death, as it selectively penetrates dead cells and binds to their nucleic acids.

Confocal microscopy images revealed that PSMα3 induces significant toxicity as indicated by membrane damage, intracellular PSMα3 aggregate formation, and positive PI staining. PSMα3 also enters the nuclei, with a notable concentration in the nucleolus, indicated by the foci of green fluorescence (marked by white arrows in Fig. 5). The composite images demonstrated clear colocalization of PSMα3 with nucleic acids in the nucleolus, as seen by the overlap of the green PSMα3-FITC signal and the red fluorescence from PI-stained nucleic acids.

Colocalization of PSMα3 with nucleic acids in HeLa cells.

(A) Confocal microscopy images showing the localization and colocalization of 20 µM PSMα3-FITC (green) and nucleic acids stained with PI (red) inside the nucleolus of HeLa cells (indicated in arrows). The left panel illustrates the distribution of PSMα3 within the cell. The middle panel shows the nucleic acids stained with PI. The right panel is a composite image that demonstrates the colocalization of PSMα3 with nucleic acids. Scale bars represent 15 µm. (B) FRAP analysis of 20 µM FITC-labeled PSMα3 inside the nucleolus (indicated by the arrow) of HeLa cells, showing fluorescence recovery after 60 seconds. Scale bars represent 10 µm.

FRAP analysis of 20 µM PSMα3-FITC within the nucleolus of HeLa cells indicated mobility (Fig. 5B). The Pre-bleach panel shows the initial fluorescence distribution, with the nucleolus marked by the white arrow. Following photobleaching, a localized reduction in fluorescence was observed. Notably, within 60 seconds, significant fluorescence recovery was observed in the nucleolus, highlighting the dynamic and reversible molecular behavior of PSMα3 in this cellular compartment. The results overall suggest an interaction between PSMα3 and nucleolar RNA/DNA within cells, which mirrors our in vitro findings where PSMα3 interacted, and underwent LLPS and formed solid aggregates in the presence of RNA.

EGCG directly binds PSMα3 and inhibits its fibrillation and bioactivity, even in the presence of RNA

Since RNA appears to protect or maintain PSMα3’s toxic functions, potentially by influencing its fibril formation and morphology, we investigated the corresponding and combined effects of an inhibitor of fibril formation. One such inhibitor is epigallocatechin gallate (EGCG), the most abundant catechin in tea and a known amyloid inhibitor, including of PSMs4.

The addition of EGCG to PSMα3 at a 1:1 molar ratio did not significantly alter its cytotoxicity against HeLa cells. However, at a fivefold molar excess, EGCG completely abolished PSMα3’s cytotoxic effect (Fig. 6C). To examine the structural changes underlying this effect, we analyzed fibril formation kinetics and morphology in the presence of EGCG. TEM micrographs revealed that a fivefold molar excess of EGCG disrupted fibril formation of 100 µM PSMα3, instead inducing amorphous aggregates (Fig. 6A). Consistently, kinetics assays of fibril formation showed that EGCG inhibited the Thioflavin-T (ThT) fluorescence curve otherwise indicating the fibril formation of 100 µM PSMα3 (Fig. S4). These findings suggest that EGCG reduces PSMα3 toxicity by modulating its fibril formation and morphology. The addition of RNA did not counteract the effect of EGCG or restore PSMα3 cytotoxicity (Fig. 6C), highlighting the strong impact of EGCG on PSMα3 morphology and properties.

EGCG modulates PSMα3 aggregation and reduces toxicity against HeLa cells.

(A) TEM micrographs of 100 μM PSMα3 incubated for 24 hours, without (left) and with (right) a fivefold molar excess of EGCG. Scale bars: 500 nm. (B) Live-cell confocal microscopy of HeLa cells treated with 20 μM PSMα3-FITC (green) without (top) and with EGCG (bottom), imaged immediately after preparation (t = 0) and after 10 minutes. Hoechst 33342 (blue) marks the nuclei, while PI staining (red) indicates membrane disruption. Scale bars: 10 μm. (C) Cytotoxicity of PSMα3 against HeLa cells in the presence and absence of EGCG and Poly (AU) RNA at two different concentrations, assessed via LDH assay. The experiment was performed in triplicate and repeated on three separate days for consistency. Cytotoxicity percentages were averaged across all replicates, with error bars representing the standard error of the mean. Statistical significance: *p < 0.05, **p < 0.01, ***p < 0.001, ****p < 0.0001 (one-way ANOVA, GraphPad Prism v11).

To investigate residue-specific interactions between PSMα3 and EGCG, we performed NMR spectroscopy using a 2:1 PSMα3:EGCG molar ratio. The 1D 1H-NMR spectrum revealed distinct chemical shift changes and peak broadening upon EGCG addition, particularly in residues Glu8/Met1 and Glu2, suggesting specific interactions between these sites and the EGCG molecule (Fig. 7A). Peak broadening was also observed in the aromatic region of EGCG, especially for protons H1 and H2, compared to the reference spectrum of EGCG alone, indicating that the interaction is visible from the EGCG side as well. Notably, slight opalescence was observed in the sample following preparation, potentially reflecting early-stage aggregate formation. Additionally, the presence of multiple cross-peaks between non-sequential residues (i+3 or i+4) in the two-dimensional (2D) 1H-1H Nuclear Overhauser Effect Spectroscopy (NOESY) spectrum indicates a well-defined structure of PSMα3 in this condition (Fig. 7B). The 2D 1H-1H Total Correlation Spectroscopy (TOCSY) spectrum displayed connectivity between backbone amide (HN) and alpha protons (Hα) for several assigned residues, while the 2D 1H-1H NOESY spectrum revealed spatial correlations between nearby residues (i+1).

Residue-specific interactions between PSMα3 and EGCG.

(A) One-dimensional (1D) 1H NMR spectra of 1.0 mM PSMα3 alone (red) and in complex with 0.5 mM EGCG (blue), recorded at 35 °C. Specific residues, including Glu8/Met1 and Glu2, show chemical shift changes suggestive of direct interaction with EGCG (highlighted in the upper left). Dashed boxes mark proton signals corresponding to EGCG. Peak broadening of H1/H2 protons, compared to the EGCG-only reference sample (light gray), indicates interaction from the EGCG side. Slight opalescence observed in the sample suggests potential aggregate formation. (B) Two-dimensional (2D) 1H–1H TOCSY and NOESY spectra of the PSMα3:EGCG complex at a 2:1 ratio, recorded at 35 °C. The cross-peaks in the TOCSY spectrum allow the identification of the spin system and direct connections through the scalar coupling between the proton amide (HN) and the alpha protons (Hα) of the same residue. The cross-peaks in the NOESY spectrum establish a sequential connection between neighboring residues (i+1), allowing spectral assignment. The remaining cross-peaks (i+3 or i+4) in the spectrum support PSMα3 secondary structure in the experimental conditions. (C) Temporal stability of the PSMα3:EGCG sample over 3 days.

To assess the temporal stability of the sample, we monitored signal intensities over a 3-day period. PSMα3 signals remained stable, with only a ∼10% decrease in intensity, whereas EGCG signals exhibited substantial degradation, with nearly 50% loss over the same time frame (Fig. 7C). Given the ∼1.5-day duration of the 2D NMR measurements, it is estimated that over 60% of EGCG remained in solution during data acquisition, allowing for reliable observation of its interaction with the peptide. Of note, EGCG’s activity has been shown to depend on its chemical stability and the surrounding conditions. Specifically, at neutral pH, EGCG may undergo oxidation, and its inhibitory effects could be attributed to its degradation products rather than the intact compound itself52.

Live-cell confocal microscopy provided further insights into how EGCG binding affect PSMα3 interactions with cells. A fivefold molar excess of EGCG prevented the toxic effects of 20 µM PSMα3, blocking cell penetration and PI staining (Fig. 6B, top panel). Instead, PSMα3-FITC formed extracellular aggregates and showed no interaction with cell membranes, as indicated by the absence of PI staining and intact cell membranes (Fig. 6B, bottom panel).

Furthermore, EGCG at a fivefold molar excess also reduced the antibacterial activity of PSMα3 against E. coli, maintaining bacterial cell viability (Fig. S5). Super-resolution fluorescence microscopy showed that while 20 µM PSMα3 typically aggregates on the bacterial membrane, causing membrane disruption and PI staining indicative of cell death (Fig. S5), EGCG at a fivefold molar excess prevented membrane aggregation and PI staining, thereby preserving membrane integrity.

RNA modulates the human host defense peptide LL-37 phase behavior and cytotoxicity without compromising antimicrobial activity

LL-37, a human host defense peptide, shares functional, sequence, and structural similarities with PSMα3, including its ability to self-assemble and interact with nucleic acids28,37. We therefore examined whether RNA similarly influences LL-37 activity and properties.

Fluorescence microscopy showed that 100 µM LL-37-FITC incubated with 100 ng/µL PI-labeled Poly (AU) RNA displayed formation of aggregates, but with no indication for LLPS or droplet formation (Fig. S6). Since diverse stresses trigger phase separation of RNA-binding proteins and their coalescence into stress granules53 , we explored the effect of RNA on LL-37 under thermal stress. We applied heat shock at 65°C for 15 minutes and found that RNA significantly affected LL-37’s phase behavior in a concentration-dependent manner. Specifically, with the addition of 100 ng/µL RNA, LL-37 formed condensates that colocalized with the RNA (Fig. 8A). At 200 ng/µL RNA, LL-37 still formed droplets, but aggregation began to appear, indicating a transition from a liquid-like state to a more solid-like phase (Fig. 8B). At 400 ng/µL RNA, droplets were no longer observed, and extensive aggregation occurred, with strong colocalization between LL-37 and RNA (Fig. 8C). As a comparison of stress response, PSMα3 displayed droplet formation after the heat shock of 65°C for 15 minutes only at more conducive conditions of pH 4, but not at pH 7.4 (Figs. S7-S8). The results overall suggest that for both LL-37 and PSMα3, higher RNA concentrations promote aggregation over phase separation.

Effect of RNA concentration on LL37 phase separation, aggregation, and activity at pH 7 after heat shock.

Fluorescence microscopy images showing LL37 (FITC-labeled, green) in the presence of increasing concentrations: 100 ng/μL (A), 200 ng/μL (B) and 400 ng/μL (C) of Poly (AU) RNA (PI-labeled, red) after heat shock at 65°C for 15 minutes at pH 7. Scale bars represent 20 µm. (D) LL37 cytotoxicity, with and without Poly (AU) RNA and EGCG at varying concentrations, was assessed in HeLa cells using the LDH colorimetric assay. The experiment was performed in triplicate and repeated on three separate days for reproducibility. Cytotoxicity percentages represent the average of all replicates, with error bars indicating the standard error of the mean. (E) The antimicrobial activity of LL37, with and without Poly (AU) RNA and EGCG at varying concentrations, was evaluated against E. coli using the PrestoBlue Cell Viability assay. The experiment was performed in triplicate and repeated on three separate days for reproducibility. Bacterial viability percentages represent the average of all replicates, with error bars indicating the standard error of the mean. Statistical significance (D-E): *p < 0.05, **p < 0.01, ***p < 0.001, ****p < 0.0001 (one-way ANOVA, GraphPad Prism v11).

Freshly dissolved LL-37 exhibited cytotoxicity against HeLa cells, which, in contrast to PSMα3, remained consistent even after incubation for up to 24 hours. Notably, RNA significantly reduced LL-37 cytotoxicity in a concentration-dependent manner, irrespective of the incubation duration (freshly dissolved, 2 hours, or 24 hours) (Fig. 8D). This contrasts with RNA’s rescuing effect on the incubation time-dependent loss of PSMα3 cytotoxicity. The antimicrobial activity of LL-37 against E. coli also remained unchanged upon incubation. However, unlike its effect on cytotoxicity, RNA had minimal impact on antibacterial activity, with only high concentrations of 400 ng/µL reducing LL-37 activity (Fig. 8E). Similar to PSMα3, LL-37 cytotoxicity against HeLa cells and its antimicrobial activity was attenuated by EGCG in a concentration-dependent manner (Fig. 8E).

TEM micrographs of 100 µM LL-37 revealed a variety of aggregative and fibrous structures, including thin, curli-like fibrils, although no consistent morphology was observed (Fig. S9A). RNA induced a distinct morphological change, leading to increased aggregation and the formation of mostly amorphous species, but also some thicker, ribbon-like fibrils were observed. These structures resemble those previously observed for a segment of the LL-37 active core (residues 17–29), which is similar to PSMα3 in sequence and its ability to form fibrils of densely packed amphipathic α-helices28. Heat shock further intensified aggregation, resulting in denser, amorphous condensates and larger fibrillar assemblies, with and without RNA (Fig. S9B). The distinct effects of RNA on the aggregation and morphologies of LL-37 and PSMα3 may underlie the observed differences in their cytotoxic and antibacterial activities.

Discussion

This study demonstrates that Poly (AU) RNA modulates the cytotoxic and antimicrobial activities of PSMα3 and LL-37 in a concentration-dependent and peptide-specific manner, revealing a regulatory mechanism that differentially influences their biological functions. The effects of RNA on these peptides are distinctly linked to their aggregation dynamics and morphological outcomes, underscoring its role as a context-specific modulator of peptide bioactivity.

Interestingly, although the supramolecular structure of LL-37 does not exhibit the characteristic amyloid cross-β or the cross-α architecture, with molecular stacking perpendicular to the fibril axis27, it nonetheless shows functional links to amyloids. Notably, LL-37 has been reported to bind the Alzheimer’s-associated amyloid-β (Aβ) peptide and inhibit its fibrillization, likely by stabilizing prefibrillar intermediates54. These findings suggest that LL-37 may modulate Aβ aggregation dynamics, with potential implications for aging, immune suppression, and infection54.

Our findings demonstrate that RNA consistently reduces LL-37’s cytotoxicity against human cells, independent of RNA concentration or incubation duration (Fig. 8D). Notably, this reduction in cytotoxicity does not compromise LL-37’s antimicrobial activity against E. coli (Fig. 8E), indicating that RNA selectively protects mammalian cells while preserving its host defense functions. Unlike PSMα3, LL-37 does not undergo RNA-induced LLPS under physiological conditions, exhibiting this behavior only after thermal stress (Fig. 8). RNA rather leads to amorphous aggregates (Fig. S9), which likely underlies the selective attenuation of cytotoxicity, highlighting RNA’s role in modulating LL-37’s functional states.

In contrast to the results for LL-37, RNA exhibits a dynamic and concentration-dependent regulation of PSMα3’s cytotoxicity and antimicrobial activity, revealing a complex interplay between phase separation and aggregation. Unlike LL-37, PSMα3 shows a time-dependent reduction in activity (Fig. 4), likely due to a transition from soluble species to irreversible fibril morphology (Fig. 3). At lower RNA concentrations, PSMα3’s cytotoxic activity against human cells is initially preserved but diminishes over time. This time-dependent decrease likely reflects a transition from soluble liquid-like droplets to solid aggregates within the RNA-PSMα3 complex. Conversely, at higher RNA concentrations, PSMα3’s cytotoxicity is partially maintained even after prolonged incubation, suggesting that RNA induces a specific fibril morphology that remains reversible and bioactive.

The antimicrobial activity of PSMα3 also shows temporal modulation, with a noticeable decline after 24 hours of incubation (Fig. 4A). This delayed reduction compared to its activity against human cells suggests distinct aggregation pathways or interactions with the bacterial environment. However, RNA preserves PSMα3’s antimicrobial function even after 24 hours (Fig. 4A), likely by stabilizing intermediate or alternative polymorph states, which are reversible and enable the reactivation of PSMα3.

This behavior of dynamic phase transitions and aggregation mirrors those observed in human amyloids45,46,55–64. For example, FUS, TDP-43, hnRNPA1, and tau rely on LLPS to perform their normal cellular functions. However, these proteins are also associated with neurodegenerative diseases like amyotrophic lateral sclerosis, frontotemporal dementia, tauopathies, and Alzheimer’s disease. While LLPS enables their proper functioning under healthy conditions, mutations or modifications can disrupt this process, leading to irreversible aggregation and disease progression55–66. Similar to human amyloids, AMPs can also induce LLPS with nucleic acids67. In bacterial cells, this interaction compacts nucleic acids and disrupts critical processes like transcription and translation67. PSMα3 interaction with nucleic acids within human cells, particularly its localization in the nucleolus (Fig. 5), supports a comparable mechanism leading to cell death mediated by this bacterial virulence factor.

These findings reveal a sophisticated and context-specific regulation of PSMα3 and LL-37 by RNA, demonstrating that RNA can fine-tune peptide bioactivity with remarkable specificity across cell types. The specificity of RNA’s effects on cytotoxicity and antimicrobial activity points to a sophisticated regulatory mechanism with significant biological implications: For PSMα3, RNA-driven LLPS and aggregation dynamics provide a flexible strategy for modulating cytotoxicity and antimicrobial activity, enabling S. aureus to adapt its virulence based on environmental cues. This dynamic structural plasticity may facilitate immune evasion, biofilm formation, and enhanced survival in diverse host environments. For LL-37, RNA’s selective attenuation of cytotoxicity without compromising antibacterial function suggests an evolutionary strategy to balance host defense and tissue protection. This regulatory mechanism minimizes collateral damage to host cells while maintaining effective antimicrobial activity, highlighting dual functionality in immune regulation.

These findings reveal a previously unrecognized role for RNA in regulating amyloid-like peptides, opening new avenues for RNA-targeted therapeutic interventions against bacterial infections and immune regulation. Moreover, the parallels between RNA-driven phase transitions in PSMα3 and human amyloids suggest broader implications for understanding amyloid-associated diseases. By elucidating the dual role of RNA in promoting reversible LLPS and inducing stable fibril formation, this study contributes to the growing understanding of phase behavior in functional and pathological amyloids.

Methods

Peptide and Poly (AU) RNA preparation

Unlabeled and C-terminal FITC-labeled PSMα3 and LL-37 (PSMα3-FITC and LL-37-FITC) (>98% purity) were purchased from GL Biochem (Shanghai) Ltd. Poly (AU) RNA was purchased from Sigma Aldrich. A stock solution of the peptides was prepared at a concentration of 1 mM in a mixture of 20% dimethyl sulfoxide (DMSO) and 80% ultra-pure water. For light microscopy experiments, we used a mixture of 20% FITC-labeled and 80% unlabeled PSMα3 or LL-37. The Poly (AU) RNA stock solution was prepared at a concentration of 2000 ng/µl in UPW, the stock solution was stored at −80 °C until further use. Labeled Poly (AU) RNA was prepared by introducing 0.02 mg/ml of PI dye to the Poly (AU) RNA stock solution. For the experiments, samples containing 100 µM PSMα3 or LL-37 with Poly (AU) RNA at 10, 50, 100, 200, or 400 ng/µl were prepared in 50 mM HEPES buffer with 150 mM sodium chloride (NaCl), adjusted to pH 7.4, or in 20 μM Tris, 20 μM Bis-Tris, and 20 μM sodium acetate adjusted to pH 4.

Cytotoxicity against HeLa cells tested using the lactate dehydrogenase (LDH) release assay

Human cervical carcinoma HeLa cells (ATCC® CCL-2™) were routinely cultured in Dulbecco’s Modified Eagle’s Medium - high glucose (DMEM) (Sigma, Israel) with L-glutamine, and supplemented with penicillin (100 U/ml), streptomycin (0.1 mg/ml), and with 10% fetal calf serum (Sigma, Israel). The cells grown at 37°C and 5% CO2. One day before the experiment, cells were resuspended in growing medium (DMEM supplemented with 10% fetal calf serum) 1X105 cells/ml. 50 μL of cell suspension were pipetted into a 96-well plate and grown over night. Thirty minutes before the experiment cells were washed and resuspended in 50 μL of DMEM medium supplemented with L-glutamine (100 U/mL), and with penicillin (100 U/mL), streptomycin (0.1 mg/mL), and with 0.5% heat-inactivated fetal calf serum (assay medium). For the cytotoxicity assay, 100 µM PSMα3 was incubated with or without EGCG at molar ratios of 1:1 or 1:5, or with Poly (AU) RNA at concentrations of 50 ng/µL or 400 ng/µL, for the designated incubation times. Similarly, 100 µM LL-37 was prepared under the same conditions, with or without EGCG at 1:1 or 1:5 molar ratios, or Poly (AU) RNA at 100 ng/µL or 400 ng/µL, following the specified incubation times.

Serial two-fold dilutions in assay medium were performed, and 50 μL of each dilution were pipetted into the three different 96-well with the cells. The plates were incubated for 30 min at 37°C and 5% CO2 and then cell lysis was quantified using the LDH release colorimetric assay according to the manufacturer’s instructions, including all recommended controls (LDH; Cytotoxicity Detection Kit Plus, Roche Applied Science, Germany). Cell-free assay medium was measured as background. Cells subjected to the same experimental conditions apart from peptide addition were used as a control to account for spontaneous LDH release. Cells subjected to the same experimental conditions apart from peptide addition and treated with manufacturer-supplied lysis buffer were used as a control to account for maximum LDH release. Absorbance at 490 and 690 nm was measured in a plate reader (FLUOstar Omega, BMG Labtech, Germany). Absorbance at 690 nm was subtracted from 490 nm readings to correct for background. The mean absorbance of triplicate samples and controls was calculated, followed by background subtraction. Data was obtained from at least three independent biological replicates, with the arithmetic mean used for averaging. Error bars represent the standard deviation (SD). Statistical analysis, including one-way ANOVA, was performed using GraphPad Prism 10.

Assessment of Bacterial Viability Using PrestoBlue™ HS Cell Viability Reagent

Escherichia coli (E. coli) cultures were grown overnight in Luria Broth (LB) medium at 37 °C with shaking at 220 rpm. PSMα3 and LL-37 stock solutions were prepared by dissolving the peptides at a concentration of 1 mM in a solvent mixture of 20% DMSO and 80% ultra-pure water. From these stock solutions, the following sample preparations were made:

For PSMα3: (1)100 µM PSMα3 (2)100 µM PSMα3 with 50 ng/µL Poly (AU) RNA (3) 100 µM PSMα3 with 400 ng/µL Poly (AU) RNA (4) 100 µM PSMα3 with 500 µM EGCG. For LL-37: (1)100 µM LL-37 (2)100 µM LL-37 with 100 ng/µL Poly (AU) RNA (3) 100 µM LL-37 with 400 ng/µL Poly (AU) RNA (4) 100 µM LL-37 with 500 µM EGCG.

All solutions were prepared in 50 mM HEPES buffer pH 7.4 containing 150 mM NaCl. Where applicable, samples were incubated for the specified durations of 2 hours or 24 hours.

On the day of the experiment, dilutions of the PSMα3 solutions to 20 µM and 10 µM were prepared in LB medium and dispensed into a sterile 96-well black flat-bottom plate (Greiner bio-one). E. coli cultures were diluted to an optical density (OD) of 0.2, and bacterial suspensions were added to the wells. The plate was incubated at 37 °C with shaking at 220 rpm for 1 hour to allow the reaction to proceed.

Following incubation, 10X PrestoBlue™ HS Cell Viability Reagent (Invitrogen) was added to the wells. Wells containing only LB medium served as negative controls, while wells containing only E. coli served as positive controls. Plates were sealed with a thermal seal film (EXCEL Scientific) and incubated in a plate reader (CLARIOstar). Bacterial viability was assessed using PrestoBlue fluorescence (excitation: 535–560 nm; emission: 590–615 nm) over time. Each condition was tested in triplicate across three independent experiments. Fluorescence values were averaged, and blank readings were subtracted. Antimicrobial activity was determined at the 2-hour time point, when fluorescence was highest in the positive control. Data were obtained from at least three independent biological replicates, with the arithmetic mean used for averaging. Error bars represent the standard deviation (SD). Statistical analysis, including one-way ANOVA, was performed using GraphPad Prism 10.

Transmission electron microscopy (TEM)

For TEM analysis, 4–5 µL of 100 µM PSMα3, with or without EGCG or Poly(AU) RNA at concentrations of 10 ng/µL, 50 ng/µL, and 400 ng/µL, as well as 100 µM LL-37, with or without Poly(AU) RNA at 100 ng/µL and 400 ng/µL, were directly applied onto glow-discharged 400-mesh copper grids (easiGlow; Pelco, Clovis, CA, USA) with a grid hole size of 42 µm, stabilized with Formvar/carbon (Ted Pella, Inc.). The grids were glow-discharged using a 15-mA current with a negative charge for 25 seconds. Samples were allowed to adhere to the grids for 45 seconds before being stained with a 1% uranyl acetate solution (Electron Microscopy Science, 22400-1) for 45 seconds. The samples were then examined using a ThermoFisher Scientific (FEI) Talos F200C transmission electron microscope, operating at 200 kV and equipped with a Ceta 16M CMOS camera, at the Ilse Katz Institute for Nanoscale Science and Technology, Ben Gurion University of the Negev, Israel.

Turbidity measurements

Turbidity measurements were performed using a FLUOstar Omega plate reader (BMG Labtech) set to a wavelength of 400 nm. For each sample, the protein concentration was maintained at 100 µM, while RNA concentrations were varied across the following levels: 10 ng/µl, 20 ng/µl, 50 ng/µl, 100 ng /µl, 200 ng/µl, and 400 ng/µl. The turbidity of the mixtures was monitored over several hours following resuspension. The maximum absorbance recorded within the first 30 minutes was used as an indicator of the dense phase volume. The experiments were conducted three times with similar observations.

Electrophoretic mobility shift assay (EMSA)

RNA molecules were synthesized by IDT, with oligoA modified by the IRDye 800CW fluorescent dye. Double-stranded RNA was prepared by annealing 10 µM oligoA-IRDye 800CW with 10 µM oligoU in an annealing buffer containing 10 mM Tris-HCl (pH 7.5), 50 mM NaCl, and 1 mM EDTA. The mixture was heated to 95°C for 5 minutes, followed by gradual cooling to 25°C at a rate of 2°C per minute. The final RNA concentration used in the binding reactions was 40 nM (∼400ng/µL).

RNA-protein complexes were incubated at 37°C for 30 minutes before being resolved on a 2.5% agarose gel. Gel electrophoresis was performed in a 0.5× TBE buffer for 7 minutes. The complexes were visualized using a Li-Cor Odyssey FC imaging system in the 800 nm channel with a 30-second exposure. For the EMSA, binding reactions were carried out in a buffer composed of 20 mM HEPES (pH 7.9), 50 mM KCl, 1 mM Dithiothreitol (DTT), 0.1 mM Ethylenediaminetetraacetic acid (EDTA), 5% glycerol, and 0.05% nonyl phenoxypolyethoxylethanol (NP-40). PSMα3 was added to the reactions at final concentrations of 0, 40, 80, 160, and 320 µM. The experiments were conducted three times with similar observations.

Fluorescent microscopy imaging of peptides with RNA in vitro

Prior to imaging, 100 µM PSMα3 or LL-37 (containing 20% FITC-labeled and 80% unlabeled peptide) were combined with PI-labeled Poly (AU) RNA at different concentrations in 50 mM HEPES buffer pH 7.4 containing 150 mM NaCl. Sample were also tested after 65°C heat shock for 15 minutes. In addition, the PSMα3 and RNA mixture was also tested in a different buffer containing 20 μM Tris, 20 μM Bis-Tris, and 20 μM sodium acetate at pH 4 and were subjected to heat shock at 65°C for 15 minutes.

Following preparation, with or without heat shock, a 10 µL aliquot of each sample was transferred to a µ-Slide 8 Well ibidi chamber or to a clean 24 × 60 mm No. 1.5 glass slide for imaging. Fluorescence microscopy was performed using a Leica DMI8 inverted fluorescent microscope equipped with a 63× immersion oil objective (Numerical Aperture, NA = 1.4) at the Life Sciences and Engineering (LS&E) Infrastructure Center, Technion-Israel Institute of Technology, Haifa, Israel. Data were processed and analyzed using Fiji-ImageJ software. The experiment was conducted three times with similar observations.

Fluorescence recovery after photobleaching (FRAP) measurements

FRAP experiments were conducted using a Zeiss LSM 710 laser scanning confocal microscope equipped with a 63× Plan-Apochromat oil immersion objective, NA1.4, at the Life Sciences and Engineering (LS&E) Infrastructure Center, Technion-Israel Institute of Technology, Haifa, Israel. The experiments were performed on condensates from by 100 µM PSMα3-FITC or LL-37-FITC mixed with PI-labeled Poly (AU) RNA at various concentrations as indicated in the figures. Photobleaching was performed using 405 nm and 488 nm laser lines. The experiments were conducted three times with similar observations.

For FRAP analysis inside the nucleolus of HeLa cells, PSMα3 was added to the cells at a final concentration of 20 µM containing 20% FITC-labeled and 80% unlabeled peptide. The cells were incubated with the peptide for 20 minutes before conducting the experiment. Data was processed and analyzed using ZEISS ZEN software and Fiji-ImageJ software. The experiments were conducted three times with similar observations.

Total internal reflection fluorescence (TIRF) microscopy

TIRF microscopy was performed using a ZEISS Elyra 7 Super-Resolution Microscope equipped with a 63× Apochromat alpha Plan-Apochromat Oil immersion DIC objective, NA1.46, at the Life Sciences and Engineering (LS&E) Infrastructure Center, Technion-Israel Institute of Technology, Haifa, Israel. Images were acquired using the pco.edge sCMOS cameras, and data were processed and analyzed with ZEISS ZEN software to accurately represent the observed phenomena. The condensates and aggregates analyzed were generated using 100 µM PSMα3-FITC with either 50 ng/µL or 400 ng/µL Poly (AU) RNA. AmyTracker 630 (Ebba Biotech) was added at a 1:500 (peptide: AmyTracker) molar ratio to each sample. The experiments were conducted three times with similar observations.

Confocal microscopy visualization of PSMα3 interaction with HeLa cells

HeLa cells were pre-cultured one day before the experiment by preparing a suspension containing 350,000 cells/ml and plating 150 µl of this suspension into each well of a µ-Slide 8 well glass-bottom chamber. The cells were then incubated overnight under standard growth conditions (37°C, 5% CO2) to allow for adherence and growth. On the day of the experiment, the cells were washed three times with phosphate-buffered saline (PBS) to remove any residual media. Hoechst 33342 dye (10 mg/mL stock) was diluted 1:2000 in fresh cell media and added to the cells. The cells were incubated with Hoechst for 10 minutes at 37°C and 5% CO2. After incubation, cells were washed three times with PBS to remove Hoechst residuals. A working solution of PI was prepared by diluting a 1 mg/ml PI stock solution to a final concentration of 0.02 mg/ml in fresh cell growth media. Immediately prior to imaging, PSMα3 was added to the cells at a final concentration of 20 µM containing 20% FITC-labeled and 80% unlabeled peptide. The cells were then imaged using a Ti2-E microscope by Nikon with a CSU-W1 spinning disk confocal unit by Yokogawa, equipped with a 100X CFI SR HP Plan Apochromat Lambda S silicone immersion objective, NA1.35, at the Life Sciences and Engineering (LS&E) Infrastructure Center, Technion-Israel Institute of Technology, Haifa, Israel. Confocal movies and images were captured by Photometrics BSI sCMOS cameras to observe the interaction and localization of PSMα3 within the cells, focusing particularly on its colocalization with nucleic acids in the nucleolar region. The acquired images and movies were subsequently analyzed using Imaris Image Analysis Software (Oxford Instruments). The experiments were conducted three times with similar observations.

Super resolution light microscopy visualization of PSMα3 interaction with E. coli

E. coli lptD4213 cells were a kind gift from Prof. Sima Yaron from the Technion – Israel Institute of Technology. An inoculum was pre-cultured in Luria-Bertani medium (LB) medium at 220 rpm and 37°C for 24 hours prior to the experiment. On the day of the experiment, the optical density (OD) of the culture was measured, and the cells were diluted to an OD of 0.4. The cells were then incubated with 20 μM PSMα3, with and without EGCG at 1:5 molar ratio, alongside a bacteria-only control. Samples were incubated for 30 minutes at 220 rpm and 37°C. Following incubation, the samples were centrifuged three times for 3 minutes at 1.5g, with the supernatant discarded and the pellet resuspended in 1X PBS after each spin. The resuspended cells were then stained with 50 μg/mL wheat germ agglutinin (WGA) CF633 for 30 minutes. After staining, the cells were washed three times by centrifugation with PBS to remove excess WGA. During the final wash, the cells were resuspended in 0.1 mg/mL PI solution.

Prior to imaging, the samples were loaded into an ibidi µ-Slide VI 0.4. Imaging was performed using a ZEISS Elyra 7 Super-Resolution Microscope equipped with a 63× Apochromat alpha Plan-Apochromat Oil immersion DIC objective, NA 1.4, and pco.edge sCMOS cameras. Lattice structured illumination images were acquired and processed using the ZEN black software at the Life Sciences and Engineering (LS&E) Infrastructure Center, Technion-Israel Institute of Technology, Haifa, Israel. The experiments were conducted three times with similar observations.

Thioflavin T fluorescence fibrillation kinetics assay

Thioflavin T (ThT) (Sigma Aldrich) is a widely used fluorescent dye for detecting and analyzing amyloid fibril formation kinetics. Fibrillation curves in the presence of ThT typically exhibit an initial lag phase followed by rapid fibril elongation. To accurately capture the fibrillation lag time, PSMα3 was pre-treated before the experiment. The peptide was dissolved in a 1:1 mixture of 1,1,1,3,3,3-Hexafluoroisopropanol (HFIP-Sigma Aldrich) and Trifluoroacetic acid (TFA-Sigma Aldrich) to a final concentration of 1 mg/mL, followed by bath sonication for 10 minutes at room temperature. The organic solvents were then evaporated using a mini-rotational vacuum concentrator (Christ, Germany) at 1,000 rpm for 2 hours at room temperature.

For the experiment, freshly prepared 100 µM PSMα3 peptides, with or without 500 µM Epigallocatechin gallate (EGCG) at a 1:5 molar ratio, were prepared in 50 mM HEPES buffer pH 7.4 containing 150 mM NaCl. ThT was prepared by diluting a stock solution in ultrapure double-distilled water (UPddw) and filtered before use to reach a final concentration of 200 µM Blank control solutions containing all components except for the peptide were prepared for each reaction.

The assay was conducted in black 96-well flat-bottom plates (Greiner Bio-One), which were sealed with a thermal seal film (EXCEL Scientific) to prevent evaporation. Samples were incubated in a plate reader (OMEGA) at 37°C, shaking at 500 rpm for 85 seconds before each reading cycle, for up to 1,000 cycles of 6 minutes each, totaling approximately 100 hours. Fluorescence was measured in triplicate using an excitation wavelength of 438 ± 20 nm and an emission wavelength of 490 ± 20 nm. All values were averaged, background fluorescence was subtracted using blank controls, and the results were plotted over time. Standard error of the mean is shown as error bars. The entire experiment was independently repeated at least three times on different days to ensure reproducibility.

Solid-State Circular Dichroism Spectroscopy

Solid-state circular dichroism spectroscopy(ss-CD) was conducted with PSMa3 to assess their fibrillar secondary structure components in presence and absence of Poly (AU) RNA. For preparation of peptide and RNA samples, see Peptide and Poly (AU) preparation section. PSMa3 was prepared at a working concentration of 100 µM in 50 mM HEPES, pH 7.4, 150 mM NaCl at 0 ng/µL, 50 ng/µL or 400 ng/µL Poly (AU) RNA. The reaction mixes were incubated for 0 min, 2 h and 72 h at 37 °C in a non-shaking thermocycler. Following incubation, soluble reaction components were removed via dilution washing with 150 µL water and centrifugation at 12.500 rcf for 30 min. 20 µL of the sediment were resuspended in 180 µL water for a second washing centrifugation step at 12.500 rcf for 30 min. 180 µL of the supernatant were discarded and 18 µL of sediment applied to a Chirascan Series fused silica disc (AP/CSSD, Applied Photophysics) in three times 6 µL steps. After every sample application onto to the silica disc, all liquid was evaporated at 37 °C for 5–10 min until only an opaque film remained visible. For secondary structure determination of aggregated/fibrillar peptide components on the silica disc, the disc was inserted into a Chirascan solid sample holder (CS/SSH, Applied Photophysics) and the circular dichroism recorded in the far-UV (180–250 nm, step size 1 nm, time-per-point 1 s) using a CD spectropolarimeter (Applied Photophysics). To counteract sample anisotropy, CD spectra were recorded at least four different disc rotations and the results averaged. Prior to data analysis, silica disc and buffer/water backgrounds were subtracted from recorded sample spectra. Poly (AU) RNA-only background spectra showed complex and high amplitude signals that did not appear for samples in combination with peptides and were therefore not subtracted from sample spectra.

NMR experiments

To prepare the reference samples for NMR analysis, 0.521 mg of PSMα3 was dissolved in 180 µL of 20 mM MES (2-(N-morpholino)ethanesulfonic acid) buffer at pH 6.0, supplemented with 5% (v/v) deuterium oxide (D₂O). From this solution, 170 µL was transferred into a 3.0 mm NMR tube for spectral acquisition. In parallel, a 4 mM stock solution of EGCG was prepared in the same 20 mM MES buffer (pH 6.0). From this, a working solution containing 0.5 mM EGCG in 20 mM MES with 5% (v/v) D₂O was prepared to match the buffer conditions of the peptide sample. A volume of 170 µL was then transferred into a separate 3.0 mm NMR tube for use as the EGCG reference.

All NMR experiments were recorded on a Bruker 700 MHz spectrometer equipped with a triple-resonance cryoprobe (Prodigy) and an AVANCE NEO console. One-dimensional (1D) ¹H-NMR spectra were acquired using the Bruker library pulse sequence zggpw5 for water suppression. The acquisition parameters were: acquisition time (AQ) = 1.90 s, spectral width (SW) = 12.31 ppm, recovery delay (D1) = 1.5 s, and 64 scans. All experiments were conducted at 35 °C, under which optimal spectral quality was obtained.

Two-dimensional (2D) ¹H–¹H TOCSY (Total Correlation Spectroscopy) spectra were collected using the Bruker pulse sequence dipsi2gpph19. Acquisition parameters were: t₂ = 2048 complex points (AQ₂ = 0.118 s), t₁ = 512 complex points (AQ₁ = 0.029 s), SW₁ = SW₂ = 12.31 ppm, with 64 scans and a mixing time (tmix) of 80 ms. Recovery delay (D1) was set to 1.5 s. 2D ¹H–¹H NOESY (Nuclear Overhauser Effect Spectroscopy) spectra were acquired using the noesyfpgpph19 pulse sequence, with acquisition parameters identical to those used for TOCSY. The mixing time was set to 300 ms. Total experimental time for the combined 2D TOCSY and NOESY experiments was approximately 30 hours. NMR data processing was carried out using TopSpin 4.5.0 (Bruker), and spectral assignments for the 2D TOCSY and NOESY spectra were performed using POKY software68. Signal integration was carried out using in-house scripts written in Scilab69.

eLife Assessment

This study investigates how RNA molecules modulate phase separation, aggregation, and cytotoxicity of the staphylococcal virulent peptide PSMα3 and the human host‑defence peptide LL‑37 using an array of biophysical and cell‑based assays. If validated, these findings would be important, as they suggest that nucleic acids can tune the material state and bioactivity of amyloids, with implications for host-pathogen interactions and for the design of therapeutics that target phase behaviour. However, the evidence is incomplete: many key claims rest on qualitative imaging and contested assumptions about "functional" amyloids, and the absence of quantitative binding data, phase diagrams, and appropriate controls limits confidence in the conclusions.

Reviewer #1 (Public review):

Summary:

The manuscript by Rayan et al. aims to elucidate the role of RNA as a context-dependent modulator of liquid-liquid phase separation (LLPS), aggregation, and bioactivity of the amyloidogenic peptides PSMα3 and LL-37, motivated by their structural and functional similarities.

Strengths:

The authors combine extensive biophysical characterization with cell-based assays to investigate how RNA differentially regulates peptide aggregation states and associated cytotoxic and antimicrobial functions.

Weaknesses:

While the study addresses an interesting and timely question with potentially broad implications for host-pathogen interactions and amyloid biology, several aspects of the experimental design and data analysis require further clarification and strengthening.

Major Comments:

(1) In Figure 1A, the author showed "stronger binding affinity" based on shifts at lower peptide concentrations, but no quantitative binding parameters (e.g., apparent Kd, fraction bound, or densitometric analysis) are presented. This claim would be better supported by including: (i) A binding curve with quantification of free vs bound RNA band intensities (ii) Replicates and error estimates (mean {plus minus} SD).

(2) The authors report droplet formation at low RNA (50 ng/µL) but protein aggregation at high RNA (400 ng/µL) through fluorescence microscopy. However, no intermediate RNA concentrations (e.g., 100-300 ng/µL) are tested or discussed, leaving a critical gap in understanding the full phase diagram and transition mechanisms. Additionally, the behaviour of PSMα3 in the absence of RNA under LLPS conditions is not shown. Without protein-only data, it is difficult to assess if droplets are RNA-induced or if protein has a weak baseline LLPS that RNA tunes. The saturation concentration (csat) for PSMα3 phase separation, either in the absence or presence of RNA, should be reported.

(3) For a convincing LLPS claim, it is important to show: Quantitative FRAP curves (mobile fraction and half-time of recovery) rather than only microscopy images and qualitative statements.

(4) The manuscript highly relies on fluorescence microscopy to show colocalization. However, the colocalization is presented in a qualitative manner only. The manuscript would benefit from the inclusion of quantitative metrics (e.g., Pearson's correlation coefficient, Manders' overlap coefficients, or intensity correlation analysis).

(5) In Figures 3 B and 3C, the contrast between "no AT630 at 30 min, strong at 2 h" (50 ng/μL) and "strong at 30 min" (400 ng/μL) is compelling, but a simple quantification (e.g., mean fluorescence intensity per area) would greatly increase rigor.

(6) In Figure S3 ssCD data, if possible, indicate whether the α-helical signal increases with RNA concentration or shows a non-linear dependence, which might link to the LLPS vs solid aggregate regimes.

(7) In Figure 5B, FRAP recovery in dying cells may reflect artifactual mobility rather than biological relevance. Additionally, the absence of quantification data limits interpretation; providing recovery curves would clarify relevance.

(8) The narrative conflates cytotoxicity endpoints (membrane damage, PI staining, aggregates) with localization data (nucleolar foci), creating ambiguity about whether nucleolar targeting drives toxicity or is a consequence of cell death. Separating toxicity assessment from localization analysis, or clearly demonstrating that nucleolar accumulation precedes cytotoxicity, would resolve this ambiguity.

(9) In Figure 8, to strengthen the LLPS assignment for LL-37, additional evidence, such as FRAP analysis or observation of droplet fusion events, would be valuable. This is particularly relevant given that the heat shock conditions (65{degree sign}C for 15 minutes) could potentially induce partial denaturation or nonspecific coacervation.

Reviewer #2 (Public review):

In this paper, Rayan et al. report that RNA influences cytotoxic activity of the staphylococcal secreted peptide cytolysin PSMalpha3 versus human cells and E. coli by impacting its aggregation. The authors used sophisticated methods of structural analysis and described the associated liquid-liquid phase separation. They also compare the influence of RNA on the aggregation and activity of LL-37, which shows differences from that on PSMalpha3.

Strengths:

That RNA impacts PSM cytotoxicity when co-incubated in vitro becomes clear.

Weaknesses:

I have two major and fundamental problems with this study:

(1) The premise, as stated in the introduction and elsewhere, that PSMalpha3 amyloids are biologically functional, is highly debatable and has never been conclusively substantiated. The property that matters most for the present study, cytotoxicity, is generally attributed to PSM monomers, not amyloids. The likely erroneous notion that PSM amyloids are the predominant cytotoxic form is derived from an earlier study by the authors that has described a specific amyloid structure of aggregated PSMalpha3. Other authors have later produced evidence that, quite unsurprisingly, indicated that aggregation into amyloids decreases, rather than increases, PSM cytotoxicity. Unfortunately, yet other groups have, in the meantime, published in-vitro studies on "functional amyloids" by PSMs without critically challenging the concept of PSM amyloid "functionality". Of note, the authors' own data in the present study, which show strongly decreased cytotoxicity of PSMalpha3 after prolonged incubation, are in agreement with monomer-associated cytotoxicity as they can be easily explained by the removal of biologically active monomers from the solution.

(2) That RNA may interfere with PSM aggregation and influence activity is not very surprising, given that PSM attachment to nucleic acids - while not studied in as much detail as here - has been described. Importantly, it does not become clear whether this effect has biologically significant consequences beyond influencing, again not surprisingly, cytotoxicity in vitro. The authors do show in nice microscopic analyses that labeled PSMalpha3 attaches to nuclei when incubated with HeLa cells. However, given that the cells are killed rapidly by membrane perturbation by the applied PSM concentrations, it remains unclear and untested whether the attachment to nucleic acids in dying cells makes any contribution to PSM-induced cell death or has any other biological significance.

Overall, the findings can be explained in a much more straightforward way with the common concept of cytotoxicity being due to monomeric PSMs, and the impact of nucleic acids on cytotoxicity being due to lowering of the concentration of that active form by RNA attachment. Further limiting the significance of the findings, whether this interaction has any biological significance on the physiology or infectivity of the PSM producer remains largely unexplored.

Reviewer #3 (Public review):

Summary:

The manuscript by Rayan et al. aims to investigate the role of RNA in modulating both virulent amyloid and host-defense peptides, with the objective of understanding their self-assembly mechanisms, morphological features, and aggregation pathways.

Strengths:

The overall content is well-structured with a logical flow of ideas that effectively conveys the research objectives.

Weaknesses:

(1) Figure 2 displays representative FRAP images demonstrating fluorescence recovery within seconds. To gain a more comprehensive understanding of how recovery after photobleaching varies under different conditions, it is recommended to supplement these images with corresponding quantitative fluorescence recovery curves for analysis.

(2) Ostwald ripening typically leads to the shrinkage or even disappearance of smaller droplets, accompanied by the further growth of large droplets. However, the droplet size in Figure 2D decreases significantly after 2 h of incubation. This observation prompts the question, what is the driving force underlying RNA-regulated phase separation and phase transition?

(3) The manuscript aims to study the role of RNA in modulating PSMα3 aggregation by using solution-state NMR to obtain residue-specific structural information. The current NMR data, as described in the method and figure captions, were recorded in the absence of RNA. Whether RNA binding induces conformational changes of PSMα3, and how these changes alter the NMR spectra? Also, the sequential NOE walk between neighboring residues can be annotated on the spectrum for clarity.

(4) The authors claim that LL-37 shares functional, sequence, and structural similarities with PSMα3. However, no droplet formation was observed of LL-37 in the presence of RNA only. The authors then applied thermal stress to induce phase separation of LL-37. What are the main factors contributing to the different phase behaviors exhibited by LL-37 and PSMα3? What are the differences in the conformation of amyloid aggregates and the kinetics of aggregation between the condensation-induced aggregation in the presence of RNA and the conventional nucleation-elongation process in the absence of RNA for these two proteins?

Author response:

The following is the authors’ response to the original reviews.

Public Reviews:

Reviewer #1 (Public review):

Summary:

The manuscript by Rayan et al. aims to elucidate the role of RNA as a context-dependent modulator of liquid-liquid phase separation (LLPS), aggregation, and bioactivity of the amyloidogenic peptides PSMα3 and LL-37, motivated by their structural and functional similarities.

Strengths:

The authors combine extensive biophysical characterization with cell-based assays to investigate how RNA differentially regulates peptide aggregation states and associated cytotoxic and antimicrobial functions.

Weaknesses:

While the study addresses an interesting and timely question with potentially broad implications for host-pathogen interactions and amyloid biology, several aspects of the experimental design and data analysis require further clarification and strengthening.

Major Comments:

(1) In Figure 1A, the author showed "stronger binding affinity" based on shifts at lower peptide concentrations, but no quantitative binding parameters (e.g., apparent Kd, fraction bound, or densitometric analysis) are presented. This claim would be better supported by including: (i) A binding curve with quantification of free vs bound RNA band intensities (ii) Replicates and error estimates (mean {plus minus} SD).

We thank the reviewer for this suggestion. To quantitatively support the binding differences observed in Figure 1A, we have now performed densitometric analysis of the EMSA data and included the results in Figure S1. The analysis showed that the Kd for PSMα3 binding to polyAU and polyA RNA is in the same order of magnitude but lower for the polyAU, indicating a stronger binding. A description was added to the results in lines 137-145 of the revised version.

(2) The authors report droplet formation at low RNA (50 ng/µL) but protein aggregation at high RNA (400 ng/µL) through fluorescence microscopy. However, no intermediate RNA concentrations (e.g., 100-300 ng/µL) are tested or discussed, leaving a critical gap in understanding the full phase diagram and transition mechanisms.

Our initial choice of 50 ng/µL (low RNA) and 400 ng/µL (high RNA) was guided by a broader RNA titration performed by turbidity measurements across 0, 10, 20, 50, 100, 200, and 400 ng/µL (Figure S2 in the revised version). In this screen, turbidity increased up to 50 ng/µL and then decreased dose-dependently from 100–400 ng/µL. We interpret this non-monotonic behavior as consistent with a transition from a droplet rich regime (maximal light scattering at intermediate dense-phase volume) toward conditions where assemblies become larger and/or more compact and sediment out of the optical path. This is described in lines 158-161 of the revised version.

Of note, additional intermediate RNA conditions (100 and 200 ng/µL) are included in Figure S14 (of the revised version). While these experiments were performed under the heat-shock perturbation, they nevertheless support the central point that RNA tunes assembly state across intermediate concentrations rather than producing a binary low/high outcome.

Importantly, we agree with the reviewer that a full phase diagram would be the most rigorous way to define the transition mechanism. However, establishing csat and constructing a complete phase diagram would require systematic measurements of dilute-phase concentrations (e.g., centrifugation/quantification or fluorescence calibration), controlled ionic strength titrations, and time-resolved mapping, which is beyond the scope of the present study. We have therefore revised the text to avoid implying that we provide a complete phase diagram. Instead, we frame our results as a qualitative with multi-assay characterization showing that RNA concentration drives a shift from liquid-like condensates (at low RNA) toward solid-like assemblies (at high RNA), with an intermediate regime suggested by the turbidity transition and supported by additional imaging under stress. Finally, to address the “critical gap” concern directly, we add a sentence (lines 239-241) stating that: “Future work will be required to quantitatively define the phase boundaries and delineate the dominant mechanisms, such as sedimentation, dissolution, or coarsening/aging, across intermediate RNA concentrations”.

(3) Additionally, the behaviour of PSMα3 in the absence of RNA under LLPS conditions is not shown. Without protein-only data, it is difficult to assess if droplets are RNA-induced or if protein has a weak baseline LLPS that RNA tunes. The saturation concentration (csat) for PSMα3 phase separation, either in the absence or presence of RNA, should be reported.

In response to the reviewer’s request, we have added Figure 2F, which shows PSMα3 alone in the absence of RNA under the same conditions. PSMα3 does not form droplets in this condition, indicating that condensate formation is RNA-dependent in the tested conditions. This is referred to in the text in lines 190-193 of the revised version. Please see our response about determining the csat in the response to the previous comment.

(4) For a convincing LLPS claim, it is important to show: Quantitative FRAP curves (mobile fraction and half-time of recovery) rather than only microscopy images and qualitative statements.

We have included quantitative FRAP analysis in Figure S4 of the revised version, showing normalized recovery curves along with extracted mobile fractions and half-times of recovery (t₁/₂). These quantitative measurements support the dynamic nature of the PSMα3–RNA. This is referred to in the text in lines 179-184 of the revised version.

(5) The manuscript highly relies on fluorescence microscopy to show colocalization. However, the colocalization is presented in a qualitative manner only. The manuscript would benefit from the inclusion of quantitative metrics (e.g., Pearson's correlation coefficient, Manders' overlap coefficients, or intensity correlation analysis).

In response, we have added quantitative colocalization analysis to the revised manuscript. Specifically, we now report Pearson’s correlation coefficients and Manders’ overlap coefficients for the dual-channel fluorescence microscopy datasets in Figure S5 of the revised version. These metrics provide an objective measure of co-distribution and complement the qualitative imaging.

The analysis supports that at low RNA concentrations (droplet/condensate conditions), PSMα3 and RNA show strong colocalization, consistent with RNA being incorporated within, or closely associated with, the peptide-rich phase. In contrast, at high RNA concentrations, where the assemblies are more solid-like/amyloid-positive, the quantitative coefficients decrease, consistent with reduced overlap and an apparent spatial demixing in which RNA becomes partially excluded from the peptide-rich structures. This is referred to in the text in lines 194-203 of the revised version.

(6) In Figures 3 B and 3C, the contrast between "no AT630 at 30 min, strong at 2 h" (50 ng/μL) and "strong at 30 min" (400 ng/μL) is compelling, but a simple quantification (e.g., mean fluorescence intensity per area) would greatly increase rigor.

We have included quantitative analysis of AmyTracker630 fluorescence intensity in Figure S6 of the revised version, reporting the mean fluorescence intensity per area for the indicated conditions and time points. This quantification supports the qualitative differences observed in Figures 3B and 3C. This is now referred to in the text in lines 233-236 of the revised version.

(7) In Figure S3 ssCD data, if possible, indicate whether the α-helical signal increases with RNA concentration or shows a non-linear dependence, which might link to the LLPS vs solid aggregate regimes.

The ssCD spectra displayed in Figure S7 in the revised version (corresponding to Figure S3 in the original submission) show that the α-helical signature of PSMα3 is markedly enhanced in the presence of RNA compared to peptide alone, as evidenced by increased signal intensity, deeper minima, and more pronounced spectral features characteristic of α-helical structure. Importantly, this enhancement is more pronounced at 400 ng/µL Poly(AU) RNA than at 50 ng/µL, particularly after 2 hours of coincubation, indicating that RNA concentration influences the stabilization of α-helical assemblies. This is now more specifically detailed in the text in lines 258-263 of the revised version.

We note that solid-state CD does not allow direct quantitative deconvolution of secondary structure content (e.g., % helix) in the same manner as solution CD, due to sample anisotropy, scattering, and orientation effects inherent to dried or aggregated films. Consequently, our interpretation is qualitative rather than strictly quantitative. The ssCD data therefore suggest a non-linear dependence on RNA concentration, rather than a simple linear dose–response. This is also expected considering that phase transition, suggested by the other findings, is intrinsically non-linear.

(8) In Figure 5B, FRAP recovery in dying cells may reflect artifactual mobility rather than biological relevance. Additionally, the absence of quantification data limits interpretation; providing recovery curves would clarify relevance.”

We added quantitative FRAP analysis of the effect on PSMα3 within HeLa cells, shown in Figure S8 of the revised version. Compared to PSMα3 assemblies in vitro, nucleolar PSMα3 exhibits slower fluorescence recovery and a reduced mobile fraction. The nucleolus represents a highly crowded, RNA-rich cellular environment, which is expected to impose additional constraints on molecular mobility and likely contributes to the slower recovery kinetics observed in cells. This is now more specifically detailed in the text in lines 324-333 and discussed in lines 597-607 of the revised version.

(9) The narrative conflates cytotoxicity endpoints (membrane damage, PI staining, aggregates) with localization data (nucleolar foci), creating ambiguity about whether nucleolar targeting drives toxicity or is a consequence of cell death. Separating toxicity assessment from localization analysis, or clearly demonstrating that nucleolar accumulation precedes cytotoxicity, would resolve this ambiguity.

We thank the reviewer for raising this important point. We agree that, in the current dataset, cytotoxicity readouts (membrane damage, PI staining, aggregate formation) and subcellular localization (nucleolar accumulation) are observed in close temporal proximity, which limits our ability to unambiguously assign causality. In the experiments presented here, PSMα3 was applied at concentrations known to induce rapid membrane disruption and cytotoxicity in HeLa cells. Under these conditions, PSMα3 accumulates on cellular membranes and penetrates into the cell and nucleus on very short timescales (seconds to minutes), likely preceding the temporal resolution accessible by standard live-cell fluorescence microscopy. As a result, nucleolar accumulation and cytotoxic endpoints are detected essentially concurrently, precluding a definitive determination of whether nucleolar association actively drives toxicity or occurs as a downstream consequence of membrane permeabilization and cell damage.

We therefore emphasize that, in this study, nucleolar localization is presented as a phenomenological observation consistent with RNA-rich compartment association, rather than as a demonstrated causal mechanism of cytotoxicity. We have revised the Discussion (lines 597-607) to clarify this distinction and to avoid implying that nucleolar targeting is the primary driver of cell death.

We agree that resolving this ambiguity would require systematic time-resolved and concentration-dependent experiments, including analysis at sub-toxic PSMα3 concentrations below the membrane-disruptive threshold, combined with orthogonal imaging approaches. Such experiments are planned for future work but are beyond the scope of the present study.

(10) In Figure 8, to strengthen the LLPS assignment for LL-37, additional evidence, such as FRAP analysis or observation of droplet fusion events, would be valuable. This is particularly relevant given that the heat shock conditions (65 °C for 15 minutes) could potentially induce partial denaturation or nonspecific coacervation.

In response to this comment, we have added FRAP analysis of LL-37 assemblies in the revised manuscript (Figure S12), including representative images and corresponding fluorescence recovery curves. The FRAP measurements show minimal fluorescence recovery over the acquisition window, indicating that the LL-37–RNA assemblies formed under these conditions are largely immobile and solid-like, rather than liquid-like droplets. This is now referred to in the text in lines 458-462 of the revised version.

Reviewer #2 (Public review):

In this paper, Rayan et al. report that RNA influences cytotoxic activity of the staphylococcal secreted peptide cytolysin PSMalpha3 versus human cells and E. coli by impacting its aggregation. The authors used sophisticated methods of structural analysis and described the associated liquid-liquid phase separation. They also compare the influence of RNA on the aggregation and activity of LL-37, which shows differences from that on PSMalpha3.

Strengths:

That RNA impacts PSM cytotoxicity when co-incubated in vitro becomes clear.

Weaknesses:

I have two major and fundamental problems with this study:

(1) The premise, as stated in the introduction and elsewhere, that PSMalpha3 amyloids are biologically functional, is highly debatable and has never been conclusively substantiated. The property that matters most for the present study, cytotoxicity, is generally attributed to PSM monomers, not amyloids. The likely erroneous notion that PSM amyloids are the predominant cytotoxic form is derived from an earlier study by the authors that has described a specific amyloid structure of aggregated PSMalpha3. Other authors have later produced evidence that, quite unsurprisingly, indicated that aggregation into amyloids decreases, rather than increases, PSM cytotoxicity. Unfortunately, yet other groups have, in the meantime, published in-vitro studies on "functional amyloids" by PSMs without critically challenging the concept of PSM amyloid "functionality". Of note, the authors' own data in the present study, which show strongly decreased cytotoxicity of PSMalpha3 after prolonged incubation, are in agreement with monomer-associated cytotoxicity as they can be easily explained by the removal of biologically active monomers from the solution.

We thank the reviewer for this important critique and agree that direct cytotoxicity is most plausibly mediated by soluble PSM species, while extensive fibrillation generally reduces toxicity by depleting these forms, a conclusion supported by our data and by other studies (e.g., Zheng et al 2018 and Yao et al 2019). We do not propose mature amyloid fibrils as the primary toxic entities. Rather, we use the term functional amyloid in a regulatory sense, consistent with other biological amyloids whose fibrillar states modulate activity (e.g., hormone storage amyloids or RNA-binding proteins).

In line with emerging findings, we interpret PSMα3 toxicity as arising from a dynamic assembly process rather than from a single static molecular species. We previously showed that PSMα3 forms cross-α fibrils that are thermodynamically and mechanically less stable than cross-β amyloids and readily disassemble upon heat stress, fully restoring cytotoxic activity (Rayan et al., 2023). This behavior contrasts with PSMα1, which forms highly stable cross-β fibrils that do not recover activity after heat shock, suggesting that the limited thermostability of PSMα3 is an evolved feature enabling reversible switching between inactive (stored) and active states.

Consistent with this view, both PSMα1 and PSMα3 are cytotoxic in their soluble states, yet mutants unable to fibrillate lose activity, indicating that fibrillation is required but not itself the toxic end state (Tayeb-Fligelman et al., 2017, 2020; Malishev et al., 2018). Our other studies further show that cytotoxicity toward human cells correlates with inherent or lipid-induced α-helical assemblies, rather than with inert β-sheet amyloids (RagonisBachar et al., 2022, 2026; Salinas 2020, Bücker 2022). Together, these findings support a model in which membrane-associated, dynamic α-helical assembly, which requires continuous exchange between soluble species and growing fibrils, drives membrane disruption, potentially through lipid recruitment or extraction, analogous to mechanisms proposed for human amyloids such as islet amyloid polypeptide (Sparr et al., 2004).

In the present study, we further show that RNA reshapes this dynamic landscape: while PSMα3 alone progressively loses activity upon incubation, co-incubation with RNA preserves cytotoxicity by stabilizing bioactive polymorphs and condensate-like states, whereas high RNA concentrations promote solid aggregation but nevertheless preserve activity. Thus, aggregation is neither inherently functional nor toxic, but context dependent and environmentally regulated. Taken together, our data support a model in which PSMα3 amyloids act as a dynamic reservoir, enabling S. aureus to tune virulence by reversibly shifting between dormant and active states in response to environmental cues such as heat or RNA.

This is now discussed in lines 56-76 and 523-553 of the revised version.

(2) That RNA may interfere with PSM aggregation and influence activity is not very surprising, given that PSM attachment to nucleic acids - while not studied in as much detail as here - has been described. Importantly, it does not become clear whether this effect has biologically significant consequences beyond influencing, again not surprisingly, cytotoxicity in vitro. The authors do show in nice microscopic analyses that labeled PSMalpha3 attaches to nuclei when incubated with HeLa cells. However, given that the cells are killed rapidly by membrane perturbation by the applied PSM concentrations, it remains unclear and untested whether the attachment to nucleic acids in dying cells makes any contribution to PSM-induced cell death or has any other biological significance.

We thank the reviewer for this important point and agree that PSM–nucleic acid interactions are not unexpected and that our data do not support a direct intracellular role for RNA binding in mediating cytotoxicity. Accordingly, we do not propose nucleolar or nuclear association of PSMα3 as a causal mechanism of cell death. At the concentrations used, PSMα3 induces rapid membrane disruption, and nucleic acid association is observed along with membrane attachment, precluding conclusions about intracellular function. This limitation is now explicitly clarified in the revised manuscript. The biological significance of our findings lies instead in extracellular and environmental contexts, where PSMα3 encounters abundant nucleic acids, such as RNA or DNA released from damaged host cells or present in biofilms as now addressed in lines 622631. Our data show that RNA modulates PSMα3 aggregation trajectories, shifting the balance between liquid-like condensates and solid aggregates, and thereby regulates the persistence and timing of cytotoxic activity. In this framework, RNA acts as a context dependent regulator of virulence, rather than as an intracellular cytotoxic cofactor, an aspect which would be studied in depth in future work. This is now addressed in the text in lines 597-607 of the revised version.

Reviewer #3 (Public review):

Summary:

The manuscript by Rayan et al. aims to investigate the role of RNA in modulating both virulent amyloid and host-defense peptides, with the objective of understanding their self-assembly mechanisms, morphological features, and aggregation pathways.

Strengths:

The overall content is well-structured with a logical flow of ideas that effectively conveys the research objectives.

Weaknesses:

(1) Figure 2 displays representative FRAP images demonstrating fluorescence recovery within seconds. To gain a more comprehensive understanding of how recovery after photobleaching varies under different conditions, it is recommended to supplement these images with corresponding quantitative fluorescence recovery curves for analysis.

In response to this comment, we have supplemented the representative FRAP images with quantitative fluorescence recovery curves, reporting normalized recovery kinetics for the indicated conditions. These data are now provided in Figure S4 of the revised manuscript, allowing direct comparison of recovery behavior across conditions (shown by microscopy in Figure 2). In addition, we have included quantitative FRAP analyses for the cellular imaging shown in Figure 5 (presented in Figure S8) and for LL-37 assemblies formed under heat-shock conditions (Figure S12). Together, these additions provide a quantitative framework for interpreting the FRAP results and strengthen the distinction between liquid-like and solid-like assembly states.

(2) Ostwald ripening typically leads to the shrinkage or even disappearance of smaller droplets, accompanied by the further growth of large droplets. However, the droplet size in Figure 2D decreases significantly after 2 h of incubation. This observation prompts the question, what is the driving force underlying RNA-regulated phase separation and phase transition?”

We thank the reviewer for this observation. Across multiple samples, we consistently observe a coexistence of small droplets and larger aggregates, rather than systematic growth of larger droplets at the expense of smaller ones or a uniform decrease in droplet size. In addition, the timescales examined do not allow us to reliably assess whether diffusion-driven droplet coalescence is fast enough to draw firm conclusions about droplet size evolution. This is now addressed in the text in lines 181-184 of the revised version.

A decrease in droplet size over time is nevertheless observed in some instances and is more consistent with a time-dependent conversion of initially liquid-like condensates into more solid-like assemblies, which would reduce molecular mobility and suppress droplet coalescence. In parallel, progressive fibril formation may act as a sink for soluble peptide, leading to partial dissolution or shrinkage of less mature condensates. Together, these observations are consistent with a non-equilibrium aging process, in which RNAregulated assemblies evolve from dynamic condensates toward more solid structures rather than following equilibrium Ostwald ripening.

(3) The manuscript aims to study the role of RNA in modulating PSMα3 aggregation by using solution-state NMR to obtain residue-specific structural information. The current NMR data, as described in the method and figure captions, were recorded in the absence of RNA. Whether RNA binding induces conformational changes of PSMα3, and how these changes alter the NMR spectra? Also, the sequential NOE walk between neighboring residues can be annotated on the spectrum for clarity.

The solution-state NMR experiments were performed specifically to characterize the potential binding of EGCG to PSMα3. Due to the strong tendency of PSMα3 to undergo rapid aggregation and line broadening upon RNA addition, solution state NMR spectra in the presence of RNA could not be obtained at sufficient quality for residue-specific analysis. As suggested, we have updated and annotated the sequential NOE walk between neighboring residues on the relevant NOESY spectra to improve clarity.

(4) The authors claim that LL-37 shares functional, sequence, and structural similarities with PSMα3. However, no droplet formation was observed of LL-37 in the presence of RNA only. The authors then applied thermal stress to induce phase separation of LL-37. What are the main factors contributing to the different phase behaviors exhibited by LL37 and PSMα3? What are the differences in the conformation of amyloid aggregates and the kinetics of aggregation between the condensation-induced aggregation in the presence of RNA and the conventional nucleation-elongation process in the absence of RNA for these two proteins?

We appreciate this important question and have clarified both the basis of the comparison and the origin of the divergent phase behaviors of LL-37 and PSMα3. While PSMα3 and LL-37 share key properties as short, cationic, amphipathic α-helical peptides that self-assemble and interact with nucleic acids, they differ fundamentally in their assembly architectures. PSMα3 is an amyloidogenic peptide that forms cross-α amyloid fibrils, in which α-helices stack perpendicular to the fibril axis. In contrast, LL-37 can form fibrillar or sheet-like assemblies (observed in cryo grids), but these lack canonical amyloid features without clear cross-α or cross-β amyloid order, as so far observed by crystal structures. This is now clarified in different parts of the text of the revised version. Thus, the comparison between the two peptides is functional and physicochemical rather than implying identical amyloid mechanisms. These structural differences likely underlie their distinct phase behaviors.

Because LL-37 does not follow a classical amyloid nucleation–elongation pathway, and high-resolution structural information (e.g., cryo-EM) is currently lacking, partly due to its sheet-like, non-twisted morphology (unpublished results), it is not possible to directly compare aggregation kinetics or nucleation mechanisms between LL-37 and PSMα3. It is possible that amyloidogenic systems such as PSMα3 exhibit greater flexibility in prefibrillar and fibrillar polymorphism, enabling RNA-regulated phase behavior, whereas non amyloid assemblies such as LL-37 are more prone to stress-induced solid aggregation. We note that this interpretation is necessarily tentative and does not imply a general rule, but rather reflects differences evident in the present system.

Recommendations for the authors:

Reviewer #1 (Recommendations for the authors):

Minor Comments:

(1) In the abstract, replacing the word "overriding" with "counteracting" may provide a scientifically neutral tone.

In the course of revision, the abstract was substantially rewritten to more precisely convey the mechanistic framework and key conclusions of the study. As part of this rewrite, the term "overriding" was removed and the language throughout was revised to adopt a more scientifically neutral tone, consistent with the reviewer's suggestion.

(2) In abstract, the final sentence is ambitious but heavy. It may benefit from being split into two shorter sentences, for example:

"These findings establish RNA as a potent, context-dependent modulator of both virulent amyloids and host-defense peptides. They further reveal phase transitions as tunable regulators of peptide activity and potential therapeutic targets across infectious and neurodegenerative diseases."

As part of the broader abstract revision, the final sentence was restructured and the abstract as a whole was rewritten to improve clarity and readability, in the spirit of the reviewer's recommendation.

(3) In the Introduction section,

The phenol-soluble modulins (PSMs) produced by Staphylococci contain amyloid-forming short peptides which play multiple functional roles...", consider "Staphylococcal phenolsoluble modulins (PSMs) are short, amyloidogenic peptides that perform multiple roles central to pathogenesis....

In accordance with the suggestion, the sentence has been revised.

(4) To improve narrative flow in the final paragraph of the Introduction, a short bridging sentence could be added, such as:

"Given these nucleic acid interactions, we next examined whether RNA can drive phase separation or structural reorganization of these amyloidogenic peptides."

We thank the reviewer for this helpful suggestion. It provided an opportunity to clarify an important distinction between the two peptides studied. While LL-37 can self-assemble into higher-order α-helical structures, it is not amyloidogenic, in contrast to PSMα3. We therefore revised the bridging sentence in the final paragraph of the Introduction to read: “Given their shared cationic, amphipathic α-helical character, but distinct amyloidogenic properties, we sought to examine whether RNA differentially influences the assembly landscapes and bioactivity of PSMα3 and LL-37. “

(5) The rationale for selecting Poly(A) and Poly(AU) would benefit from further clarification. It would be helpful to specify whether these RNAs are intended to model particular host or bacterial RNA species, such as AU-rich elements, rRNA-like sequences, or mRNA-like contexts.

Poly(A) and Poly(AU) RNAs were selected as simplified, well-defined model RNAs to probe general peptide–RNA interactions in an unbiased manner, as no prior information was available regarding whether such interactions occur or which specific RNA species might be involved. This rationale is now clarified in the revised text (lines 128–131).

These RNAs are not intended to represent a single biological transcript, but rather generic RNA features relevant to both host and bacterial contexts, including single-stranded homopolymeric regions and AU-rich elements commonly found in mRNAs and stress srelated RNAs. The use of such reductionist RNA models to study RNA–protein interactions, phase behavior, and RNA-modulated aggregation is well established. We nevertheless agree that RNA sequence and structure may influence peptide assembly and activity, and future studies will address sequence-specific and biologically derived RNAs.

(6) In Figure 1A, essential EMSA controls- RNA alone, peptide alone, and a nonspecific peptide or PSMα3 should be included to distinguish specific complexes from artifacts, even if presented in the supplementary information. In addition, a competition assay using unlabeled RNA would help confirm binding specificity and rule out predominantly nonspecific electrostatic interactions; these data could also be reported in the supplementary figures.

An RNA-alone control is already included in Figure 1A of the revised version. The first lane (“0 µM”) shows free Poly(A) or Poly(AU) RNA in the absence of peptide and serves as the negative control against which PSMα3-induced mobility shifts are evaluated. A peptide-alone EMSA cannot be performed, as PSMα3 is highly cationic and does not migrate into the gel in the absence of RNA; moreover, EMSA in this format reports on RNA mobility rather than peptide migration.

With respect to binding specificity, we compared Poly(A) and Poly(AU) RNAs and observed distinct binding behaviors, which would not be expected for purely nonspecific electrostatic interactions. In addition, the extracted Hill coefficients (>1) are consistent with cooperative binding, further arguing against simple charge-driven association. Finally, the RNA-dependent association of PSMα3 is independently supported by fluorescence microscopy and quantitative colocalization analyses, which corroborate the EMSA results. Together, these orthogonal approaches support the relevance of the observed peptide–RNA interactions.

(7) In Figure 1B, there is a time mismatch between EMSA (30 minutes) and TEM (2 hours). If aggregation progresses over time, the EMSA pattern at 2 hours may differ. This point could be acknowledged or experimentally addressed, as RNA-peptide assemblies may evolve from liquid-like condensates to more solid aggregates.

The EMSA and TEM experiments were intentionally performed at different time points to capture distinct stages of the PSMα3–RNA assembly process. The EMSA assay (30 minutes) was designed to probe early RNA–peptide complex formation and binding interactions, before extensive higher-order aggregation occurs. At this stage, we aim to detect mobility shifts reflecting complex formation rather than mature assemblies. In contrast, TEM was performed after 2 hours to visualize later-stage structural outcomes, including fibrillation and morphological reorganization. As aggregation progresses over time, the assemblies evolve from early RNA–peptide complexes into more ordered fibrillar structures, which are best assessed by electron microscopy at later time points. To improve clarity and avoid potential confusion, we have streamlined Figure 1 to focus on the EMSA data, which specifically addresses early binding events. The TEM data were removed from Figure 1 and are now presented in Figure 3, where later-stage structural transitions and fibrillation are shown more comprehensively and in the appropriate mechanistic context.

(8) In Figure 1B, if feasible, complementing TEM with a confirmatory fibril assay (e.g., ThT kinetics) under the same conditions would strengthen the conclusion that the morphology difference is robust, but it is not mandatory.

We attempted to perform ThT fibrillation kinetics under the same RNA containing conditions; however, these assays were not informative for this system. PSMα3 aggregates extremely rapidly, producing an immediate and steep increase in ThT fluorescence (Fig. S9 in the revised version), which prevents reliable resolution of RNA dependent differences in aggregation kinetics or lag phases. In addition, Poly(AU) RNA interferes with ThT readout through electrostatic interactions between the negatively charged RNA and the cationic dye, as well as through RNA-induced changes in fibril morphology, both of which complicate quantitative interpretation of fluorescence kinetics. Based on these technical constraints and prior experience with RNA–amyloid systems, ThT kinetics under identical RNA conditions would not provide a robust or interpretable confirmation of the morphological differences observed by TEM.

(9) In Figure 1B, PSMα3 alone control is missing in TEM images.

A TEM image of PSMα3 alone is included in Figure 3, where we systematically present fibrillation outcomes across different RNA concentrations alongside the peptide-only control. Figure 1 was streamlined to focus on early RNA– peptide interactions assessed by EMSA, whereas Figure 3 provides a comprehensive TEM analysis of later-stage structural outcomes. This organization was chosen to clearly separate early binding events from subsequent assembly transitions and to avoid redundant presentation of TEM images under similar conditions.

(10) Although it is experimentally practical to focus on Poly(AU), the justification is very one-sided. The Poly(A) condition, which yields amorphous aggregates, may be equally informative for understanding toxicity, LLPS, or nonfibrillar states and could be discussed more explicitly.

We agree that Poly(A)-induced amorphous aggregation is informative for understanding non fibrillar assembly states. However, the primary aim of this study was to dissect RNA-dependent regulation of fibrillar assembly and phase behavior, which is most clearly captured using Poly(AU). Poly(A) was therefore included as a comparative condition rather than as a focus for detailed mechanistic analysis. A more systematic comparison of different RNA classes and their effects on non fibrillar states and toxicity is an important direction for future work but is beyond the scope of the present study.

(11) To improve readability of the manuscript, the main text should follow the order of the figure panels (e.g., A, B, C, D, and E) and numbers (Figure 1, 2...) sequentially, so that readers can easily align with the corresponding images.

We have revised the manuscript to improve alignment between the main text and the figures, adjusting panel ordering and numbering where appropriate so that the text now follows the figure panels and figure numbers more sequentially. These changes were made to enhance readability while maintaining a logical visual flow within the figures.

(12) In the result section of Figure 2, the analogy to Ddx4-like systems is a helpful concept, but should be clearly framed as an analogy, not evidence. It would be more accurate to say that the behavior is "conceptually similar to" those systems, while noting that the molecular context is significantly different.

We have revised the text to explicitly frame the comparison to Ddx4-like systems as a conceptual analogy rather than evidence: lines 158-161 in the revised version.

(13) In Figure 4, inclusion of positive and negative controls to validate assay performance (e.g., untreated bacteria or HeLa cells, lysis buffer, media alone) would strengthen confidence in the bioactivity measurements.

We wish to clarify that appropriate positive and negative controls were included in all bioactivity assays and were used to normalize the data presented in Figure 4. For the HeLa cytotoxicity assay (LDH), untreated cells were used to determine spontaneous LDH release (negative control), and cells treated with the manufacturer supplied lysis buffer were used to determine maximum LDH release (positive control). The percent cytotoxicity shown in Figure 4B was calculated relative to these internal controls, as described in the Methods. For the antibacterial assay (PrestoBlue), wells containing E. coli without peptide served as the positive control for 100% viability, while wells containing sterile LB medium alone were used as blanks. Viability values in Figure 4A were normalized to these controls. We have ensured that the Methods section explicitly describes these controls to reinforce confidence in the bioactivity measurements.

(14) To enhance clarity, consider presenting the RNA concentration and time-dependent effects on PSMα3 bioactivity in a comparison table within the main text or as a supplementary figure.

We appreciate this suggestion and carefully considered presenting the data in tabular form. However, we found that graphical representation more effectively conveys the trends, transitions, and comparative patterns between conditions. A table would not adequately capture these relationships.

Reviewer #2 (Recommendations for the authors):

Further remarks:

(1) Circumstantial evidence based on the "amyloid inhibitor", EGCG: The results with EGCG, which has been shown to have a moderate amyloid-reducing effect on PSMalpha 1 and PSMalpha4, should not be taken as evidence for amyloid-based cytotoxicity. While increased concentrations of EGCG reduced the cytotoxic effect of PSMalpha3, it is not convincingly shown that this is due to a lower concentration of amyloid vs. monomeric PSM.

We agree that the effects of EGCG should not be interpreted as evidence for amyloid fibrils being the cytotoxic species. Our data instead support a mechanism in which EGCG primarily targets soluble PSMα3, thereby redirecting its assembly pathway and depleting bioactive species. Specifically, solution-state NMR (Fig. 7) shows that EGCG binds defined residues of monomeric PSMα3, consistent with sequestration of soluble peptide rather than selective inhibition of fibrils. Complementary light and electron microscopy, together with kinetic measurements, indicate that EGCG does not simply stabilize monomers but instead diverts PSMα3 into amorphous, non-functional aggregates, as visualized by TEM (Fig. 6B) and reflected in altered ThT responses (Fig. S9). Importantly, these EGCG-induced aggregates are non-cytotoxic (Fig. 6A/C) and fail to associate with membranes or cells, in contrast to untreated PSMα3, which forms membrane-associated assemblies and induces disruption (newly added Movies S1-S2). Thus, EGCG potentially reduces cytotoxicity by remodeling the aggregation landscape and depleting active soluble species, rather than by selectively inhibiting specific fibril formation. This clarification is now added to the Discussion in lines 554-564 of the revised version.

(2) It is appreciated that the authors refrain from presenting the unsubstantiated concept of "functional" PSM amyloids in the discussion. However, wording in that direction must also be removed from other parts of the manuscript (e.g. "bioactive fibrillar polymorphs". "The formation of cross-alpha amyloids has been correlated with toxic activity", etc.), generally refraining from uncritically implying that amyloid formation underlies PSM biological activity, and rather discussing that the much more likely explanation of the findings is a lowering of cytolytically active, monomeric PSM concentration.

As detailed in our response to Major Comment #1, we agree that uncritical language implying that amyloid fibrils themselves are the cytotoxic species should be avoided. Accordingly, we have revised the manuscript to consistently frame amyloid formation in regulatory terms. Aggregation, depending on context, modulates activity by altering the availability, persistence, and assembly pathways of these species. Distinct aggregation states are therefore presented as correlated with, but not equivalent to, cytotoxic activity, and as components of a dynamic assembly landscape rather than as direct toxic entities.

(3) Discussion: "PSM alpha3 interaction with nucleic acids within human cells ...supports a comparable mechanism...". Please delete this as it is unsubstantiated.

We agree that the original phrasing overstated the evidence. The sentence was removed and the Discussion was revised to clearly frame nucleolar accumulation as a phenomenological observation reflecting PSMα3's intrinsic nucleic acid–binding capacity, rather than as evidence for a comparable intracellular mechanism. Specifically, the revised Discussion (lines 597–607) states that nucleolar localization is "unlikely to represent a distinct intracellular toxic mechanism" and instead "reflects binding competence within RNA-rich compartments following cellular entry." The biological relevance of this interaction, particularly at sub-cytotoxic concentrations, is noted as an open question requiring further investigation.

(4) The authors should also cite papers that have argued against their central hypothesis of "functional" PSM amyloids.

We thank the reviewer for this suggestion. Accordingly, we have revised the manuscript to explicitly cite and discuss studies that argue against amyloid fibrils as the primary cytotoxic species, and that instead attribute PSM cytotoxicity to soluble or membrane-associated forms. These perspectives are now incorporated in the Discussion to provide a balanced view of the field and to clarify how our findings align with, and differ from, existing models of PSM activity.

Version 2

Abstract

Amyloids, classically associated with neurodegenerative disease, also play important roles in infection and immunity. Phenol-soluble modulins (PSMs) from Staphylococcus aureus are amyloid-forming virulence peptides that contribute to cytotoxicity, immune modulation, and biofilm stability. PSMα3 forms cross-α amyloid fibrils and shares sequence and α-helical self-assembly features with LL-37, a human host-defence peptide that forms non-amyloid α-helical assemblies. Here, we identify RNA as a context-dependent regulator of their assembly pathways and biological activity. RNA consistently reduces LL-37 cytotoxicity toward human cells without compromising its antibacterial function, suggesting a host-protective effect. In contrast, RNA preserves PSMα3 cytotoxic and antimicrobial activity over time by reshaping its assembly landscape, promoting liquid–liquid phase separation at low concentrations and stabilizing dynamic α-helical intermediates. At higher RNA concentrations, both peptides transition into distinct aggregated states, amorphous for LL-37 and fibrillar for PSMα3, correlating with divergent functional outcomes. The amyloid inhibitor EGCG abolishes the bioactivity of both peptides by redirecting assembly into non-functional aggregates, highlighting that activity depends on supramolecular architecture and reversibility rather than aggregation per se. Together, these findings establish RNA as an environmental regulator of α-helical peptide assemblies and reveal phase transitions as tuneable determinants of peptide function, with implications for microbial virulence, innate immunity, and therapeutic intervention in infectious and protein-aggregation–associated diseases.

Introduction

Bacterial amyloids are proteins capable of forming ordered fibrils, playing a crucial role in biofilm stability, surface adherence, cell toxicity and immune responses1–6. These amyloids are vital for the survival and virulence of many bacteria, enabling them to colonize and persist in diverse environments7,8. Some examples of bacterial amyloids include the Csg and Fap proteins from Enterobacteria and Pseudomonas species1,6,7,9–11, respectively, which play a role in biofilm stability. Staphylococcal phenol-soluble modulins (PSMs) are short peptides that are capable of forming amyloid structures12–20 and perform multiple roles central to pathogenesis and immune evasion in Staphylococcus aureus21–36. In this context, defining PSMs as “functional amyloids” implies that amyloid formation can serve regulatory purposes, including controlled storage, stabilization, or spatial organization of bioactive molecules, as described for peptide hormones37 and RNA-binding proteins38,39. Accordingly, amyloid assembly may function as a reversible structural state that modulates the availability and timing of virulence-associated activities.

Among bacterial amyloids, PSMα3 is notable for its strong cytotoxicity toward a broad range of microbial and human cell types28,34,40–42 and for adopting a unique cross-α amyloid architecture, in which α-helices stack perpendicular to the fibril axis in paired sheets14–16. Recent studies on fibril-forming antimicrobial peptides have suggested a correlation between α-helical conformations in the solid state and cytotoxicity43–47. This observation parallels findings in human disease-associated amyloids, where α-helical oligomeric intermediates are widely considered to represent the toxic species48–50. Similarly, within the PSMα family, distinct amyloid polymorphs appear to encode different biological functions31: the metastable and reversible cross-α assemblies of PSMα3 enable dynamic switching between inactive and active states, whereas the more stable and irreversible cross-β fibrils formed by the homologous PSMα1 peptide19,51 may contribute to robust biofilm architecture. In this framework, toxicity is proposed to arise from transient assembly intermediates along the fibrillation pathway rather than from the mature fibrils themselves52.

PSMα3 shares self-assembly of α-helices with LL-37, a human host defence peptide53–55, even though LL-37 is not an amyloid and does not form canonical cross-α or cross-β fibrils. Moreover, the LL-37 antimicrobial active core (residues 17–29) shows sequence similarity to PSMα3, and the formation of non-amyloid fibrils composed of densely packed cationic amphipathic α-helices56. The sequence and α-helices self-assembly similarities supported the hypothesis that PSMα3 might recapitulate functions of LL-37, and indeed, they share the activation of the signal inhibitory receptor on leukocytes-1 (SIRL-1) which is a negative regulator of myeloid cell function and dampens antimicrobial responses57. Such common functions might be related to molecular mimicry and the ability of bacterial peptides to hijack immunomodulation pathways58. Indeed, PSMαs interact with the formyl peptide receptor 2 (FPR2) on immune cells, initiating immune responses such as chemotaxis, phagocytosis, and the secretion of pro-inflammatory cytokines. These reactions are critical for recruiting and activating immune cells, thereby amplifying the inflammatory response23,59–64.

LL-37 has been shown to bind RNA released from dying cells, forming stable complexes that protect the RNA from enzymatic degradation and promote its uptake by dendritic cells65. Once internalized, these RNA–LL-37 complexes activate Toll-like receptors (TLRs), triggering cytokine production and dendritic cell maturation65,66. Beyond LL-37, other human AMPs, such as β-defensins, also form higher-order assemblies that organize nucleic acids into nanostructures with defined periodicity. These nanocomplexes potently stimulate TLRs, highlighting that innate immune activation is influenced not only by ligand identity, but also by the nanoscale geometry and spatial organization of the complexes67–69.

Interestingly, nucleic acid interactions also play a key role in the immune activation triggered by bacterial amyloids. For example, PSMs co-assemble with extracellular DNA (eDNA), and CpG-rich DNA specifically promotes PSMα3 fibrillation, forming co-aggregates that colocalize with DNA in biofilms. These complexes activate TLR pathways and induce anti-dsDNA autoantibody responses in mice27. Similar mechanisms are seen in E. coli and Salmonella, where curli fibers bind eDNA, enhance fibrillation, and engage TLRs signaling, also promoting autoimmunity70,71. These findings suggest that amyloid–DNA complexes in bacterial biofilms are not only structural components but also potent immunomodulators with potential links to autoimmunity.

Given their shared cationic, amphipathic α-helical character, but distinct amyloidogenic properties, we sought to examine whether RNA differentially influences the assembly landscapes and bioactivity of PSMα3 and LL-37. RNA is increasingly recognized as a key regulator of protein phase behaviour, capable of promoting liquid–liquid phase separation (LLPS), altering aggregation pathways, and reshaping supramolecular organization. LLPS is a fundamental mechanism of cellular organization, in which biomolecules demix into dynamic, membraneless compartments that regulate processes such as RNA metabolism, stress responses, and signal transduction72–75. Whether bacterial virulence-associated peptides are similarly governed by RNA-mediated phase behaviour remains largely unexplored.

To address this, we investigated how defined RNA species modulate the structural transitions, material properties, and bioactivity of PSMα3 and LL-37. In parallel, we examined the impact of epigallocatechin gallate (EGCG), a small-molecule assembly modulator widely studied as an amyloid inhibitor32,76–78, to compare RNA-driven regulation with pharmacological redirection of peptide assembly. By integrating biophysical, structural, and cellular approaches, this work aims to define how environmental factors reshape α-helical peptide assemblies and thereby influence bacterial virulence and host-defence mechanisms.

Results

PSMα3 interacts with RNA in vitro

The interactions between PSMα3 and RNA were assessed using an Electrophoretic Mobility Shift Assay (EMSA), comparing the binding affinity of PSMα3 to single-stranded PolyA RNA and double-stranded Poly (AU) RNA. These RNAs were selected as simplified, well-defined model RNAs to probe general peptide–RNA interactions in an unbiased manner, as no prior information was available regarding whether such interactions occur, or which specific RNA species might be involved. The RNA molecules were labelled with the IR800CW fluorescent dye (as detailed in the Methods section). Freshly dissolved PSMα3, at varying concentrations, was incubated with each RNA type at a constant concentration for 30 minutes at 37°C. The EMSA results revealed that PSMα3 exhibits a stronger binding affinity to double-stranded Poly (AU) RNA, as evidenced by significant shifts at lower peptide concentrations compared to single-stranded Poly (A) (Fig. 1).

The binding and effect of single- and double-stranded RNA on PSMα3.

EMSA assay illustrating the interaction between PSMα3 and RNA. The assay compares the effects of increasing concentrations of PSMα3 (0 µM, 40 µM, 80 µM, 160 µM, and 320 µM), shown in a gradient from left to right, on the mobility shift of single-stranded Poly (A) RNA (left panel) and double-stranded Poly (AU) RNA (right panel) at ∼400ng/µL.

Quantification of free and shifted RNA band intensities revealed a progressive increase in the fraction of RNA bound as a function of peptide concentration for both RNA molecules (Fig. S1). Poly (AU) RNA reached saturation at lower PSMα3 concentrations than Poly (A), consistent with stronger apparent binding. Fitting the binding curves to a Hill model yielded an apparent dissociation constant (Kd) of 61.8 ± 1.6 µM for Poly (AU) and 94.5 ± 5.4 µM for Poly (A) (Fig. S1). Hill coefficients greater than 1 were obtained for both RNAs, consistent with cooperative binding behaviour. Based on these Kd differences, subsequent experiments focused on PSMα3 behaviour in the presence of double-stranded Poly (AU) RNA (hereafter referred to as RNA).

Non-monotonic turbidity response of PSMα3 to RNA

In order to assess the concentration-dependent effect of RNA on PSMα3 we used turbidity assays. The results showed that a 30 min co-incubation of 100 µM PSMα3 with varying concentrations of RNA displayed an increased turbidity up to 50 ng/µl, while higher RNA concentrations displayed a decrease in turbidity (Fig. S2). The decrease in particle size above 50 ng/µL can be a result of either the formation of smaller particles or aggregation into larger clusters, which can settle out of suspension, removing them from the light-scattering medium and reducing turbidity. Alternatively, this non-monotonic turbidity behaviour is conceptually similar to charge-driven re-entrant transitions described in other protein–RNA phase-separating systems, such as Ddx479, however, the molecular context and assembly architecture of PSMα3 are fundamentally distinct. According to the turbidity results, we hypothesized that 50 ng/µl can be a critical concentration between soluble species, which might be able to phase separate, and the formation of aggregates.

RNA concentration regulates formation and liquid-to-solid transition of PSMα3 condensates

To better understand the nature of the PSMα3-RNA interaction, we visualized their mixtures via fluorescence microscopy, using PSMα3 labelled with Fluorescein isothiocyanate (FITC) at the C-terminus (PSMα3-FITC), which is known to form fibrils14. All fluorescence microscopy experiments used a mixture of 20% FITC-labelled and 80% unlabelled peptide to minimize potential effects of the fluorophore on assembly behaviour. 100 µM PSMα3-FITC was incubated with either 50 ng/µl or 400 ng/µl RNA, labelled using propidium iodide (PI), in a close to physiological buffer of 50 mM HEPES (pH 7.4) containing 150 mM sodium chloride (NaCl).

Upon addition of 50 ng/µL RNA, PSMα3 formed liquid–liquid phase-separated droplets that clearly colocalized with RNA (Fig. 2A). We further observed encapsulation of RNA within PSMα3–FITC–coated condensates (Fig. S3), indicating co-assembly rather than simple surface association. The liquid-like nature of these droplets was confirmed by fluorescence recovery after photobleaching (FRAP), which showed rapid signal recovery shortly after bleaching (Fig. 2B). Quantitative analysis revealed a half-time of recovery of 3.1 seconds and a mobile fraction of 68% (immobile fraction 32%) (Fig. S4), consistent with dynamic molecular exchange. Across multiple samples, we consistently observed coexistence of small droplets and larger aggregates. The experimental timescales examined do not allow us to reliably determine whether diffusion-driven coalescence kinetics would support classical droplet ripening dynamics80,81. Notably, after 2 hours of incubation, fluorescence recovery was no longer detected (Fig. 2C), indicating that the condensates undergo time-dependent aging and transition toward a more solid-like or aggregated state.

Colocalization, droplet formation, and texture of PSMα3 mixed with varying Poly (AU) RNA concentrations.

Widefield fluorescence microscopy images of 100 µM of 20% FITC-PSMα3 (green) and 80% unlabelled PSMα3 incubated with Poly (AU) RNA (red) at 50 ng/µL (A) or 400 ng/µL (D), shown as individual fluorescence channels and merged images. (B-C) FRAP analysis of PSMα3-FITC in the presence of 50 ng/µL Poly (AU) RNA after 10 min (B) or 2 h (C) of co-incubation. (E) FRAP analysis of PSMα3-FITC in the presence of 400 ng/µL Poly (AU) RNA after 10 min of co-incubation. For panels B, C, and E, images were acquired before photobleaching, immediately after photobleaching, and 40 s post-photobleaching. (F) 100 µM of 20% FITC-PSMα3 (green) and 80% unlabelled PSMα3 in the absence of RNA. All scale bars represent 20 µm.

In contrast to the effect of 50 ng/µl RNA, the addition of a higher RNA concentration of 400 ng/µl to PSMα3 resulted in immediate aggregation, with no observed phase of droplet formation (Fig. 2D). FRAP analysis indicated no recovery of signal after bleaching, likely due to the formation of solid structures rather than droplets with liquid-like properties (Fig. 2E). In the absence of RNA, PSMα3 did not form condensates and instead appeared as dispersed or irregular assemblies (Fig. 2F), establishing that RNA is required to induce the droplet state observed under these conditions.

Colocalization between PSMα3-FITC and RNA was quantified using Pearson’s correlation coefficient (above the Costes threshold) and Manders’ overlap coefficients (tM1 and tM2) (Fig. S5). All metrics indicate substantially stronger spatial correlation and signal overlap at 50 ng/µL RNA compared with 400 ng/µL RNA. At low RNA concentration, Pearson’s correlation is positive (R ≈ 0.53), accompanied by high Manders’ overlap values (tM1 ≈ 0.68; tM2 ≈ 0.63). This is consistent with coordinated association of PSMα3 and RNA within liquid-like droplets. In contrast, at 400 ng/µL RNA, Pearson’s correlation becomes negative (R ≈ −0.24) and Manders’ coefficients are markedly reduced. This indicates a transition from co-localized droplet assemblies to spatially segregated aggregated structures in which RNA becomes partially excluded from the peptide-rich structures.

RNA concentration-driven changes in PSMα3 fibrillar morphology and aggregation

The TEM micrographs showed that in the absence of RNA, 100 µM PSMα3 formed nanotube-like fibrils after 2 hours, growing wider or with some twist after 24-hours co-incubation (Fig. 3A). Incubating PSMα3 with a low concentration of 10 ng/µL RNA displayed accelerated fibril formation, with more strongly twisted, wide, sheet-like fibrils observed after both 2 hours and 24 hours (Fig. 3A). With the addition of 50 ng/µL RNA, PSMα3 fibrils have a similar twisted morphology after 2 hours, while at a longer incubation time of 24 hours, we observed a significant morphological shift into more thin amorphous aggregates (Fig. 3A). At the higher concentration of 400 ng/µL RNA, PSMα3 formed dense, thin fibrils after 2 hours, but with a possible fragmentation into smaller species or rearrangement into amorphous aggregates after 24 hours of co-incubation. This indicates that RNA concentration and time of co-incubation affect the density and morphology of PSMα3 fibrils. This corresponds to the differences observed by light microscopy of co-aggregates contexture of LLPS droplets vs solid aggregates (Fig. 2).

TEM and TIRF visualization of PSMα3 aggregation and morphology with different Poly (AU) RNA concentrations and incubation times.

(A) TEM micrographs of 100 µM PSMα3 incubated with or without Poly (AU) RNA at varying concentrations of 10 ng/µL, 50 ng/µL, and 400 ng/µL for 2 hours (top row) and 24 hours (bottom row). Scale bars represent 500 nm. (B) TIRF microscopy images showing 100 µM of 20% FITC-PSMα3 (green) and 80% unlabelled PSMα3 co-incubated with 50 ng/µL Poly (AU) RNA and the amyloid indicator AT630 (magenta) for 30 minutes and 2 hours. Scale bars represent 20 μm. (C) TIRF microscopy images of 100 µM of 20% FITC-PSMα3 (green) and 80% unlabelled PSMα3 co-incubated with 400 ng/µL Poly (AU) RNA and AT630 (magenta) for 30 min. Scale bars represent 20 µm.

These observations were further supported by TIRF microscopy using the amyloid-specific dye AmyTracker630 (AT630). With the addition of 50 ng/µL RNA to PSMα3-FITC, a strong AT630 fluorescence signal was detected only after 2 hours of co-incubation, but not after 30 minutes (Fig. 3B, quantified in Fig. S6). This suggests a time-dependent transition into amyloid-like species. Conversely, at 400 ng/µL RNA, a significant AT630 fluorescence was observed already after 30 minutes of co-incubation, consistent with a rapid formation of amyloid fibrils (Figs. 3C and S6). These findings highlight a concentration- and time-dependent modulation of PSMα3 phase separation and structural transitions, where RNA promotes LLPS at lower concentrations and drives rapid amyloid formation and unique fibrillar morphologies at higher concentrations. Future work will be required to quantitatively define the phase boundaries and delineate the dominant mechanisms, such as sedimentation, dissolution, or coarsening/aging, across intermediate RNA concentrations.

RNA enhances α-helical structure of PSMα3

Solid-state circular dichroism (ssCD) spectroscopy reveals that PSMα3 lacks a defined secondary structure both immediately after preparation and following 2 hours of incubation (Fig. S7). However, an α-helical signature of PSMα3 is markedly enhanced in the presence of RNA compared to peptide alone, as evidenced by increased signal intensity, deeper minima, and more pronounced spectral features characteristic of α-helical structure. This enhancement is more pronounced at 400 ng/µL RNA than at 50 ng/µL, particularly after 2 hours of co-incubation, indicating that RNA concentration influences the stabilization of α-helical assemblies. This supports the notion that RNA not only accelerates aggregation but also promotes or stabilizes the α-helical fibrillar architecture, potentially consistent with cross-α amyloid structures.

RNA modulates the antibacterial activity of PSMα3

The antimicrobial activity of PSMα3 against Escherichia coli was evaluated under varying RNA concentrations and incubation times using the PrestoBlue Cell Viability Assay (Fig. 4A). Freshly dissolved 10 or 20 µM PSMα3 exhibited potent antibacterial activity, completely abolishing bacterial viability. After 2 hours of incubation, PSMα3 retained its full activity, comparable to its freshly prepared state. The presence of RNA at 50 ng/µL and 400 ng/µL had no impact on its antimicrobial function within this timeframe (Fig. 4A).

Impact of Poly (AU) RNA on PSMα3 cytotoxicity and antibacterial activity.

Antimicrobial activity of PSMα3 against E. Coli using the PrestoBlue cell viability assay (A) and its cytotoxicity against HeLa cells using the LDH colorimetric assay (B) were assessed with and without Poly (AU) RNA at varying concentrations. The experiments were performed in at least three replicates and repeated across three independent days to ensure result reliability. Cytotoxicity and bacterial cell viability percentages were calculated as the mean of all replicates, with error bars representing the standard error of the mean (SEM). Statistical significance was determined using one-way ANOVA for normally distributed data in GraphPad Prism (version 11). Significance levels are indicated as follows: *p<0.05, **p<0.01, ***p<0.001, ****p<0.0001.

In contrast, after 24 hours of incubation, PSMα3’s antibacterial activity was significantly reduced (Fig. 4A), suggesting a decrease in its effective concentration or changes in its morphology. Notably, the addition of RNA at 50 ng/µL or 400 ng/µL prevented this loss of activity, indicating that RNA plays a stabilizing role in toxic species or their reservoir. These findings suggest that RNA influences PSMα3 aggregation dynamics and morphology, thereby modulating its long-term antimicrobial effectiveness.

RNA modulates PSMα3 cytotoxicity against human HeLa cells

The effect of RNA on the cytotoxicity of PSMα3 against human HeLa cells was evaluated by measuring lactate dehydrogenase (LDH) release, an indicator of cell membrane damage (Fig. 4B). Freshly dissolved PSMα3 exhibited substantial cytotoxicity, causing approximately 80% cell death at 20 µM and 50% at 10 µM. The presence of RNA did not significantly alter the toxicity of freshly dissolved PSMα3 (Fig. 4B), similar to the antibacterial activity. However, following 2 hours of incubation, PSMα3 cytotoxicity was significantly reduced, possibly due to aggregation and a loss of active peptide concentration. This reduction occurred earlier than the loss of antibacterial activity, which was only observed after 24 hours of incubation (Fig. 4A).

Notably, co-incubation with RNA helped maintain PSMα3 cytotoxicity in a concentration-dependent manner. With 50 ng/µL RNA, the 2-hour incubated PSMα3 retained cytotoxicity comparable to its freshly dissolved form, suggesting that RNA prevents activity loss due to incubation. However, after 24 hours, 50 ng/µL RNA was insufficient to preserve cytotoxicity. In contrast, at 400 ng/µL RNA, partial cytotoxicity was maintained for both 2-hour and 24-hour incubated PSMα3, compared to the freshly dissolved sample.

Overall, these findings suggest that RNA prevents the incubation-induced loss of both antibacterial activity and cytotoxicity (Fig. 4). This effect appears to be concentration- and co- incubation time-dependent, likely linked to RNA-induced morphological variations of PSMα3 species (Figs. 2-3).

PSMα3 targets nucleolar nucleic acids in HeLa cells with LLPS dynamics

The interactions and localization of PSMα3 were next explored in HeLa cells with the addition of 20µM PSMα3 just before imaging. PI was added to the cell medium as a marker for cell death, as it selectively penetrates dead cells and binds to their nucleic acids.

Confocal microscopy images revealed that PSMα3 induces significant toxicity as indicated by membrane damage and blebbing, intracellular PSMα3 aggregate formation, and positive PI staining (Fig. 5 and Movie. S1). PSMα3 also enters the nuclei, with a notable concentration in the nucleolus, indicated by the foci of green fluorescence (marked by white arrows in Fig. 5). The composite images demonstrated clear colocalization of PSMα3 with nucleic acids in the nucleolus, as seen by the overlap of the green PSMα3-FITC signal and the red fluorescence from PI-stained nucleic acids.

Colocalization of PSMα3 with nucleic acids in HeLa cells.

(A) Confocal microscopy images showing the localization and colocalization of 20 µM of 20% FITC-PSMα3 (green) and 80% unlabelled PSMα3 and nucleic acids stained with PI (red) inside the nucleolus of HeLa cells (indicated in arrows). The left panel illustrates the distribution of PSMα3 within the cell. The middle panel shows the nucleic acids stained with PI. The right panel is a composite image that demonstrates the colocalization of PSMα3 with nucleic acids. Scale bars represent 15 µm. (B) FRAP analysis of 20 µM of 20% FITC-PSMα3 (green) and 80% unlabelled PSMα3 inside the nucleolus (indicated by the arrow) of HeLa cells, showing fluorescence recovery after 60 seconds. Scale bars represent 10 µm.

FRAP analysis of 20 µM PSMα3 within the nucleolus of HeLa cells revealed measurable but constrained mobility (Fig. 5B). Following photobleaching of a defined nucleolar region (indicated by the white arrow), fluorescence gradually recovered over the 60 s acquisition window, indicating dynamic exchange of PSMα3 within this compartment. However, recovery was slow and incomplete during this time frame, and did not reach a clear plateau, precluding reliable determination of recovery half-time or precise mobile and immobile fractions (Fig. S8), and indicating dynamic but constrained exchange within this RNA-rich compartment. This behaviour differs from the rapid and near-complete recovery observed for PSMα3–RNA droplets formed in vitro (Fig. 2C), indicating that nucleolar-associated PSMα3 exhibits more restricted dynamics than liquid-like condensates in defined buffer systems.

EGCG directly binds PSMα3 and inhibits its fibrillation and bioactivity, even in the presence of RNA

Since RNA appears to protect or maintain PSMα3’s toxic functions, potentially by influencing its fibril formation and morphology, we investigated the corresponding and combined effects of an inhibitor of fibril formation. One such inhibitor is epigallocatechin gallate (EGCG), the most abundant catechin in tea and a known amyloid inhibitor, including PSMs32.

The addition of EGCG to PSMα3 at a 1:1 molar ratio did not significantly alter its cytotoxicity against HeLa cells. However, at a fivefold molar excess, EGCG completely abolished PSMα3’s cytotoxic effect (Fig. 6A). To examine the structural changes underlying this effect, we analysed fibril formation kinetics and morphology in the presence of EGCG. TEM micrographs revealed that a fivefold molar excess of EGCG disrupted fibril formation of 100 µM PSMα3, instead inducing amorphous aggregates (Fig. 6B). Consistently, kinetics assays of fibril formation showed that EGCG inhibited the ThT fluorescence curve otherwise indicating the fibril formation of 100 µM PSMα3 (Fig. S9). These findings suggest that EGCG reduces PSMα3 toxicity by modulating its fibril formation and morphology. The addition of RNA did not counteract the effect of EGCG or restore PSMα3 cytotoxicity (Fig. 6A), highlighting the strong impact of EGCG on PSMα3 morphology and properties.

EGCG modulates PSMα3 aggregation and reduces toxicity against HeLa cells.

(A) Cytotoxicity of PSMα3 against HeLa cells in the presence and absence of EGCG and Poly (AU) RNA at two different concentrations, assessed via LDH assay. The experiment was performed in triplicate and repeated on three separate days for consistency. Cytotoxicity percentages were averaged across all replicates, with error bars representing the standard error of the mean. Statistical significance: *p < 0.05, **p < 0.01, ***p < 0.001, ****p < 0.0001 (one-way ANOVA, GraphPad Prism v11). (B) TEM micrographs of 100 μM PSMα3 incubated for 24 hours, without (left) and with (right) a fivefold molar excess of EGCG. Scale bars: 500 nm. (C) Live-cell confocal microscopy of HeLa cells treated with 20 μM of 20% FITC-PSMα3 (green) and 80% unlabelled PSMα3 without (top) and with EGCG (bottom), imaged immediately after preparation (t = 0) and after 10 minutes. Hoechst 33342 (blue) marks the nuclei, while PI staining (red) indicates membrane disruption. Scale bars: 10 μm.

Live-cell confocal microscopy provided further insights into how EGCG binding affects PSMα3 interactions with cells. In accordance with the LDH measurements (Fig. 6A), in the absence of EGCG, 20 µM PSMα3 readily penetrated cells and induced robust PI staining, indicating membrane disruption (Fig. 6C, top panel and Movie. S1). In contrast, a fivefold molar excess of EGCG prevented the toxic effects of 20 µM PSMα3, blocking cell penetration and PI staining (Fig. 6C, bottom panel and Movie. S2). Instead, PSMα3-FITC formed extracellular aggregates and showed no interaction with cell membranes, as indicated by the absence of PI staining and intact cell membranes.

Furthermore, EGCG at a fivefold molar excess also reduced the antibacterial activity of PSMα3 against E. coli, maintaining bacterial cell viability (Fig. S10). Super-resolution fluorescence microscopy showed that while 20 µM PSMα3 typically aggregates on the bacterial membrane, causing membrane disruption and PI staining indicative of cell death (Fig. S10), EGCG at a fivefold molar excess prevented membrane aggregation and PI staining, thereby preserving membrane integrity.

To investigate residue-specific interactions between PSMα3 and EGCG, we performed NMR spectroscopy using a 2:1 PSMα3:EGCG molar ratio. The 1D 1H-NMR spectrum revealed distinct chemical shift changes and peak broadening upon EGCG addition, particularly in residues Met1, Glu2, and V4/N21, suggesting specific interactions between these sites and the EGCG molecule (Fig. 7A). Peak broadening was also observed in the aromatic region of EGCG, especially for protons H1 and H2, compared to the reference spectrum of EGCG alone, indicating that the interaction is visible from the EGCG side as well. Notably, slight opalescence was observed in the sample following preparation, potentially reflecting early-stage aggregate formation. Additionally, the presence of multiple cross-peaks between non-sequential residues (i+3 or i+4) in the two-dimensional (2D) 1H-1H Nuclear Overhauser Effect Spectroscopy (NOESY) spectrum indicates a well-defined structure of PSMα3 in this condition (Fig. 7B). The 2D 1H-1H Total Correlation Spectroscopy (TOCSY) spectrum displayed connectivity between backbone amide (HN) and alpha protons (Hα) for several assigned residues, while the 2D 1H-1H NOESY spectrum revealed spatial correlations between nearby residues (i+1).

Residue-specific interactions between PSMα3 and EGCG.

(A) One-dimensional (1D) 1H NMR spectra of 1.0 mM PSMα3 alone (red) and in complex with 0.5 mM EGCG (blue), recorded at 35 °C. Specific residues, including Met1, Glu2 and V4/N21, show chemical shift changes suggestive of direct interaction with EGCG (highlighted in the upper left). Dashed boxes mark proton signals corresponding to EGCG. Peak broadening of H1/H2 protons, compared to the EGCG-only reference sample (light grey), indicates interaction from the EGCG side. Slight opalescence observed in the sample suggests potential aggregate formation. (B) Two-dimensional (2D) 1H–1H TOCSY and NOESY spectra of the PSMα3:EGCG complex at a 2:1 ratio, recorded at 35 °C. The cross-peaks in the TOCSY spectrum allow the identification of the spin system and direct connections through the scalar coupling between the proton amide (HN) and the alpha protons (Hα) of the same residue. The cross-peaks in the NOESY spectrum establish a sequential connection between neighbouring residues (i+1), allowing spectral assignment. The remaining cross-peaks (i+3 or i+4) in the spectrum support PSMα3 secondary structure in the experimental conditions. (C) Temporal stability of the PSMα3:EGCG sample over 3 days.

To assess the temporal stability of the sample, we monitored signal intensities over a 3-day period. PSMα3 signals remained stable, with only a ∼10% decrease in intensity, whereas EGCG signals exhibited substantial degradation, with nearly 50% loss over the same time frame (Fig. 7C). Given the ∼1.5-day duration of the 2D NMR measurements, it is estimated that over 60% of EGCG remained in solution during data acquisition, allowing for reliable observation of its interaction with the peptide. Of note, EGCG’s activity has been shown to depend on its chemical stability and the surrounding conditions. Specifically, at neutral pH, EGCG may undergo oxidation, and its inhibitory effects could be attributed to its degradation products rather than the intact compound itself82.

RNA modulates the human host defence peptide LL-37 phase behaviour and cytotoxicity without compromising antimicrobial activity

LL-37, a human host defence peptide, shares similarities with PSMα3, including its ability to self-assemble into α-helical supramolecular structures and to interact with nucleic acids56,65. We therefore examined whether RNA similarly influences LL-37 activity and properties.

Fluorescence microscopy showed that 100 µM LL-37-FITC incubated with 100 ng/µL PI-labelled Poly (AU) RNA formed aggregates, with no evidence of droplet formation or LLPS under these conditions (Fig. S11). Because diverse cellular stresses are known to induce condensation of RNA-binding proteins and promote the formation of stress-associated assemblies83, we next examined the effect of RNA on LL-37 under thermal stress. Following heat shock at 65 °C for 15 minutes, RNA markedly altered LL-37 assembly behaviour in a concentration-dependent manner. At 100 ng/µL RNA, LL-37 formed rounded condensates that colocalized with RNA (Fig. 8A). To assess the material properties of these assemblies, we performed FRAP analysis on LL-37–FITC under this condition. Photobleaching resulted in minimal fluorescence recovery over the acquisition window (Fig. S12), indicating that the LL-37–RNA assemblies formed after heat shock are predominantly solid-like rather than liquid-like. At 200 ng/µL RNA, LL-37 assemblies exhibited mixed features, with rounded structures still present but accompanied by increasing aggregation, consistent with a transition toward a more solid state (Fig. 8B). At 400 ng/µL RNA, discrete condensates were no longer observed and LL-37 instead formed extensive, irregular aggregates with strong RNA colocalization (Fig. 8C).

Effect of RNA concentration on LL-37 phase separation, aggregation, and activity at pH 7 after heat shock.

Fluorescence microscopy images showing 20% FITC-labelled (green) and 80% unlabelled 100 μM LL-37 in the presence of increasing concentrations: 100 ng/μL (A), 200 ng/μL (B) and 400 ng/μL (C) of Poly (AU) RNA (PI-labelled, red) after heat shock at 65°C for 15 minutes at pH 7. Scale bars represent 20 µm. (D) LL-37 cytotoxicity, with and without Poly (AU) RNA and EGCG at varying concentrations, was assessed in HeLa cells using the LDH colorimetric assay. The experiment was performed in triplicate and repeated on three separate days for reproducibility. Cytotoxicity percentages represent the average of all replicates, with error bars indicating the standard error of the mean. (E) The antimicrobial activity of LL-37, with and without Poly (AU) RNA and EGCG at varying concentrations, was evaluated against E. coli using the PrestoBlue Cell Viability assay. The experiment was performed in triplicate and repeated on three separate days for reproducibility. Bacterial viability percentages represent the average of all replicates, with error bars indicating the standard error of the mean. Statistical significance (D-E): *p < 0.05, **p < 0.01, ***p < 0.001, ****p < 0.0001 (one-way ANOVA, GraphPad Prism v11).

For comparison, PSMα3 subjected to the same heat-shock protocol exhibited droplet formation only under more permissive conditions (pH 4), but not at physiological pH (7.4) (Figs. S13-S14). Together, these results indicate that, for both LL-37 and PSMα3, increasing RNA concentration favours aggregation over phase separation, and that LL-37 assemblies formed under thermal stress rapidly adopt solid-like properties rather than maintaining liquid-like dynamics.

Freshly dissolved LL-37 exhibited cytotoxicity against HeLa cells, which, in contrast to PSMα3, remained consistent even after incubation for up to 24 hours. Notably, RNA significantly reduced LL-37 cytotoxicity in a concentration-dependent manner, irrespective of the incubation duration (freshly dissolved, 2 hours, or 24 hours) (Fig. 8D). This contrasts with RNA’s rescuing effect on the incubation time-dependent loss of PSMα3 cytotoxicity. The antimicrobial activity of LL-37 against E. coli also remained unchanged upon incubation. However, unlike its effect on cytotoxicity, RNA had minimal impact on antibacterial activity, with only high concentrations of 400 ng/µL reducing LL-37 activity (Fig. 8E). Similar to PSMα3, LL-37 cytotoxicity against HeLa cells and its antimicrobial activity were attenuated by EGCG in a concentration-dependent manner (Fig. 8E).

TEM micrographs of 100 µM LL-37 revealed a variety of aggregative and fibrous structures, including thin, curli-like fibrils, although no consistent morphology was observed (Fig. S15A). RNA induced a distinct morphological change, leading to increased aggregation and the formation of mostly amorphous species, but also some thicker, ribbon-like fibrils were observed. These structures resemble those previously observed for a segment of the LL-37 active core (residues 17–29), which is similar to PSMα3 in sequence and its ability to form fibrils of densely packed amphipathic α-helices56. Heat shock further intensified aggregation, resulting in denser, amorphous condensates and larger fibrillar assemblies, with and without RNA (Fig. S15B). The distinct effects of RNA on the aggregation and morphologies of LL-37 and PSMα3 may underlie the observed differences in their cytotoxic and antibacterial activities.

Discussion

This study identifies RNA as a context-dependent regulator of α-helical assembly-prone peptides, demonstrating that RNA reshapes their supramolecular landscape and thereby tunes bioactivity rather than merely sequestering peptide species. This is supported by the emergence of distinct RNA-dependent material states (from droplets to aged solids) and RNA-dependent fibrillar morphologies with divergent activity profiles across time and concentration. By comparing the bacterial virulence factor PSMα3, which forms canonical cross-α amyloid assemblies, with the human host-defence peptide LL-37, which forms α-helical fibrils but lacks amyloid-like stacking, we reveal that RNA differentially modulates structurally distinct systems through peptide-specific effects on phase behaviour and aggregation dynamics.

Toxicity and antibacterial activity arise from dynamic intermediates

Our data support a model in which PSMα3 toxicity does not arise from a single static species, such as monomers or mature fibrils, but from dynamic assembly processes that include transient α-helical assembly intermediates along the fibrillation pathway. While soluble species mediate cytotoxic activity, mutants that retain α-helical monomeric structure yet fail to fibrillate lose toxicity14,15, and fully matured fibrils are largely inactive84,85. This apparent paradox is resolved by proposing that the toxic window resides within dynamic, membrane-associated assembly intermediates52. In this framework, membrane-associated fibril growth may destabilize bilayers through lipid recruitment and extraction86, analogous to mechanisms proposed for certain human amyloids such as islet amyloid polypeptide87.

RNA preserves PSMα3 bioactivity by regulating this toxic window and redirecting its assembly pathway

In the absence of RNA, PSMα3 undergoes time-dependent aggregation31, accompanied by progressive loss of cytotoxic and antimicrobial activity (Fig. 4), coinciding with the formation of dense, fibrillar assemblies (Figs. 2-3). These mature aggregates likely represent inert end states that sequester membrane-active intermediates84,85. RNA reshapes this trajectory in a concentration-dependent manner. At low RNA concentrations, PSMα3 initially retains activity (Fig. 4B) but gradually loses it as liquid-like RNA–PSMα3 condensates mature into more rigid assemblies (Figs. 2-4 & S3), indicating that RNA regulates the lifetime of dynamic intermediates. At higher RNA concentrations, RNA promotes alternative fibrillar polymorphs that retain substantial cytotoxic activity even after prolonged incubation (Fig. 4B). Similarly, RNA preserves antibacterial potency over time (Fig. 4A), consistent with stabilization of assembly states that remain competent for membrane interaction. Thus, RNA does not simply suppress aggregation but modulates the reversibility and supramolecular architecture of α-helical assemblies (Fig. S7), thereby controlling the duration and accessibility of the toxic window.

The temporal differences between cytotoxicity (Fig. 4B) and antibacterial activity (Fig. 4A) may reflect distinct membrane environments, aggregation kinetics, or interactions with bacterial surfaces versus mammalian membranes. Nonetheless, in both contexts, activity correlates with assembly state rather than peptide abundance alone. RNA therefore acts as a structural modulator that controls the lifetime and reversibility of functionally relevant intermediates.

Regulatory modulation versus irreversible inhibition

The contrasting effects of RNA and the amyloid inhibitor EGCG highlight that bioactivity is governed by supramolecular architecture and reversibility rather than aggregation per se. RNA preserves PSMα3 bioactivity by stabilizing dynamic, α-helical assembly states. In contrast, EGCG binds soluble PSMα3 and diverts it into stable amorphous aggregates that represent structural dead ends, abolishing activity (Figs. 6-7). Thus, EGCG should not be viewed as evidence that amyloid fibrils are the toxic species; instead, it abolishes activity by removing/redirecting soluble bioactive species into non-productive amorphous assemblies. Overall, RNA functions as a regulatory modulator of assembly dynamics, whereas EGCG enforces irreversible pathway diversion. These results underscore that assembly state, not simply peptide presence, determines biological function.

LL-37: non-amyloid α-helical assemblies under RNA control

In contrast to PSMα3, LL-37 does not form canonical amyloid architectures. Although it can assemble into α-helical fibrillar structures53,56, these lack the ordered cross-β or cross-α stacking characteristic of amyloids. Nevertheless, RNA strongly modulates LL-37 behaviour. RNA can selectively attenuate cytotoxicity while preserving antimicrobial function, correlating with the formation of amorphous RNA–LL-37 co-assemblies rather than structured fibrils (Figs. 8, S11 & S15). RNA-dependent condensate-like assemblies are observed only under thermal stress, and these rapidly adopt solid-like properties (Figs. 8 and S12). For LL-37, RNA therefore appears to serve a protective role, minimizing collateral host damage while maintaining immune defence. This suggests that RNA-mediated modulation is a broader principle of α-helical peptide regulation, yet with a functional divergence that may reflect differences in the architecture of the resulting co-assemblies: RNA drives PSMα3 into structured α-helical fibrillar polymorphs, and LL-37 into disordered, amorphous co-aggregates.

Interestingly, although LL-37 does not adopt canonical amyloid architectures, it displays multiple functional connections to amyloid systems. LL-37 has been reported to bind the Alzheimer’s-associated amyloid-β (Aβ) peptide and inhibit its fibrillization, likely through stabilization of prefibrillar intermediates88. This observation suggests that LL-37 can influence amyloid assembly pathways rather than simply interacting with mature fibrils, with potential implications for aging, immune regulation, and infection. In addition, LL-37 has been shown to interact with human α-synuclein as well as with bacterial curli amyloids, where it modulates biofilm properties89–91. Together, these findings position LL-37 as a cross-reactive regulator of amyloid assemblies across host and microbial contexts. The mechanistic basis and physiological relevance of these interactions remain to be further elucidated.

Parallels with phase-transition biology

The dynamic phase transitions observed for PSMα3 mirror those described for human RNA-binding proteins such as FUS, TDP-43, hnRNPA1, and tau, where LLPS enables physiological function but aberrant maturation contributes to neurodegenerative diseases73,74,92–103.

Antimicrobial peptides can also induce LLPS with nucleic acids and disrupt critical processes like transcription and translation104. While the antimicrobial LL-37 is not an amyloid, its RNA-dependent co-assembly behaviour reinforces the broader concept that nucleic acids can regulate peptide phase behaviour across structurally distinct systems.

The nucleolar accumulation of PSMα3 (Fig. 5) is consistent with its intrinsic nucleic acid – binding capacity and provides cellular continuity with the RNA-dependent assemblies observed in vitro (Figs. 2-3). However, given the rapid membrane disruption induced by PSMα3 at cytotoxic concentrations (Figs. 5-6 and Movies S1-S2), nucleolar localization is unlikely to represent a distinct intracellular toxic mechanism. Rather, it reflects binding competence within RNA-rich compartments following cellular entry. The slow and incomplete fluorescence recovery observed in FRAP measurements (Fig. S8) further supports a dynamic yet constrained exchange of PSMα3 within the nucleolus, indicative of interaction with endogenous nucleic acids rather than freely diffusing peptide. The biological relevance of this interaction, particularly at sub-cytotoxic concentrations where membrane disruption does not dominate, remains to be determined and warrants further investigation.

Biological and therapeutic implications

Collectively, these findings position RNA as an environmental regulator of α-helical peptide assemblies, tuning bioactivity by modulating supramolecular organization and reversibility. Nucleic acids are abundant in infection-relevant settings where PSMs operate, including biofilms (rich in extracellular DNA and RNA), abscesses, neutrophil extracellular traps, and damaged tissues that release host RNA. In such contexts, RNA could control the persistence and timing of membrane-active PSM species by shifting assemblies between condensate-like reservoirs and more inert aggregates, thereby modulating virulence without invoking an intracellular mechanism. The divergent outcomes of RNA-mediated modulation-preservation of PSMα3 virulence versus attenuation of LL-37 cytotoxicity thus reflect peptide-specific assembly architectures rather than a uniform effect: RNA sustains dynamic, membrane-competent assemblies for the amyloidogenic PSMα3 while driving the non-amyloidogenic LL-37 into inactive amorphous states, effectively tilting the host-pathogen balance toward bacterial advantage.

While Poly (AU) served here as a defined, reductionist probe of peptide–RNA interactions, multiple endogenous RNA pools are plausible physiological partners. Within S. aureus biofilms, extracellular RNA is an abundant matrix component that contributes structurally to biofilm organization, raising the possibility that it directly modulates PSM assembly and thereby impacts biofilm stability and virulence-factor availability105,106. At infection sites, host-derived extracellular RNA released from stressed or dying cells can function as a danger-associated molecular pattern (DAMP)107,108 and may further shift the balance between PSMα3-driven cytotoxicity and LL-37–mediated host protection within inflammatory milieus. Defining which endogenous RNA species, concentrations, and structural features govern these assembly-dependent effects in vivo is therefore an important next step.

For PSMα3, RNA-sensitive assembly dynamics may provide a means for S. aureus to tune virulence in response to local cues, supporting immune evasion, biofilm-associated persistence, and survival across diverse host environments. For LL-37, RNA-mediated attenuation of cytotoxicity without loss of antimicrobial function suggests a host-protective mechanism that helps balance pathogen control with limiting collateral tissue damage. More broadly, this work shifts the conceptual emphasis from amyloid formation as a static endpoint to environmentally regulated phase/assembly transitions and shows that RNA-mediated control extends beyond canonical amyloids to non-amyloid α-helical assemblies. These insights motivate therapeutic strategies aimed at reshaping assembly pathways and material states, rather than simply blocking aggregation.

Methods

Peptide and Poly (AU) RNA preparation

Unlabelled and C-terminal FITC-labelled PSMα3 and LL-37 (PSMα3-FITC and LL-37-FITC) (>98% purity) were purchased from GL Biochem (Shanghai) Ltd. Poly (AU) RNA was purchased from Sigma Aldrich. A stock solution of the peptides was prepared at a concentration of 1 mM in a mixture of 20% dimethyl sulfoxide (DMSO) and 80% ultra-pure water. For light microscopy experiments, we used a mixture of 20% FITC-labelled and 80% unlabelled PSMα3 or LL-37. The Poly (AU) RNA stock solution was prepared at a concentration of 2000 ng/µl in UPW, the stock solution was stored at −80 °C until further use. Labelled Poly (AU) RNA was prepared by introducing 0.02 mg/ml of PI dye to the Poly (AU) RNA stock solution. For the experiments, samples containing 100 µM PSMα3 or LL-37 with Poly (AU) RNA at 10, 50, 100, 200, or 400 ng/µl were prepared in 50 mM HEPES buffer with 150 mM sodium chloride (NaCl), adjusted to pH 7.4, or in 20 μM Tris, 20 μM Bis-Tris, and 20 μM sodium acetate adjusted to pH 4.

Cytotoxicity against HeLa cells tested using the lactate dehydrogenase (LDH) release assay

Human cervical carcinoma HeLa cells (ATCC® CCL-2™) were routinely cultured in Dulbecco’s Modified Eagle’s Medium - high glucose (DMEM) (Sigma, Israel) with L-glutamine and supplemented with penicillin (100 U/ml), streptomycin (0.1 mg/ml), and with 10% foetal calf serum (Sigma, Israel). The cells grown at 37°C and 5% CO2. One day before the experiment, cells were resuspended in growing medium (DMEM supplemented with 10% foetal calf serum) 1X105 cells/ml. 50 μL of cell suspension were pipetted into a 96-well plate and grown over night. Thirty minutes before the experiment cells were washed and resuspended in 50 μL of DMEM medium supplemented with L-glutamine (100 U/mL), and with penicillin (100 U/mL), streptomycin (0.1 mg/mL), and with 0.5% heat-inactivated foetal calf serum (assay medium). For the cytotoxicity assay, 100 µM PSMα3 was incubated with or without EGCG at molar ratios of 1:1 or 1:5, or with Poly (AU) RNA at concentrations of 50 ng/µL or 400 ng/µL, for the designated incubation times. Similarly, 100 µM LL-37 was prepared under the same conditions, with or without EGCG at 1:1 or 1:5 molar ratios, or Poly (AU) RNA at 100 ng/µL or 400 ng/µL, following the specified incubation times.

Serial two-fold dilutions in assay medium were performed, and 50 μL of each dilution were pipetted into the three different 96-well with the cells. The plates were incubated for 30 min at 37°C and 5% CO2 and then cell lysis was quantified using the LDH release colorimetric assay according to the manufacturer’s instructions, including all recommended controls (LDH; Cytotoxicity Detection Kit Plus, Roche Applied Science, Germany). Cell-free assay medium was measured as background. Cells subjected to the same experimental conditions apart from peptide addition were used as a control to account for spontaneous LDH release. Cells subjected to the same experimental conditions apart from peptide addition and treated with manufacturer-supplied lysis buffer were used as a control to account for maximum LDH release. Absorbance at 490 and 690 nm was measured in a plate reader (FLUOstar Omega, BMG Labtech, Germany). Absorbance at 690 nm was subtracted from 490 nm readings to correct for background. The mean absorbance of triplicate samples and controls was calculated, followed by background subtraction. Data was obtained from at least three independent biological replicates, with the arithmetic mean used for averaging. Error bars represent the standard deviation (SD). Statistical analysis, including one-way ANOVA, was performed using GraphPad Prism 10.

Assessment of Bacterial Viability Using PrestoBlue™ HS Cell Viability Reagent

Escherichia coli (E. coli) cultures were grown overnight in Luria Broth (LB) medium at 37 °C with shaking at 220 rpm. PSMα3 and LL-37 stock solutions were prepared by dissolving the peptides at a concentration of 1 mM in a solvent mixture of 20% DMSO and 80% ultra-pure water. From these stock solutions, the following sample preparations were made:

For PSMα3: (1)100 µM PSMα3 (2)100 µM PSMα3 with 50 ng/µL Poly (AU) RNA (3) 100 µM PSMα3 with 400 ng/µL Poly (AU) RNA (4) 100 µM PSMα3 with 500 µM EGCG. For LL-37: (1)100 µM LL-37 (2)100 µM LL-37 with 100 ng/µL Poly (AU) RNA (3) 100 µM LL-37 with 400 ng/µL Poly (AU) RNA (4) 100 µM LL-37 with 500 µM EGCG.

All solutions were prepared in 50 mM HEPES buffer pH 7.4 containing 150 mM NaCl. Where applicable, samples were incubated for the specified durations of 2 hours or 24 hours.

On the day of the experiment, dilutions of the PSMα3 solutions to 20 µM and 10 µM were prepared in LB medium and dispensed into a sterile 96-well black flat-bottom plate (Greiner bio-one). E. coli cultures were diluted to an optical density (OD) of 0.2, and bacterial suspensions were added to the wells. The plate was incubated at 37 °C with shaking at 220 rpm for 1 hour to allow the reaction to proceed.

Following incubation, 10X PrestoBlue™ HS Cell Viability Reagent (Invitrogen) was added to the wells. Wells containing only LB medium served as negative controls, while wells containing only E. coli served as positive controls. Plates were sealed with a thermal seal film (EXCEL Scientific) and incubated in a plate reader (CLARIOstar). Bacterial viability was assessed using PrestoBlue fluorescence (excitation: 535–560 nm; emission: 590–615 nm) over time. Each condition was tested in triplicate across three independent experiments. Fluorescence values were averaged, and blank readings were subtracted. Antimicrobial activity was determined at the 2-hour time point, when fluorescence was highest in the positive control. Data were obtained from at least three independent biological replicates, with the arithmetic mean used for averaging. Error bars represent the standard deviation (SD). Statistical analysis, including one-way ANOVA, was performed using GraphPad Prism 10.

Transmission electron microscopy (TEM)

For TEM analysis, 4–5 µL of 100 µM PSMα3, with or without EGCG or Poly (AU) RNA at concentrations of 10 ng/µL, 50 ng/µL, and 400 ng/µL, as well as 100 µM LL-37, with or without Poly (AU) RNA at 100 ng/µL and 400 ng/µL, were directly applied onto glow-discharged 400-mesh copper grids (easiGlow; Pelco, Clovis, CA, USA) with a grid hole size of 42 µm, stabilized with Formvar/carbon (Ted Pella, Inc.). The grids were glow-discharged using a 15-mA current with a negative charge for 25 seconds. Samples were allowed to adhere to the grids for 45 seconds before being stained with a 1% uranyl acetate solution (Electron Microscopy Science, 22400-1) for 45 seconds. The samples were then examined using a ThermoFisher Scientific (FEI) Talos F200C transmission electron microscope, operating at 200 kV and equipped with a Ceta 16M CMOS camera, at the Ilse Katz Institute for Nanoscale Science and Technology, Ben Gurion University of the Negev, Israel.

Turbidity measurements

Turbidity measurements were performed using a FLUOstar Omega plate reader (BMG Labtech) set to a wavelength of 400 nm. For each sample, the protein concentration was maintained at 100 µM, while RNA concentrations were varied across the following levels: 10 ng/µl, 20 ng/µl, 50 ng/µl, 100 ng /µl, 200 ng/µl, and 400 ng/µl. The turbidity of the mixtures was monitored over several hours following resuspension. The maximum absorbance recorded within the first 30 minutes was used as an indicator of the dense phase volume. The experiments were conducted three times with similar observations.

Electrophoretic mobility shift assay (EMSA)

RNA molecules were synthesized by IDT, with oligoA modified by the IRDye 800CW fluorescent dye. Double-stranded RNA was prepared by annealing 10 µM oligoA-IRDye 800CW with 10 µM oligoU in an annealing buffer containing 10 mM Tris-HCl (pH 7.5), 50 mM NaCl, and 1 mM EDTA. The mixture was heated to 95°C for 5 minutes, followed by gradual cooling to 25°C at a rate of 2°C per minute. The final RNA concentration used in the binding reactions was 40 nM (∼400ng/µL).

RNA-protein complexes were incubated at 37°C for 30 minutes before being resolved on a 2.5% agarose gel. Gel electrophoresis was performed in a 0.5× TBE buffer for 7 minutes. The complexes were visualized using a Li-Cor Odyssey FC imaging system in the 800 nm channel with a 30-second exposure. For the EMSA, binding reactions were carried out in a buffer composed of 20 mM HEPES (pH 7.9), 50 mM KCl, 1 mM Dithiothreitol (DTT), 0.1 mM Ethylenediaminetetraacetic acid (EDTA), 5% glycerol, and 0.05% nonyl phenoxypolyethoxylethanol (NP-40). PSMα3 was added to the reactions at final concentrations of 0, 40, 80, 160, and 320 µM. The experiments were conducted three times with similar observations.

Fluorescent microscopy imaging of peptides with RNA in vitro

Prior to imaging, 100 µM PSMα3 or LL-37 (containing 20% FITC-labelled and 80% unlabelled peptide) were combined with PI-labelled Poly (AU) RNA at different concentrations in 50 mM HEPES buffer pH 7.4 containing 150 mM NaCl. Sample were also tested after 65°C heat shock for 15 minutes. In addition, the PSMα3 and RNA mixture was also tested in a different buffer containing 20 μM Tris, 20 μM Bis-Tris, and 20 μM sodium acetate at pH 4 and were subjected to heat shock at 65°C for 15 minutes.

Following preparation, with or without heat shock, a 10 µL aliquot of each sample was transferred to a µ-Slide 8 Well ibidi chamber or to a clean 24 × 60 mm No. 1.5 glass slide for imaging. Fluorescence microscopy was performed using a Leica DMI8 inverted fluorescent microscope equipped with a 63× immersion oil objective (Numerical Aperture, NA = 1.4) at the Life Sciences and Engineering (LS&E) Infrastructure Center, Technion-Israel Institute of Technology, Haifa, Israel. Data were processed and analyzed using Fiji-ImageJ software. The experiment was conducted three times with similar observations.

Fluorescence recovery after photobleaching (FRAP) measurements

FRAP experiments were conducted using a Zeiss LSM 710 laser scanning confocal microscope equipped with a 63× Plan-Apochromat oil immersion objective, NA1.4, at the Life Sciences and Engineering (LS&E) Infrastructure Center, Technion-Israel Institute of Technology, Haifa, Israel. The experiments were performed on condensates from by 100 µM PSMα3 or LL-37 (containing 20% FITC-labelled and 80% unlabelled peptide) mixed with PI-labelled Poly (AU) RNA at various concentrations as indicated in the figures. Photobleaching was performed using 405 nm and 488 nm laser lines. The experiments were conducted three times with similar observations.

For FRAP analysis inside the nucleolus of HeLa cells, PSMα3 was added to the cells at a final concentration of 20 µM containing 20% FITC-labelled and 80% unlabelled peptide. The cells were incubated with the peptide for 20 minutes before conducting the experiment. Data was processed and analysed using ZEISS ZEN software and Fiji-ImageJ software. The experiments were conducted three times with similar observations.

Total internal reflection fluorescence (TIRF) microscopy

TIRF microscopy was performed using a ZEISS Elyra 7 Super-Resolution Microscope equipped with a 63× Apochromat alpha Plan-Apochromat Oil immersion DIC objective, NA1.46, at the Life Sciences and Engineering (LS&E) Infrastructure Center, Technion-Israel Institute of Technology, Haifa, Israel. Images were acquired using the pco.edge sCMOS cameras, and data were processed and analysed with ZEISS ZEN software to accurately represent the observed phenomena. The condensates and aggregates analysed were generated using 100 µM of 20% FITC-PSMα3 (green) and 80% unlabelled PSMα3 with either 50 ng/µL or 400 ng/µL Poly (AU) RNA. AmyTracker 630 (Ebba Biotech) was added at a 1:500 (peptide: AmyTracker) molar ratio to each sample. The experiments were conducted three times with similar observations.

Confocal microscopy visualization of PSMα3 interaction with HeLa cells

HeLa cells were pre-cultured one day before the experiment by preparing a suspension containing 350,000 cells/ml and plating 150 µl of this suspension into each well of a µ-Slide 8 well glass-bottom chamber. The cells were then incubated overnight under standard growth conditions (37°C, 5% CO2) to allow for adherence and growth. On the day of the experiment, the cells were washed three times with phosphate-buffered saline (PBS) to remove any residual media. Hoechst 33342 dye (10 mg/mL stock) was diluted 1:2000 in fresh cell media and added to the cells. The cells were incubated with Hoechst for 10 minutes at 37°C and 5% CO2. After incubation, cells were washed three times with PBS to remove Hoechst residuals. A working solution of PI was prepared by diluting a 1 mg/ml PI stock solution to a final concentration of 0.02 mg/ml in fresh cell growth media. Immediately prior to imaging, PSMα3 was added to the cells at a final concentration of 20 µM containing 20% FITC-labelled and 80% unlabelled peptide. The cells were then imaged using a Ti2-E microscope by Nikon with a CSU-W1 spinning disk confocal unit by Yokogawa, equipped with a 100X CFI SR HP Plan Apochromat Lambda S silicone immersion objective, NA1.35, at the Life Sciences and Engineering (LS&E) Infrastructure Center, Technion-Israel Institute of Technology, Haifa, Israel. Confocal movies and images were captured by Photometrics BSI sCMOS cameras to observe the interaction and localization of PSMα3 within the cells, focusing particularly on its colocalization with nucleic acids in the nucleolar region. The acquired images and movies were subsequently analysed using Imaris Image Analysis Software (Oxford Instruments). The experiments were conducted three times with similar observations.

Super resolution light microscopy visualization of PSMα3 interaction with E. coli

E. coli lptD4213 cells were a kind gift from Prof. Sima Yaron from the Technion – Israel Institute of Technology. An inoculum was pre-cultured in Luria-Bertani medium (LB) medium at 220 rpm and 37°C for 24 hours prior to the experiment. On the day of the experiment, the optical density (OD) of the culture was measured, and the cells were diluted to an OD of 0.4. The cells were then incubated with 20 μM of 20% FITC-PSMα3 (green) and 80% unlabelled PSMα3, with and without EGCG at 1:5 molar ratio, alongside a bacteria-only control. Samples were incubated for 30 minutes at 220 rpm and 37°C. Following incubation, the samples were centrifuged three times for 3 minutes at 1.5g, with the supernatant discarded and the pellet resuspended in 1X PBS after each spin. The resuspended cells were then stained with 50 μg/mL wheat germ agglutinin (WGA) CF633 for 30 minutes. After staining, the cells were washed three times by centrifugation with PBS to remove excess WGA. During the final wash, the cells were resuspended in 0.1 mg/mL PI solution.

Prior to imaging, the samples were loaded into an ibidi µ-Slide VI 0.4. Imaging was performed using a ZEISS Elyra 7 Super-Resolution Microscope equipped with a 63× Apochromat alpha Plan-Apochromat Oil immersion DIC objective, NA 1.4, and pco.edge sCMOS cameras. Lattice structured illumination images were acquired and processed using the ZEN black software at the Life Sciences and Engineering (LS&E) Infrastructure Center, Technion-Israel Institute of Technology, Haifa, Israel. The experiments were conducted three times with similar observations.

Thioflavin T fluorescence fibrillation kinetics assay

Thioflavin T (ThT) (Sigma Aldrich) is a widely used fluorescent dye for detecting and analyzing amyloid fibril formation kinetics. Fibrillation curves in the presence of ThT typically exhibit an initial lag phase followed by rapid fibril elongation. To accurately capture the fibrillation lag time, PSMα3 was pre-treated before the experiment. The peptide was dissolved in a 1:1 mixture of 1,1,1,3,3,3-Hexafluoroisopropanol (HFIP-Sigma Aldrich) and Trifluoroacetic acid (TFA-Sigma Aldrich) to a final concentration of 1 mg/mL, followed by bath sonication for 10 minutes at room temperature. The organic solvents were then evaporated using a mini-rotational vacuum concentrator (Christ, Germany) at 1,000 rpm for 2 hours at room temperature.

For the experiment, freshly prepared 100 µM PSMα3 peptides, with or without 500 µM Epigallocatechin gallate (EGCG) at a 1:5 molar ratio, were prepared in 50 mM HEPES buffer pH 7.4 containing 150 mM NaCl. ThT was prepared by diluting a stock solution in ultrapure double-distilled water (UPddw) and filtered before use to reach a final concentration of 200 µM Blank control solutions containing all components except for the peptide were prepared for each reaction.

The assay was conducted in black 96-well flat-bottom plates (Greiner Bio-One), which were sealed with a thermal seal film (EXCEL Scientific) to prevent evaporation. Samples were incubated in a plate reader (OMEGA) at 37°C, shaking at 500 rpm for 85 seconds before each reading cycle, for up to 1,000 cycles of 6 minutes each, totalling approximately 100 hours.

Fluorescence was measured in triplicate using an excitation wavelength of 438 ± 20 nm and an emission wavelength of 490 ± 20 nm. All values were averaged, background fluorescence was subtracted using blank controls, and the results were plotted over time. Standard error of the mean is shown as error bars. The entire experiment was independently repeated at least three times on different days to ensure reproducibility.

Solid-State Circular Dichroism Spectroscopy

Solid-state circular dichroism spectroscopy(ss-CD) was conducted with PSMa3 to assess their fibrillar secondary structure components in presence and absence of Poly (AU) RNA. For preparation of peptide and RNA samples, see Peptide and Poly (AU) preparation section. PSMa3 was prepared at a working concentration of 100 µM in 50 mM HEPES, pH 7.4, 150 mM NaCl at 0 ng/µL, 50 ng/µL or 400 ng/µL Poly (AU) RNA. The reaction mixes were incubated for 0 min, 2 h and 72 h at 37 °C in a non-shaking thermocycler. Following incubation, soluble reaction components were removed via dilution washing with 150 µL water and centrifugation at 12.500 rcf for 30 min. 20 µL of the sediment were resuspended in 180 µL water for a second washing centrifugation step at 12.500 rcf for 30 min. 180 µL of the supernatant were discarded and 18 µL of sediment applied to a Chirascan Series fused silica disc (AP/CSSD, Applied Photophysics) in three times 6 µL steps. After every sample application onto to the silica disc, all liquid was evaporated at 37 °C for 5–10 min until only an opaque film remained visible. For secondary structure determination of aggregated/fibrillar peptide components on the silica disc, the disc was inserted into a Chirascan solid sample holder (CS/SSH, Applied Photophysics) and the circular dichroism recorded in the far-UV (180–250 nm, step size 1 nm, time-per-point 1 s) using a CD spectropolarimeter (Applied Photophysics). To counteract sample anisotropy, CD spectra were recorded at least four different disc rotations and the results averaged. Prior to data analysis, silica disc and buffer/water backgrounds were subtracted from recorded sample spectra. Poly (AU) RNA-only background spectra showed complex and high amplitude signals that did not appear for samples in combination with peptides and were therefore not subtracted from sample spectra.

NMR experiments

To prepare the reference samples for NMR analysis, 0.521 mg of PSMα3 was dissolved in 180 µL of 20 mM MES (2-(N-morpholino)ethanesulfonic acid) buffer at pH 6.0, supplemented with 5% (v/v) deuterium oxide (D₂O). From this solution, 170 µL was transferred into a 3.0 mm NMR tube for spectral acquisition. In parallel, a 4 mM stock solution of EGCG was prepared in the same 20 mM MES buffer (pH 6.0). From this, a working solution containing 0.5 mM EGCG in 20 mM MES with 5% (v/v) D₂O was prepared to match the buffer conditions of the peptide sample. A volume of 170 µL was then transferred into a separate 3.0 mm NMR tube for use as the EGCG reference.

All NMR experiments were recorded on a Bruker 700 MHz spectrometer equipped with a triple-resonance cryoprobe (Prodigy) and an AVANCE NEO console. One-dimensional (1D) ¹H-NMR spectra were acquired using the Bruker library pulse sequence zggpw5 for water suppression. The acquisition parameters were: acquisition time (AQ) = 1.90 s, spectral width (SW) = 12.31 ppm, recovery delay (D1) = 1.5 s, and 64 scans. All experiments were conducted at 35 °C, under which optimal spectral quality was obtained.

Two-dimensional (2D) ¹H–¹H TOCSY (Total Correlation Spectroscopy) spectra were collected using the Bruker pulse sequence dipsi2gpph19. Acquisition parameters were: t₂ = 2048 complex points (AQ₂ = 0.118 s), t₁ = 512 complex points (AQ₁ = 0.029 s), SW₁ = SW₂ = 12.31 ppm, with 64 scans and a mixing time (tmix) of 80 ms. Recovery delay (D1) was set to 1.5 s. 2D ¹H–¹H NOESY (Nuclear Overhauser Effect Spectroscopy) spectra were acquired using the noesyfpgpph19 pulse sequence, with acquisition parameters identical to those used for TOCSY. The mixing time was set to 300 ms. Total experimental time for the combined 2D TOCSY and NOESY experiments was approximately 30 hours. NMR data processing was carried out using TopSpin 4.5.0 (Bruker), and spectral assignments for the 2D TOCSY and NOESY spectra were performed using POKY software68. Signal integration was carried out using in-house scripts written in Scilab69.

eLife Assessment

This study presents valuable findings on the differential effects of RNA on the phase separation, aggregation dynamics, and bioactivity of PSMα3 and LL-37. The authors provide solid evidence from complementary biophysical and cell-based experiments that RNA influences peptide assembly and associated in vitro activities. The study is of interest for understanding interactions between amyloidogenic peptides and nucleic acids, although the physiological significance and some aspects of the mechanistic interpretation would benefit from further clarification.

Reviewer #1 (Public review):

Summary:

The manuscript by Rayan et al. aims to elucidate the role of RNA as a context-dependent modulator of liquid-liquid phase separation (LLPS), aggregation, and bioactivity of the amyloidogenic peptides PSMα3 and LL-37, motivated by their structural and functional similarities.

Strengths:

The authors combine extensive biophysical characterization with cell-based assays to investigate how RNA differentially regulates peptide aggregation states and associated cytotoxic and antimicrobial functions.

Weaknesses:

While the study addresses an interesting and timely question with potentially broad implications for host-pathogen interactions and amyloid biology, some aspects of the experimental design and data analysis require further clarification and strengthening.

Reviewer #2 (Public review):

In this paper, Rayan et al. report that RNA influences cytotoxic activity of the staphylococcal secreted peptide cytolysin PSMalpha3 versus human cells and E. coli by impacting its aggregation. The authors used sophisticated methods of structural analysis and describe the associated liquid-liquid phase separation. They also compare to the influence of RNA on aggregation and activity of LL-37, which shows differences to that on PSMalpha3.

That RNA impacts PSM cytotoxicity when co-incubated in vitro becomes clear. However, I have two major problems with this study:

(1) The premise, as stated in the introduction and elsewhere, that PSMalpha3 amyloids are biologically functional, is highly debatable and has never been conclusively substantiated. The property that matters most for the present study, cytotoxicity, is generally attributed to PSM monomers, not amyloids. The likely erroneous notion that PSM amyloids are the predominant cytotoxic form is derived from an earlier study by the authors that has described a specific amyloid structure of aggregated PSMalpha3. Other authors have later produced evidence that, quite unsurprisingly, indicated that aggregation into amyloids decreases, rather than increases, PSM cytotoxicity. Unfortunately, yet other groups have in the meantime published in-vitro studies on "functional amyloids" by PSMs without critically challenging the concept of PSM amyloid "functionality". Of note, the authors' own data in the present study that show strongly decreased cytotoxicity of PSMalpha3 after prolonged incubation are in agreement with monomer-associated cytotoxicity as they can be easily explained by the removal of biologically active monomers from the solution.

In their revision and in the rebuttal, the authors have further described their concept regarding what they call "functionality" of PSMalpha3 amyloids. They now admit that monomers are the active cytolytic form, like other researchers have stressed, whereas amyloids are not. This represents a considerable difference to earlier papers in which they ascribed functionality, i.e. cytolytic capacity, to PSMalpha3 amyloids, a claim that has raised considerable controversy. Now, they use the term "functional " to describe that PSMalpha3 amyloids, while not cytolytic, can be reversed to a cytolytic monomeric state, calling them a "dynamic reservoir". There is no evidence that such a reservoir is necessary for the cytolytic activity of the monomers to be established; also, there is no evidence that in a biological system, such an amyloid reservoir exists. To continue calling PSMalpha3 amyloids "functional" based on this - considerably changed - concept of the authors appears inappropriate, given the finally admitted absence of cytolytic activity of the PSM amyloids in addition to the continuing complete lack of evidence of any biological relevance of PSM amyloid formation.

(2) That RNA may interfere with PSM aggregation and influence activity is not very surprising, given that PSM attachment to nucleic acids - while not studied in as much detail as here - has been described. Importantly, it does not become clear whether this effect has biologically significant consequences beyond influencing, again not surprisingly, cytotoxicity in vitro. The authors do show in nice microscopic analyses that labeled PSMalpha3 attaches to nuclei when incubated with HeLa cells. However, given that the cells are killed rapidly by membrane perturbation by the applied PSM concentrations, it remains unclear and untested whether the attachment to nucleic acids in dying cells makes any contribution to PSM-induced cell death or has any other biological significance.

Overall, the findings can be explained in a much more straightforward way with the common concept of cytotoxicity being due to monomeric PSMs, and the impact of nucleic acids on cytotoxicity being due to lowering of the concentration of that active form by RNA attachment. Further limiting the significance of the findings, whether this interaction has any biological significance on the physiology or infectivity of the PSM producer remains largely unexplored.

Further remarks:

• Circumstantial evidence based on the "amyloid inhibitor", EGCG: The results with EGCG, which has been shown to have a moderate amyloid-reducing effect on PSMalpha 1 and PSMalpha4, should not be taken as evidence for amyloid-based cytotoxicity. While increased concentrations of EGCG reduced the cytotoxic effect of PSMalpha3, it is not convincingly shown that this is due to a lower concentration of amyloid vs. monomeric PSM.

• It is appreciated that the authors refrain from presenting the unsubstantiated concept of "functional" PSM amyloids in the discussion. However, wording in that direction must also be removed from other parts of the manuscript (e.g. "bioactive fibrillar polymorphs". "The formation of cross-alpha amyloids has been correlated with toxic activity", etc.), generally refraining from uncritically implying that amyloid formation underlies PSM biological activity, and rather discussing that the much more likely explanation of the findings is a lowering of cytolytically active, monomeric PSM concentration.

• Discussion: "PSM alpha3 interaction with nucleic acids within human cells ...supports a comparable mechanism...". Delete. Unsubstantiated.

• The authors should cite papers that have argued against their hypothesis and not only their own manuscripts.

Author response:

The following is the authors’ response to the previous reviews

eLife Assessment

This study presents valuable findings on the differential effects of RNA on the phase separation, aggregation dynamics, and bioactivity of PSMα3 and LL-37. The authors provide solid evidence from complementary biophysical and cell-based experiments that RNA influences peptide assembly and associated in vitro activities. The study is of interest for understanding interactions between amyloidogenic peptides and nucleic acids, although the physiological significance and some aspects of the mechanistic interpretation would benefit from further clarification.

We are grateful for the positive assessment. The two outstanding concerns about physiological significance and mechanistic interpretation are addressed in detail below through Reviewer #2's comments. We have made targeted revisions throughout the manuscript, and have been careful to distinguish genuine clarifications from reframing that would misrepresent what the data show.

Public Reviews:

Reviewer #1 (Public review):

Summary:

The manuscript by Rayan et al. aims to elucidate the role of RNA as a contextdependent modulator of liquid-liquid phase separation (LLPS), aggregation, and bioactivity of the amyloidogenic peptides PSMα3 and LL-37, motivated by their structural and functional similarities.

Strengths:

The authors combine extensive biophysical characterization with cell-based assays to investigate how RNA differentially regulates peptide aggregation states and associated cytotoxic and antimicrobial functions.

Weaknesses:

While the study addresses an interesting and timely question with potentially broad implications for host-pathogen interactions and amyloid biology, some aspects of the experimental design and data analysis require further clarification and strengthening.

We thank Reviewer #1 for the positive assessment. Previous revision round incorporated all major quantitative additions requested:

Quantitative EMSA binding analysis with Kd values and Hill coefficients (Fig. S1)

Quantitative FRAP recovery curves with mobile fractions and half-times (Figs. S4, S8, S12)

Colocalization metrics — Pearson's correlation coefficient and Manders' overlap coefficients (Fig. S5)

Quantification of AmyTracker630 amyloid signal intensity (Fig. S6)

Explicit acknowledgment of limitations regarding phase diagram boundaries and csat

Revised interpretation clarifying nucleolar localization as phenomenological, not causal

Reviewer #2 (Public review):

In this paper, Rayan et al. report that RNA influences cytotoxic activity of the staphylococcal secreted peptide cytolysin PSMalpha3 versus human cells and E. coli by impacting its aggregation. The authors used sophisticated methods of structural analysis and describe the associated liquidliquid phase separation. They also compare to the influence of RNA on aggregation and activity of LL-37, which shows differences to that on PSMalpha3.

That RNA impacts PSM cytotoxicity when co-incubated in vitro becomes clear. However, I have two major problems with this study:

The premise, as stated in the introduction and elsewhere, that PSMalpha3 amyloids are biologically functional, is highly debatable and has never been conclusively substantiated. The property that matters most for the present study, cytotoxicity, is generally attributed to PSM monomers, not amyloids. The likely erroneous notion that PSM amyloids are the predominant cytotoxic form is derived from an earlier study by the authors that has described a specific amyloid structure of aggregated PSMalpha3. Other authors have later produced evidence that, quite unsurprisingly, indicated that aggregation into amyloids decreases, rather than increases, PSM cytotoxicity. Unfortunately, yet other groups have in the meantime published in-vitro studies on "functional amyloids" by PSMs without critically challenging the concept of PSM amyloid "functionality". Of note, the authors' own data in the present study that show strongly decreased cytotoxicity of PSMalpha3 after prolonged incubation are in agreement with monomerassociated cytotoxicity as they can be easily explained by the removal of biologically active monomers from the solution.

In their revision and in the rebuttal, the authors have further described their concept regarding what they call "functionality" of PSMalpha3 amyloids. They now admit that monomers are the active cytolytic form, like other researchers have stressed, whereas amyloids are not. This represents a considerable difference to earlier papers in which they ascribed functionality, i.e. cytolytic capacity, to PSMalpha3 amyloids, a claim that has raised considerable controversy. Now, they use the term "functional " to describe that PSMalpha3 amyloids, while not cytolytic, can be reversed to a cytolytic monomeric state, calling them a "dynamic reservoir". There is no evidence that such a reservoir is necessary for the cytolytic activity of the monomers to be established; also, there is no evidence that in a biological system, such an amyloid reservoir exists. To continue calling PSMalpha3 amyloids "functional" based on this - considerably changed - concept of the authors appears inappropriate, given the finally admitted absence of cytolytic activity of the PSM amyloids in addition to the continuing complete lack of evidence of any biological relevance of PSM amyloid formation.

That RNA may interfere with PSM aggregation and influence activity is not very surprising, given that PSM attachment to nucleic acids - while not studied in as much detail as here - has been described. Importantly, it does not become clear whether this effect has biologically significant consequences beyond influencing, again not surprisingly, cytotoxicity in vitro. The authors do show in nice microscopic analyses that labeled PSMalpha3 attaches to nuclei when incubated with HeLa cells. However, given that the cells are killed rapidly by membrane perturbation by the applied PSM concentrations, it remains unclear and untested whether the attachment to nucleic acids in dying cells makes any contribution to PSM-induced cell death or has any other biological significance. Overall, the findings can be explained in a much more straightforward way with the common concept of cytotoxicity being due to monomeric PSMs, and the impact of nucleic acids on cytotoxicity being due to lowering of the concentration of that active form by RNA attachment. Further limiting the significance of the findings, whether this interaction has any biological significance on the physiology or infectivity of the PSM producer remains largely unexplored.

We thank the reviewer for the detailed comments. We appreciate the opportunity to further clarify our interpretation of the relationship between PSMα3 assembly, cytotoxicity, and RNA-mediated regulation. In the revised manuscript, and building on the previous revision round, we substantially expanded and refined the Discussion and Introduction to more clearly distinguish between mature fibrils, transient assembly intermediates, and broader assembly state-dependent mechanisms. We also incorporated additional literature representing different perspectives from the field. The revised manuscript presents a model in which biological activity is governed by dynamic assembly pathways and membrane-associated intermediates whose formation, persistence, and structural organization are modulated by environmental conditions, including RNA.

A central point raised by the reviewer is the suggestion that the RNA effects observed here can be explained simply by sequestration of active monomeric PSMα3. We respectfully disagree that this interpretation can account for the data. A monomer-depletion model makes a clear experimental prediction: conditions that promote aggregation should proportionally reduce activity by reducing the free monomer pool. However, our data show the opposite behavior. RNA promotes PSMα3 aggregation, induces liquid–liquid phase separation, and reshapes fibril morphology into distinct polymorphic assemblies, yet preserves cytotoxic and antimicrobial activity over incubation periods during which peptide alone progressively loses activity. Thus, activity does not correlate with suppression of aggregation or maintenance of soluble peptide. Instead, the data indicate that assembly trajectory and supramolecular organization are functionally relevant parameters. We state this point explicitly in the section “RNA preserves PSMα3 bioactivity,” where we added text clarifying that RNA does not prevent aggregation but redirects the assembly pathway toward structurally and functionally distinct states.

To further clarify our interpretation, we substantially revised the section “PSMα3 cytotoxicity arises from dynamic assembly intermediates.” This section now integrates multiple independent lines of evidence supporting an assembly-state-dependent model. Together, these observations argue against a simple binary model in which either monomers alone or mature fibrils alone determine activity. Instead, they support a framework in which transient intermediates formed along the assembly pathway contribute to membrane disruption and cytotoxicity. Consistent with this interpretation, our confocal and super-resolution microscopy experiments directly show PSMα3 accumulation and aggregation at bacterial and cellular membranes (Figs. 5, 6C, S10), supporting a model in which assembly occurs in direct association with membrane interfaces rather than exclusively in bulk solution prior to membrane contact. We expanded the Discussion accordingly.

We acknowledge the reviewer’s alternative interpretation that the nucleolar/nucleic-acid association observed in HeLa cells may reflect post-lysis binding following membrane permeabilization. We agree that this is a valid consideration at the cytotoxic concentrations used here, where membrane disruption is rapid (Figs. 5–6, Movies S1–S2). The Discussion therefore clarifies that nucleolar localization under these conditions is unlikely to represent a distinct intracellular toxic mechanism, but instead reflects the intrinsic nucleic-acid binding capacity of PSMα3 after cellular entry. We accordingly do not claim that intracellular nucleic-acid interactions contribute causally to cell death in these experiments. The potential biological relevance of PSMα3–nucleic acid interactions at sub-cytotoxic concentrations, where membrane disruption does not dominate, remains an important question for future investigation.

We additionally revised the manuscript to clarify the significance of the EGCG comparison. We agree with the reviewer that the EGCG data alone do not demonstrate “amyloid-mediated cytotoxicity,” and we do not make that claim. Rather, the comparison between EGCG and RNA provides evidence that different assembly trajectories produce different functional outcomes. EGCG redirects PSMα3 into amorphous, non-fibrillar assemblies that lose activity, whereas RNA promotes aggregation while preserving activity and generating distinct supramolecular morphologies. If activity depended solely on monomer concentration, both conditions would be expected to reduce activity similarly through sequestration. Instead, the divergent outcomes support the conclusion that assembly architecture and assembly pathway are functionally important.

In response to the reviewer’s concern that the manuscript overstates the concept of “functional amyloid,” we explicitly distinguish between mature fibrils and dynamic assembly processes, and we avoid wording that could be interpreted as implying that mature fibrils themselves are the active cytotoxic entities. At the same time, we note that the broader concept of functional amyloid-like assembly pathways is widely used in biology to describe assemblies whose formation regulates storage, localization, stabilization, or timing of bioactive states, including hormone-storage amyloids, RNA-binding protein assemblies, and bacterial curli systems. Within this framework, our interpretation is that PSMα3 assembly dynamics modulate the availability and lifetime of bioactive species rather than that mature fibrils themselves are directly toxic.

Importantly, we also broadened the manuscript substantially by incorporating independent studies from multiple unrelated systems supporting the principle that supramolecular organization influences biological function. These additions include: studies showing that structured fibrillar assemblies of LL-37 are required for specific antibacterial activities; work demonstrating that the nanoscale organization of β-defensin–nucleic acid complexes governs immunostimulatory potency; studies correlating α-helical solid-state conformations with cytotoxicity across fibril-forming antimicrobial peptides; salt-induced PSMα3 polymorphism studies showing distinct toxicities for amorphous versus fibrillar assemblies; and real-time AFM work demonstrating that membrane-associated protofibrillar intermediates are more disruptive than mature fibrils. We also added discussion of recent cryo-EM structures showing that RNA acts as a structural cofactor shaping tau fibril polymorphism at atomic resolution, as well as two-dimensional infrared spectroscopy studies demonstrating coexistence of cross-α and cross-β PSMα3 polymorphs. Together, these orthogonal observations from multiple systems support the broader principle that assembly architecture is a major determinant of biological behavior.

We also addressed the reviewer’s concern regarding biological relevance. We agree that direct in vivo validation remains an important future direction and state this explicitly in the revised Discussion. However, we respectfully submit that establishing the mechanistic principle that RNA regulates PSMα3 assembly state and functional output is itself a meaningful contribution independent of immediate in vivo confirmation. To better contextualize potential physiological relevance, we expanded the “Biological and therapeutic implications” section to discuss biologically plausible extracellular environments in which PSMα3 may encounter nucleic acids, including biofilms enriched in extracellular RNA, extracellular vesicles, damaged host tissues, inflammatory milieus, and host-derived extracellular RNA released as DAMPs.

Overall, the revised manuscript reflects a substantially expanded discussion of PSMα3 assemblystate-dependent activity, the role of RNA in modulating assembly trajectories, and the broader conceptual implications for membrane-active peptide assemblies.

Further remarks:

(1) Circumstantial evidence based on the "amyloid inhibitor", EGCG: The results with EGCG, which has been shown to have a moderate amyloid-reducing effect on PSMalpha 1 and PSMalpha4, should not be taken as evidence for amyloid-based cytotoxicity. While increased concentrations of EGCG reduced the cytotoxic effect of PSMalpha3, it is not convincingly shown that this is due to a lower concentration of amyloid vs. monomeric PSM.

We agree that the EGCG data alone should not be interpreted as evidence that mature amyloid fibrils are the directly cytotoxic species. Our interpretation is more limited and focuses on the effect of assembly redirection. Specifically, EGCG redirects PSMα3 into amorphous, non-fibrillar assemblies that lose activity, whereas RNA promotes aggregation while preserving activity and producing structurally distinct assemblies. The key conclusion is therefore that functional outcome depends on the nature and trajectory of assembly rather than on aggregation versus non-aggregation alone. We clarified this distinction in the revised Discussion section addressing RNA- versus EGCG-mediated modulation of PSMα3 assembly.

(2) It is appreciated that the authors refrain from presenting the unsubstantiated concept of "functional" PSM amyloids in the discussion. However, wording in that direction must also be removed from other parts of the manuscript (e.g. "bioactive fibrillar polymorphs". "The formation of cross-alpha amyloids has been correlated with toxic activity", etc.), generally refraining from uncritically implying that amyloid formation underlies PSM biological activity, and rather discussing that the much more likely explanation of the findings is a lowering of cytolytically active, monomeric PSM concentration.

In the Introduction, the phrasing 'may enable dynamic switching' has been used to soften the mechanistic claim regarding cross-α assemblies. The phrase 'bioactive fibrillar polymorphs' was revised in the previous round. At the same time, statements such as “cross-α amyloid formation has been correlated with toxic activity” are retained because they describe experimental observations reported in multiple studies (including Tayeb-Fligelman et al., 2017, 2020; Malishev 2018), without implying direct causality (correlation is not causation). We now explicitly frame these observations within a broader discussion of transient assembly intermediates and assembly-state-dependent toxicity.

(3) Discussion: "PSM alpha3 interaction with nucleic acids within human cells ...supports a comparable mechanism...". Delete. Unsubstantiated.

This sentence was removed in the previous revision round and remains absent from the current manuscript.

(4) The authors should cite papers that have argued against their hypothesis and not only their own manuscripts.

We appreciate this suggestion and agree that alternative interpretations should be represented explicitly. In the revised manuscript, we added and discussed studies including Zheng et al. (2018) and Yao et al. (2019) (already cited in both earlier versions), which support models in which advanced amyloid formation reduces cytotoxicity and active species are prefibrillar. These studies are now discussed substantively in both the Introduction and Discussion alongside our own work and that of others.

More broadly, we revised the manuscript to present the current understanding of PSMα3 toxicity as an actively debated question in the field rather than as a settled model. At the same time, we note that citing our prior studies remains necessary where the present work directly builds upon previously reported structural, biophysical, and mechanistic observations.

If the reviewer has additional specific references in mind, we welcome them and will incorporate them.

Version 3

Abstract

Amyloid-forming peptides are increasingly recognized as dynamic regulators at the host–pathogen interface, yet how environmental factors control their assembly and activity remains poorly understood. Here we show that RNA acts as a concentration-dependent regulator of two sequence-related α-helical peptides with fundamentally different assembly behaviors: the cross-α amyloid-forming Staphylococcus aureus virulence factor PSMα3 and the non-amyloidogenic human host-defense peptide LL-37. RNA drives PSMα3 through distinct assembly states, from liquid-like condensates to fibrillar polymorphs, while preserving cytotoxic and antimicrobial activity over time. In contrast, RNA attenuates LL-37 cytotoxicity toward host cells while maintaining antibacterial activity, consistent with a host-protective immunomodulatory effect. Together with the opposing effects of EGCG, which redirects both peptides into amorphous, inactive assemblies, these findings support a mechanistic model in which biological activity is governed not by monomer abundance or mature fibrils alone, but by supramolecular architecture, assembly trajectory, and dynamics. The contrasting RNA-dependent outcomes further suggest that the amyloid-forming capacity of PSMα3 enables RNA to sustain membrane-active states associated with bacterial virulence, whereas LL-37 is shifted toward less cytotoxic states that preserve antimicrobial defense while limiting host damage. More broadly, our findings identify RNA as an environmental regulator of α-helical peptide assemblies and establish assembly-state control as a tunable determinant of biological function.

Introduction

Bacterial amyloids are proteins capable of forming ordered fibrils, playing a crucial role in biofilm stability, surface adherence, cell toxicity and immune responses1–6. These amyloids are vital for the survival and virulence of many bacteria, enabling them to colonize and persist in diverse environments7,8. Some examples of bacterial amyloids include the Csg and Fap proteins from Enterobacteria and Pseudomonas species1,6,7,9–11, respectively, which play a role in biofilm stability. Staphylococcal phenol-soluble modulins (PSMs) are short peptides that predominantly adopt α-helical conformations in solution12 but are capable of forming amyloid fibrils13–21, and play multiple roles central to pathogenesis and immune evasion in Staphylococcus aureus22–35. In this context, defining PSMs as “functional amyloids” implies that amyloid formation can serve regulatory purposes, including controlled storage, stabilization, or spatial organization of bioactive molecules, as described for peptide hormones36 and RNA-binding proteins37,38. Accordingly, amyloid assembly may function as a reversible structural state that modulates the availability and timing of virulence-associated activities.

Among bacterial amyloids, PSMα3 is notable for its strong cytotoxicity toward a broad range of microbial and human cell types29,35,39–41 and for adopting a unique cross-α amyloid architecture, in which α-helices stack perpendicular to the fibril axis in paired sheets15–17. Within the PSMα family, distinct amyloid polymorphs appear to encode different biological functions32: the metastable and reversible cross-α assemblies of PSMα3 may enable dynamic switching between inactive and active states, whereas the more stable and irreversible cross-β fibrils formed by the homologous PSMα1 peptide20,42 may contribute to robust biofilm architecture. Two-dimensional infrared spectroscopy has further revealed that PSMα3 forms a heterogeneous ensemble of interconverting cross-α and cross-β assemblies, rather than a single defined structure43

The cytotoxic mechanism of PSMα3 remains debated, with studies attributing activity to soluble monomers44,45, or implicating the fibrillation process as required for full cytotoxic activity16,46. Notably, fibril-forming antimicrobial peptides show a correlation between α-helical conformations in the solid state and cytotoxicity47–51, paralleling human amyloids, where α-helical oligomeric intermediates are widely considered the primary toxic species52–54. Together, these observations support a model in which toxicity arises from transient assembly intermediates along the fibrillation pathway rather than from monomers or mature fibrils alone55.

PSMα3 shares self-assembly of α-helices with LL-37, a human host-defense peptide56–58, even though LL-37 is not an amyloid and does not form cross-α or cross-β fibrils. Moreover, the LL-37 antimicrobial active core (residues 17–29) shows sequence similarity to PSMα3 and forms non-amyloid fibrils composed of densely packed cationic amphipathic α-helices59. The sequence similarity and α-helices self-assembly behavior supported the hypothesis that PSMα3 might recapitulate functions of LL-37, and indeed, they share the activation of the signal inhibitory receptor on leukocytes-1 (SIRL-1), a negative regulator of myeloid cell function and dampens antimicrobial responses60. Such common functions might be related to molecular mimicry and the ability of bacterial peptides to hijack immunomodulation pathways61. Indeed, PSMαs interact with the formyl peptide receptor 2 (FPR2) on immune cells, initiating immune responses such as chemotaxis, phagocytosis, and the secretion of pro-inflammatory cytokines. These reactions are critical for recruiting and activating immune cells, thereby amplifying the inflammatory response24,62–67.

LL-37 has been shown to bind RNA released from dying cells, forming stable complexes that protect the RNA from enzymatic degradation and promote its uptake by dendritic cells68. Once internalized, these RNA–LL-37 complexes activate Toll-like receptors (TLRs), triggering cytokine production and dendritic cell maturation68,69. Beyond LL-37, other human AMPs, such as β-defensins, also form higher-order assemblies that organize nucleic acids into nanostructures with defined periodicity. These nanocomplexes potently stimulate TLRs, highlighting that innate immune activation is influenced not only by ligand identity, but also by the nanoscale geometry and spatial organization of the complexes70–72.

Interestingly, nucleic acid interactions also play a key role in the immune activation triggered by bacterial amyloids. For example, PSMs co-assemble with extracellular DNA (eDNA), and CpG-rich DNA specifically promotes PSMα3 fibrillation, forming co-aggregates that colocalize with DNA in biofilms. These complexes activate TLR pathways and induce anti-dsDNA autoantibody responses in mice28. Similar mechanisms are seen in E. coli and Salmonella, where curli fibers bind eDNA, enhance fibrillation, and engage TLR signaling, also promoting autoimmunity73,74. These findings suggest that amyloid–DNA complexes in bacterial biofilms are not only structural components but also potent immunomodulators with potential links to autoimmunity.

Given their shared cationic, amphipathic α-helical character, but distinct amyloidogenic properties, we sought to examine whether RNA differentially influences the assembly landscapes and bioactivity of PSMα3 and LL-37. RNA is increasingly recognized as a key regulator of protein phase behavior, capable of promoting liquid–liquid phase separation (LLPS), altering aggregation pathways, and reshaping supramolecular organization. LLPS is a fundamental mechanism of cellular organization, in which biomolecules demix into dynamic, membraneless compartments that regulate processes such as RNA metabolism, stress responses, and signal transduction75–78. Whether bacterial virulence-associated peptides are similarly governed by RNA-mediated phase behavior remains largely unexplored.

To address this, we investigated how defined RNA species modulate the structural transitions, material properties, and bioactivity of PSMα3 and LL-37. In parallel, we examined the impact of epigallocatechin gallate (EGCG), a small-molecule assembly modulator widely studied as an amyloid inhibitor33,79–81, to compare RNA-driven regulation with pharmacological redirection of peptide assembly. Here we show that RNA acts as a concentration-dependent modulator of PSMα3 and LL-37 assembly, inducing LLPS at low concentrations and distinct aggregated states at higher concentrations, with divergent functional consequences that reveal assembly architecture as a tunable determinant of peptide bioactivity.

Results

PSMα3 interacts with RNA in vitro

The interactions between PSMα3 and RNA were assessed using an Electrophoretic Mobility Shift Assay (EMSA), comparing the binding affinity of PSMα3 to single-stranded PolyA RNA and double-stranded Poly (AU) RNA. These RNAs were selected as simplified, well-defined model RNAs to probe general peptide–RNA interactions in an unbiased manner, as no prior information was available regarding whether such interactions occur, or which specific RNA species might be involved. The RNA molecules were labeled with the IR800CW fluorescent dye (as detailed in the Methods section). Freshly dissolved PSMα3, at varying concentrations, was incubated with each RNA type at a constant concentration for 30 minutes at 37°C. The EMSA results revealed that PSMα3 exhibits a stronger binding affinity to double-stranded Poly (AU) RNA, as evidenced by significant shifts at lower peptide concentrations compared to single-stranded Poly (A) (Fig. 1).

The binding and effect of single- and double-stranded RNA on PSMα3.

EMSA assay illustrating the interaction between PSMα3 and RNA. The assay compares the effects of increasing concentrations of PSMα3 (0 µM, 40 µM, 80 µM, 160 µM, and 320 µM), shown in a gradient from left to right, on the mobility shift of single-stranded Poly (A) RNA (left panel) and double-stranded Poly (AU) RNA (right panel) at ∼400ng/µL.

Quantification of free and shifted RNA band intensities revealed a progressive increase in the fraction of RNA bound as a function of peptide concentration for both RNA molecules (Fig. S1). Poly (AU) RNA reached saturation at lower PSMα3 concentrations than Poly (A), consistent with stronger apparent binding. Fitting the binding curves to a Hill model yielded an apparent dissociation constant (Kd) of 61.8 ± 1.6 µM for Poly (AU) and 94.5 ± 5.4 µM for Poly (A) (Fig. S1). Hill coefficients greater than 1 were obtained for both RNAs, consistent with cooperative binding behavior. Based on these Kd differences, subsequent experiments focused on PSMα3 behavior in the presence of double-stranded Poly (AU) RNA (hereafter referred to as RNA).

Non-monotonic turbidity response of PSMα3 to RNA

In order to assess the concentration-dependent effect of RNA on PSMα3 we used turbidity assays. The results showed that a 30 min co-incubation of 100 µM PSMα3 with varying concentrations of RNA displayed an increased turbidity up to 50 ng/µl, while higher RNA concentrations displayed a decrease in turbidity (Fig. S2). The decrease in turbidity above 50 ng/µL can be a result of either the formation of smaller particles or aggregation into larger clusters, which can settle out of suspension, removing them from the light-scattering medium and reducing turbidity. Alternatively, this non-monotonic turbidity behavior is conceptually similar to charge-driven re-entrant transitions described in other protein–RNA phase-separating systems, such as Ddx482, however, the molecular context and assembly architecture of PSMα3 are fundamentally distinct. According to the turbidity results, we hypothesized that 50 ng/µl can be a critical concentration between soluble species, which might be able to phase separate, and the formation of aggregates.

RNA concentration regulates formation and liquid-to-solid transition of PSMα3 condensates

To better understand the nature of the PSMα3-RNA interaction, we visualized their mixtures via fluorescence microscopy, using PSMα3 labeled with Fluorescein isothiocyanate (FITC) at the C-terminus (PSMα3-FITC), which is known to form fibrils15. All fluorescence microscopy experiments used a mixture of 20% FITC-labeled and 80% unlabeled peptide to minimize potential effects of the fluorophore on assembly behavior. 100 µM PSMα3-FITC was incubated with either 50 ng/µl or 400 ng/µl RNA, labeled using propidium iodide (PI), in a near-physiological buffer of 50 mM HEPES (pH 7.4) containing 150 mM sodium chloride (NaCl).

Upon addition of 50 ng/µL RNA, PSMα3 formed liquid–liquid phase-separated droplets that clearly colocalized with RNA (Fig. 2A). We further observed encapsulation of RNA within PSMα3–FITC–coated condensates (Fig. S3), indicating co-assembly rather than simple surface association. The liquid-like nature of these droplets was confirmed by fluorescence recovery after photobleaching (FRAP), which showed rapid signal recovery shortly after bleaching (Fig. 2B). Quantitative analysis revealed a half-time of recovery of 3.1 seconds and a mobile fraction of 68% (immobile fraction 32%) (Fig. S4), consistent with dynamic molecular exchange. Across multiple samples, we consistently observed coexistence of small droplets and larger aggregates. The experimental timescales examined do not allow us to reliably determine whether diffusion-driven coalescence kinetics would support classical droplet ripening dynamics83,84. Notably, after 2 hours of incubation, fluorescence recovery was no longer detected (Fig. 2C), indicating that the condensates undergo time-dependent aging and transition toward a more solid-like or aggregated state.

Colocalization, droplet formation, and texture of PSMα3 mixed with varying Poly (AU) RNA concentrations.

Widefield fluorescence microscopy images of 100 µM of 20% FITC-PSMα3 (green) and 80% unlabeled PSMα3 incubated with Poly (AU) RNA (red) at 50 ng/µL (A) or 400 ng/µL (D), shown as individual fluorescence channels and merged images. (B-C) FRAP analysis of PSMα3-FITC in the presence of 50 ng/µL Poly (AU) RNA after 10 min (B) or 2 h (C) of co-incubation. (E) FRAP analysis of PSMα3-FITC in the presence of 400 ng/µL Poly (AU) RNA after 10 min of co-incubation. For panels B, C, and E, images were acquired before photobleaching, immediately after photobleaching, and 40 s post-photobleaching. (F) 100 µM of 20% FITC-PSMα3 (green) and 80% unlabeled PSMα3 in the absence of RNA. All scale bars represent 20 µm.

In contrast to the effect of 50 ng/µl RNA, the addition of a higher RNA concentration of 400 ng/µl to PSMα3 resulted in immediate aggregation, with no observed phase of droplet formation (Fig. 2D). FRAP analysis indicated no recovery of signal after bleaching, likely due to the formation of solid structures rather than droplets with liquid-like properties (Fig. 2E). In the absence of RNA, PSMα3 did not form condensates and instead appeared as dispersed or irregular assemblies (Fig. 2F), establishing that RNA is required to induce the droplet state observed under these conditions.

Colocalization between PSMα3-FITC and RNA was quantified using Pearson’s correlation coefficient (above the Costes threshold) and Manders’ overlap coefficients (tM1 and tM2) (Fig. S5). All metrics indicate substantially stronger spatial correlation and signal overlap at 50 ng/µL RNA compared with 400 ng/µL RNA. At low RNA concentration, Pearson’s correlation is positive (R ≈ 0.53), accompanied by high Manders’ overlap values (tM1 ≈ 0.68; tM2 ≈ 0.63). This is consistent with coordinated association of PSMα3 and RNA within liquid-like droplets. In contrast, at 400 ng/µL RNA, Pearson’s correlation becomes negative (R ≈ −0.24) and Manders’ coefficients are markedly reduced. This indicates a transition from co-localized droplet assemblies to spatially segregated aggregated structures in which RNA becomes partially excluded from the peptide-rich structures.

RNA concentration-driven changes in PSMα3 fibrillar morphology and aggregation

The TEM micrographs showed that in the absence of RNA, 100 µM PSMα3 formed nanotube-like fibrils after 2 hours, growing wider or with some twist after 24-hours co-incubation (Fig. 3A). Incubating PSMα3 with a low concentration of 10 ng/µL RNA displayed accelerated fibril formation, with more strongly twisted, wide, sheet-like fibrils observed after both 2 hours and 24 hours (Fig. 3A). With the addition of 50 ng/µL RNA, PSMα3 fibrils have a similar twisted morphology after 2 hours, while at a longer incubation time of 24 hours, we observed a significant morphological shift into more thin amorphous aggregates (Fig. 3A). At the higher concentration of 400 ng/µL RNA, PSMα3 formed dense, thin fibrils after 2 hours, but with a possible fragmentation into smaller species or rearrangement into amorphous aggregates after 24 hours of co-incubation. This indicates that RNA concentration and time of co-incubation affect the density and morphology of PSMα3 fibrils. This corresponds to the differences observed by light microscopy of co-aggregates morphology of LLPS droplets vs solid aggregates (Fig. 2).

TEM and TIRF visualization of PSMα3 aggregation and morphology with different Poly (AU) RNA concentrations and incubation times.

(A) TEM micrographs of 100 µM PSMα3 incubated with or without Poly (AU) RNA at varying concentrations of 10 ng/µL, 50 ng/µL, and 400 ng/µL for 2 hours (top row) and 24 hours (bottom row). Scale bars represent 500 nm. (B) TIRF microscopy images showing 100 µM of 20% FITC-PSMα3 (green) and 80% unlabeled PSMα3 co-incubated with 50 ng/µL Poly (AU) RNA and the amyloid indicator AT630 (magenta) for 30 minutes and 2 hours. Scale bars represent 20 μm. (C) TIRF microscopy images of 100 µM of 20% FITC-PSMα3 (green) and 80% unlabeled PSMα3 co-incubated with 400 ng/µL Poly (AU) RNA and AT630 (magenta) for 30 min. Scale bars represent 20 µm.

These observations were further supported by TIRF microscopy using the amyloid-specific dye AmyTracker630 (AT630). With the addition of 50 ng/µL RNA to PSMα3-FITC, a strong AT630 fluorescence signal was detected only after 2 hours of co-incubation, but not after 30 minutes (Fig. 3B, quantified in Fig. S6). This suggests a time-dependent transition into amyloid-like species. Conversely, at 400 ng/µL RNA, a significant AT630 fluorescence was observed already after 30 minutes of co-incubation, consistent with a rapid formation of amyloid fibrils (Figs. 3C and S6). These findings highlight a concentration- and time-dependent modulation of PSMα3 phase separation and structural transitions, where RNA promotes LLPS at lower concentrations and drives rapid amyloid formation and unique fibrillar morphologies at higher concentrations. Future work will be required to quantitatively define the phase boundaries and delineate the dominant mechanisms, such as sedimentation, dissolution, or coarsening/aging, across intermediate RNA concentrations.

RNA enhances α-helical structure of PSMα3

Solid-state circular dichroism (ssCD) spectroscopy reveals that PSMα3 lacks a defined secondary structure both immediately after preparation and following 2 hours of incubation (Fig. S7). However, an α-helical signature of PSMα3 is markedly enhanced in the presence of RNA compared to peptide alone, as evidenced by increased signal intensity, deeper minima, and more pronounced spectral features characteristic of α-helical structure. This enhancement is more pronounced at 400 ng/µL RNA than at 50 ng/µL, particularly after 2 hours of co-incubation, indicating that RNA concentration influences the stabilization of α-helical assemblies. This supports the notion that RNA not only accelerates aggregation but also promotes or stabilizes the α-helical fibrillar architecture, potentially consistent with cross-α amyloid structures. This RNA-stabilized α-helical fibrillar architecture provides a structural basis for understanding why RNA preserves PSMα3 bioactivity over extended incubation times, as shown below.

RNA modulates the antibacterial activity of PSMα3

The antimicrobial activity of PSMα3 against Escherichia coli was evaluated under varying RNA concentrations and incubation times using the PrestoBlue Cell Viability Assay (Fig. 4A). Freshly dissolved 10 or 20 µM PSMα3 exhibited potent antibacterial activity, completely abolishing bacterial viability. After 2 hours of incubation, PSMα3 retained its full activity, comparable to its freshly prepared state. The presence of RNA at 50 ng/µL and 400 ng/µL had no impact on its antimicrobial function within this timeframe (Fig. 4A).

Impact of Poly (AU) RNA on PSMα3 cytotoxicity and antibacterial activity.

Antimicrobial activity of PSMα3 against E. Coli using the PrestoBlue cell viability assay (A) and its cytotoxicity against HeLa cells using the LDH colorimetric assay (B) were assessed with and without Poly (AU) RNA at varying concentrations. The experiments were performed in at least three replicates and repeated across three independent days to ensure result reliability. Cytotoxicity and bacterial cell viability percentages were calculated as the mean of all replicates, with error bars representing the standard error of the mean (SEM). Statistical significance was determined using one-way ANOVA for normally distributed data in GraphPad Prism (version 11). Significance levels are indicated as follows: *p<0.05, **p<0.01, ***p<0.001, ****p<0.0001.

In contrast, after 24 hours of incubation, PSMα3’s antibacterial activity was significantly reduced (Fig. 4A), suggesting a decrease in its effective concentration or changes in its morphology. Notably, the addition of RNA at 50 ng/µL or 400 ng/µL prevented this loss of activity, indicating that RNA plays a stabilizing role in toxic species or their reservoir. These findings suggest that RNA influences PSMα3 aggregation dynamics and morphology, thereby modulating its long-term antimicrobial effectiveness.

RNA modulates PSMα3 cytotoxicity against human HeLa cells

The effect of RNA on the cytotoxicity of PSMα3 against human HeLa cells was evaluated by measuring lactate dehydrogenase (LDH) release, an indicator of cell membrane damage (Fig. 4B). Freshly dissolved PSMα3 exhibited substantial cytotoxicity, causing approximately 80% cell death at 20 µM and 50% at 10 µM. The presence of RNA did not significantly alter the toxicity of freshly dissolved PSMα3 (Fig. 4B), similar to the antibacterial activity. However, following 2 hours of incubation, PSMα3 cytotoxicity was significantly reduced, likely due to aggregation-associated loss of membrane-active species. This reduction occurred earlier than the loss of antibacterial activity, which was only observed after 24 hours of incubation (Fig. 4A).

Notably, co-incubation with RNA helped maintain PSMα3 cytotoxicity in a concentration-dependent manner. With 50 ng/µL RNA, the 2-hour incubated PSMα3 retained cytotoxicity comparable to its freshly dissolved form, suggesting that RNA prevents activity loss due to incubation. However, after 24 hours, 50 ng/µL RNA was insufficient to preserve cytotoxicity. In contrast, at 400 ng/µL RNA, partial cytotoxicity was maintained for both 2-hour and 24-hour incubated PSMα3, compared to the freshly dissolved sample.

Overall, these findings suggest that RNA prevents the incubation-induced loss of both antibacterial activity and cytotoxicity (Fig. 4). This effect appears to be concentration- and co-incubation time-dependent, likely linked to RNA-induced morphological variations of PSMα3 species (Figs. 2-3).

PSMα3 accumulates in nucleolar RNA-rich compartments of HeLa cells

The interactions and localization of PSMα3 were next explored in HeLa cells with the addition of 20µM PSMα3 just before imaging. PI was added to the cell medium as a marker for cell death, as it selectively penetrates dead cells and binds to their nucleic acids.

Confocal microscopy images revealed that PSMα3 induces significant toxicity as indicated by membrane damage and blebbing, intracellular PSMα3 aggregate formation, and positive PI staining (Fig. 5 and Movie. S1). PSMα3 also enters the nuclei, with a notable concentration in the nucleolus, indicated by the foci of green fluorescence (marked by white arrows in Fig. 5). The composite images demonstrated clear colocalization of PSMα3 with nucleic acids in the nucleolus, as seen by the overlap of the green PSMα3-FITC signal and the red fluorescence from PI-stained nucleic acids.

Colocalization of PSMα3 with nucleic acids in HeLa cells.

(A) Confocal microscopy images showing the localization and colocalization of 20 µM of 20% FITC-PSMα3 (green) and 80% unlabeled PSMα3 and nucleic acids stained with PI (red) inside the nucleolus of HeLa cells (indicated in arrows). The left panel illustrates the distribution of PSMα3 within the cell. The middle panel shows the nucleic acids stained with PI. The right panel is a composite image that demonstrates the colocalization of PSMα3 with nucleic acids. Scale bars represent 15 µm. (B) FRAP analysis of 20 µM of 20% FITC-PSMα3 (green) and 80% unlabeled PSMα3 inside the nucleolus (indicated by the arrow) of HeLa cells, showing fluorescence recovery after 60 seconds. Scale bars represent 10 µm.

FRAP analysis of 20 µM PSMα3 within the nucleolus of HeLa cells revealed measurable but constrained mobility (Fig. 5B). Following photobleaching of a defined nucleolar region (indicated by the white arrow), fluorescence gradually recovered over the 60 s acquisition window, indicating dynamic exchange of PSMα3 within this compartment. However, recovery was slow and incomplete during this time frame, and did not reach a clear plateau, precluding reliable determination of recovery half-time or precise mobile and immobile fractions (Fig. S8), and indicating dynamic but constrained exchange within this RNA-rich compartment. This behavior differs from the rapid and near-complete recovery observed for PSMα3–RNA droplets formed in vitro (Fig. 2C), indicating that nucleolar-associated PSMα3 exhibits more restricted dynamics than liquid-like condensates in defined buffer systems.

EGCG directly binds PSMα3 and inhibits its fibrillation and bioactivity, even in the presence of RNA

Since RNA appears to protect or maintain PSMα3’s toxic functions, potentially by influencing its fibril formation and morphology, we investigated the corresponding and combined effects of an inhibitor of fibril formation. One such inhibitor is epigallocatechin gallate (EGCG), the most abundant catechin in tea and a known amyloid inhibitor, including PSMs33.

The addition of EGCG to PSMα3 at a 1:1 molar ratio did not significantly alter its cytotoxicity against HeLa cells. However, at a fivefold molar excess, EGCG completely abolished PSMα3’s cytotoxic effect (Fig. 6A). To examine the structural changes underlying this effect, we analyzed fibril formation kinetics and morphology in the presence of EGCG. TEM micrographs revealed that a fivefold molar excess of EGCG disrupted fibril formation of 100 µM PSMα3, instead inducing amorphous aggregates (Fig. 6B). Consistently, kinetics assays of fibril formation showed that EGCG inhibited the ThT fluorescence curve otherwise indicating the fibril formation of 100 µM PSMα3 (Fig. S9). These findings suggest that EGCG reduces PSMα3 toxicity by modulating its fibril formation and morphology. The addition of RNA did not counteract the effect of EGCG or restore PSMα3 cytotoxicity (Fig. 6A), highlighting the pronounced effect of EGCG on PSMα3 morphology and properties.

EGCG modulates PSMα3 aggregation and reduces toxicity against HeLa cells.

(A) Cytotoxicity of PSMα3 against HeLa cells in the presence and absence of EGCG and Poly (AU) RNA at two different concentrations, assessed via LDH assay. The experiment was performed in triplicate and repeated on three separate days for consistency. Cytotoxicity percentages were averaged across all replicates, with error bars representing the standard error of the mean. Statistical significance: *p < 0.05, **p < 0.01, ***p < 0.001, ****p < 0.0001 (one-way ANOVA, GraphPad Prism v11). (B) TEM micrographs of 100 μM PSMα3 incubated for 24 hours, without (left) and with (right) a fivefold molar excess of EGCG. Scale bars: 500 nm. (C) Live-cell confocal microscopy of HeLa cells treated with 20 μM of 20% FITC-PSMα3 (green) and 80% unlabeled PSMα3 without (top) and with EGCG (bottom), imaged immediately after preparation (t = 0) and after 10 minutes. Hoechst 33342 (blue) marks the nuclei, while PI staining (red) indicates membrane disruption. Scale bars: 10 μm.

Live-cell confocal microscopy provided further insights into how EGCG binding affects PSMα3 interactions with cells. In accordance with the LDH measurements (Fig. 6A), in the absence of EGCG, 20 µM PSMα3 readily penetrated cells and induced robust PI staining, indicating membrane disruption (Fig. 6C, top panel and Movie. S1). In contrast, a fivefold molar excess of EGCG prevented the toxic effects of 20 µM PSMα3, blocking cell penetration and PI staining (Fig. 6C, bottom panel and Movie. S2). Instead, PSMα3-FITC formed extracellular aggregates and showed no interaction with cell membranes, as indicated by the absence of PI staining and intact cell membranes.

Furthermore, EGCG at a fivefold molar excess also reduced the antibacterial activity of PSMα3 against E. coli, preserving bacterial cell viability (Fig. S10). Super-resolution fluorescence microscopy showed that while 20 µM PSMα3 typically aggregates on the bacterial membrane, causing membrane disruption and PI staining indicative of cell death (Fig. S10), EGCG at a fivefold molar excess prevented membrane aggregation and PI staining, thereby preserving membrane integrity.

To investigate residue-specific interactions between PSMα3 and EGCG, we performed NMR spectroscopy using a 2:1 PSMα3:EGCG molar ratio. The 1D 1H-NMR spectrum revealed distinct chemical shift changes and peak broadening upon EGCG addition, particularly in residues Met1, Glu2, and V4/N21, suggesting specific interactions between these sites and the EGCG molecule (Fig. 7A). Peak broadening was also observed in the aromatic region of EGCG, especially for protons H1 and H2, compared to the reference spectrum of EGCG alone, indicating that the interaction is visible from the EGCG side as well. Notably, slight opalescence was observed in the sample following preparation, potentially reflecting early-stage aggregate formation. Additionally, the presence of multiple cross-peaks between non-sequential residues (i+3 or i+4) in the two-dimensional (2D) 1H-1H Nuclear Overhauser Effect Spectroscopy (NOESY) spectrum indicates a well-defined structure of PSMα3 in this condition (Fig. 7B). The 2D 1H-1H Total Correlation Spectroscopy (TOCSY) spectrum displayed connectivity between backbone amide (HN) and alpha protons (Hα) for several assigned residues, while the 2D 1H-1H NOESY spectrum revealed spatial correlations between nearby residues (i+1).

Residue-specific interactions between PSMα3 and EGCG.

(A) One-dimensional (1D) 1H NMR spectra of 1.0 mM PSMα3 alone (red) and in complex with 0.5 mM EGCG (blue), recorded at 35 °C. Specific residues, including Met1, Glu2 and V4/N21, show chemical shift changes suggestive of direct interaction with EGCG (highlighted in the upper left). Dashed boxes mark proton signals corresponding to EGCG. Peak broadening of H1/H2 protons, compared to the EGCG-only reference sample (light grey), indicates interaction from the EGCG side. Slight opalescence observed in the sample suggests potential aggregate formation. (B) Two-dimensional (2D) 1H–1H TOCSY and NOESY spectra of the PSMα3:EGCG complex at a 2:1 ratio, recorded at 35 °C. The cross-peaks in the TOCSY spectrum allow the identification of the spin system and direct connections through the scalar coupling between the proton amide (HN) and the alpha protons (Hα) of the same residue. The cross-peaks in the NOESY spectrum establish a sequential connection between neighboring residues (i+1), allowing spectral assignment. The remaining cross-peaks (i+3 or i+4) in the spectrum support PSMα3 secondary structure in the experimental conditions. (C) Temporal stability of the PSMα3:EGCG sample over 3 days.

To assess the temporal stability of the sample, we monitored signal intensities over a 3-day period. PSMα3 signals remained stable, with only a ∼10% decrease in intensity, whereas EGCG signals exhibited substantial degradation, with nearly 50% loss over the same time frame (Fig. 7C). Given the ∼1.5-day duration of the 2D NMR measurements, it is estimated that over 60% of EGCG remained in solution during data acquisition, allowing for reliable observation of its interaction with the peptide. Of note, EGCG’s activity has been shown to depend on its chemical stability and the surrounding conditions. Specifically, at neutral pH, EGCG may undergo oxidation, and its inhibitory effects could be attributed to its degradation products rather than the intact compound itself85.

RNA modulates the human host-defense peptide LL-37 phase behavior and cytotoxicity without compromising antimicrobial activity

LL-37, a human host defense peptide, shares similarities with PSMα3, including its ability to self-assemble into α-helical supramolecular structures and to interact with nucleic acids59,68. We therefore examined whether RNA similarly influences LL-37 activity and properties.

Fluorescence microscopy showed that 100 µM LL-37-FITC incubated with 100 ng/µL PI-labeled Poly (AU) RNA formed aggregates, with no evidence of droplet formation or LLPS under these conditions (Fig. S11). Because diverse cellular stresses are known to induce condensation of RNA-binding proteins and promote the formation of stress-associated assemblies86, we next examined the effect of RNA on LL-37 under thermal stress. Following heat shock at 65 °C for 15 minutes, RNA markedly altered LL-37 assembly behavior in a concentration-dependent manner. At 100 ng/µL RNA, LL-37 formed rounded condensates that colocalized with RNA (Fig. 8A). To assess the material properties of these assemblies, we performed FRAP analysis on LL-37–FITC under this condition. Photobleaching resulted in minimal fluorescence recovery over the acquisition window (Fig. S12), indicating that the LL-37–RNA assemblies formed after heat shock are predominantly solid-like rather than liquid-like. At 200 ng/µL RNA, LL-37 assemblies exhibited mixed features, with rounded structures still present but accompanied by increasing aggregation, consistent with a transition toward a more solid state (Fig. 8B). At 400 ng/µL RNA, discrete condensates were no longer observed and LL-37 instead formed extensive, irregular aggregates with strong RNA colocalization (Fig. 8C).

Effect of RNA concentration on LL-37 phase separation, aggregation, and activity at pH 7 after heat shock.

Fluorescence microscopy images showing 20% FITC-labeled (green) and 80% unlabeled 100 μM LL-37 in the presence of increasing concentrations: 100 ng/μL (A), 200 ng/μL (B) and 400 ng/μL (C) of Poly (AU) RNA (PI-labeled, red) after heat shock at 65°C for 15 minutes at pH 7. Scale bars represent 20 µm. (D) LL-37 cytotoxicity, with and without Poly (AU) RNA and EGCG at varying concentrations, was assessed in HeLa cells using the LDH colorimetric assay. The experiment was performed in triplicate and repeated on three separate days for reproducibility. Cytotoxicity percentages represent the average of all replicates, with error bars indicating the standard error of the mean. (E) The antimicrobial activity of LL-37, with and without Poly (AU) RNA and EGCG at varying concentrations, was evaluated against E. coli using the PrestoBlue Cell Viability assay. The experiment was performed in triplicate and repeated on three separate days for reproducibility. Bacterial viability percentages represent the average of all replicates, with error bars indicating the standard error of the mean. Statistical significance (D-E): *p < 0.05, **p < 0.01, ***p < 0.001, ****p < 0.0001 (one-way ANOVA, GraphPad Prism v11).

For comparison, PSMα3 subjected to the same heat-shock protocol exhibited droplet formation only under more permissive conditions (pH 4), but not at physiological pH (7.4) (Figs. S13-S14). Together, these results indicate that, for both LL-37 and PSMα3, increasing RNA concentration favors aggregation over phase separation, and that LL-37 assemblies formed under thermal stress rapidly adopt solid-like properties rather than maintaining liquid-like dynamics.

Freshly dissolved LL-37 exhibited cytotoxicity against HeLa cells, which, in contrast to PSMα3, remained consistent even after incubation for up to 24 hours. Notably, RNA significantly reduced LL-37 cytotoxicity in a concentration-dependent manner, irrespective of the incubation duration (freshly dissolved, 2 hours, or 24 hours) (Fig. 8D). This contrasts with RNA’s rescuing effect on the incubation time-dependent loss of PSMα3 cytotoxicity. The antimicrobial activity of LL-37 against E. coli also remained unchanged upon incubation. However, unlike its effect on cytotoxicity, RNA had minimal impact on antibacterial activity, with only high concentrations of 400 ng/µL reducing LL-37 activity (Fig. 8E). Similar to PSMα3, LL-37 cytotoxicity against HeLa cells and its antimicrobial activity were attenuated by EGCG in a concentration-dependent manner (Fig. 8E).

TEM micrographs of 100 µM LL-37 revealed a variety of aggregative and fibrous structures, including thin fibrils, although no consistent morphology was observed (Fig. S15A). RNA induced a distinct morphological change, leading to increased aggregation and the formation of mostly amorphous species, but also some thicker, ribbon-like fibrils were observed. These structures resemble those previously observed for a segment of the LL-37 active core (residues 17–29), which is similar to PSMα3 in sequence and its ability to form fibrils of densely packed amphipathic α-helices59. Heat shock further intensified aggregation, resulting in denser, amorphous condensates and larger fibrillar assemblies, with and without RNA (Fig. S15B). The distinct effects of RNA on the aggregation and morphologies of LL-37 and PSMα3 may underlie the observed differences in their cytotoxic and antibacterial activities.

Discussion

This study identifies RNA as a context-dependent regulator of α-helical assembly-prone peptides, demonstrating that RNA reshapes their supramolecular landscape and thereby tunes bioactivity rather than merely sequestering peptide species. This is supported by the emergence of distinct RNA-dependent material states (from droplets to aged solids) and RNA-dependent fibrillar morphologies with divergent activity profiles across time and concentration. By comparing the bacterial virulence factor PSMα3, which forms canonical cross-α amyloid assemblies, with the human host-defense peptide LL-37, which forms α-helical fibrils but lacks amyloid-like stacking, we reveal that RNA differentially modulates structurally distinct systems through peptide-specific effects on phase behavior and aggregation dynamics.

RNA preserves PSMα3 bioactivity by redirecting its assembly pathway

Previous studies have argued that aggregation into amyloids primarily decreases PSM cytotoxicity by depleting the active soluble pool44,45, a view consistent with our observation of time-dependent activity loss upon incubation32 (Fig. 4), coinciding with the formation of dense, fibrillar assemblies (Figs. 2-3). These mature aggregates likely represent inert end states that sequester membrane-active intermediates44,45. Our data, however, indicate that RNA-dependent modulation of aggregation trajectory can preserve activity independently of simply reducing aggregation, which challenges a purely monomer-centric interpretation. Specifically, RNA reshapes this trajectory in a concentration-dependent manner. At low RNA concentrations, PSMα3 initially retains activity (Fig. 4B) but gradually loses it as liquid-like RNA–PSMα3 condensates mature into more rigid assemblies (Figs. 2-4 & S3), indicating that RNA regulates the lifetime of dynamic intermediates. At higher RNA concentrations, RNA promotes alternative fibrillar polymorphs that retain substantial cytotoxic activity even after prolonged incubation (Fig. 4B). Similarly, RNA preserves antibacterial potency over time (Fig. 4A), consistent with stabilization of assembly states that remain competent for membrane interaction. Thus, RNA does not simply suppress aggregation but potentially modulates the reversibility and supramolecular architecture of α-helical assemblies (Fig. S7), thereby controlling the duration and accessibility of the toxic window.

The temporal differences between cytotoxicity (Fig. 4B) and antibacterial activity (Fig. 4A) may reflect distinct membrane environments, aggregation kinetics, or interactions with bacterial surfaces versus mammalian membranes. Nonetheless, in both contexts, activity correlates with assembly state rather than peptide abundance alone. RNA therefore acts as a structural modulator that controls the lifetime and reversibility of functionally relevant intermediates.

Recent cryo-EM studies show that RNA acts as a structural cofactor in tau fibrils, directing polymorphic folds, stabilizing assemblies via electrostatic interactions, and forming intermediate states linked to pathology; RNase-induced disassembly confirms RNA as an integral architectural component rather than a passive trigger87. This parallels our observations for PSMα3, where RNA similarly governs supramolecular architecture and functional outcomes.

Regulatory modulation versus irreversible inhibition

The contrasting effects of RNA and the amyloid inhibitor EGCG highlight that bioactivity is governed by supramolecular architecture and reversibility rather than aggregation per se. RNA preserves PSMα3 bioactivity by stabilizing dynamic, α-helical assembly states. In contrast, EGCG binds soluble PSMα3 and diverts it into stable amorphous aggregates that represent structural dead ends, abolishing activity (Figs. 6-7). Thus, EGCG should not be viewed as evidence that amyloid fibrils are the toxic species; instead, it abolishes activity by removing or redirecting soluble bioactive species into non-productive amorphous assemblies. If cytotoxic activity were determined solely by free monomer concentration, both RNA and EGCG would be expected to reduce activity through sequestration. Instead, RNA promotes aggregation while preserving activity, whereas EGCG, which also reduces free soluble peptide, abolishes it, demonstrating that the assembly pathway and resulting structural architecture, rather than soluble peptide abundance alone, are determinants of functional outcome. Overall, RNA functions as a regulatory modulator of assembly dynamics, whereas EGCG enforces irreversible pathway diversion. These results underscore that assembly state, not simply peptide presence, determines biological function.

A parallel diversion mechanism is observed with human fibrinogen, an abundant host plasma protein encountered by PSMs at infection and wound sites: it suppresses fibrillation of most PSMs or redirects aggregation into off-pathway species, while uniquely accelerating non-fibrillar, amorphous aggregation of PSMα388. This further illustrates that host environmental factors encountered during infection can reshape PSMα3 assembly trajectories in ways that parallel the RNA- and EGCG-mediated redirections described here.

PSMα3 cytotoxicity arises from dynamic assembly intermediates

Current evidence supports a model in which PSMα3 cytotoxicity does not arise from a single static species, i.e., neither free monomers nor mature fibrils, but from dynamic assembly processes involving transient intermediates along the fibrillation pathway. PSMα3 assembly is highly sensitive to environmental conditions that shape both its aggregation trajectory and functional output. Lipid bilayers modulate aggregation kinetics in a concentration- and composition-dependent manner without redirecting the dominant fibrillar pathway89, while hydrodynamic shear comparable to vascular flow accelerates amyloid formation and promotes α-helical fibrils90. Salt-induced polymorphism further generates structurally distinct amorphous, fibrillar, and oligomeric assemblies from the same peptide sequence, with fibrillar and oligomeric species exhibiting substantially higher cytotoxicity than amorphous aggregates91, demonstrating that supramolecular architecture, rather than peptide identity or monomer availability, governs activity.

Multiple independent observations converge on membrane-associated intermediates as the primary cytotoxic species. Mutants that retain α-helical monomeric structure yet fail to fibrillate lose toxicity15,16, while mature fibrils are largely inactive44,45, indicating that neither endpoint state alone accounts for activity. Real-time AFM studies of N-formylated PSMα3, a native modification, further show that protofibrillar intermediates forming at membrane interfaces drive membrane disruption, whereas mature fibrils are less active92. Consistently, we observe peptide accumulation and aggregation at cellular and bacterial membranes (Figs. 6C and S10), supporting the view that bioactive intermediates are structurally distinct from both soluble monomers and mature fibrils. Together, these findings position the toxic window within transient, membrane-associated assembly states55.

Molecular dynamics simulations further suggest that PSMα3 monomers undergo membrane-induced folding into α-helical conformations upon contact93, implicating the membrane interface as a site of both structural organization and initiation of cytotoxic assemblies. In this framework, membrane-associated fibril growth may destabilize bilayers through lipid recruitment and extraction46, analogous to mechanisms proposed for membrane-active human amyloids such as islet amyloid polypeptide94. Together, these findings support a model in which cytotoxicity emerges from dynamic, environmentally regulated assembly intermediates rather than from free monomers or mature fibrils alone.

The RNA-dependent modulation described here represents a specific instance of this broader principle: environmental factors continuously reshape PSMα3 assembly pathways and thereby control the availability, lifetime, and structure of bioactive species. Crucially, RNA does not preserve cytotoxicity by preventing aggregation; it promotes aggregation while altering its trajectory and morphology. RNA-treated samples retain cytotoxicity over extended incubation times, whereas peptide alone loses activity as aggregation proceeds, inconsistent with a model based solely on monomer sequestration. Instead, these findings support a framework in which assembly state, fibril morphology, and aggregation reversibility, rather than free monomer concentration alone, govern cytotoxic function.

LL-37: non-amyloid α-helical assemblies under RNA control

In contrast to PSMα3, LL-37 does not form canonical amyloid architectures. This structural distinction of amyloidogenic PSMα3 versus non-amyloidogenic LL-37 provides a natural test of whether RNA-mediated phase regulation is specific to amyloid-competent assemblies or represents a broader principle of α-helical peptide regulation. Although LL-37 can assemble into α-helical fibrillar structures56,59, these lack the ordered cross-β or cross-α stacking characteristic of amyloids. Nevertheless, RNA strongly modulates LL-37 behavior. RNA can selectively attenuate cytotoxicity while preserving antimicrobial function, correlating with the formation of amorphous RNA–LL-37 co-assemblies rather than structured fibrils (Figs. 8, S11 & S15). RNA-dependent condensate-like assemblies are observed only under thermal stress, and these rapidly adopt solid-like properties (Figs. 8 and S12). For LL-37, RNA therefore appears to serve a protective role, minimizing collateral host damage while maintaining immune defense. This suggests that RNA-mediated modulation is a broader principle of α-helical peptide regulation, yet with a functional divergence that may reflect differences in the architecture of the resulting co-assemblies: RNA drives PSMα3 into structured α-helical fibrillar polymorphs, and LL-37 into disordered, amorphous co-aggregates.

Interestingly, although LL-37 does not adopt canonical amyloid architectures, it displays multiple functional connections to amyloid systems. LL-37 has been reported to bind the Alzheimer’s-associated amyloid-β (Aβ) peptide and inhibit its fibrillization, likely through stabilization of prefibrillar intermediates95. This observation suggests that LL-37 can influence amyloid assembly pathways rather than simply interacting with mature fibrils, with potential implications for aging, immune regulation, and infection. In addition, LL-37 has been shown to interact with human α-synuclein as well as with bacterial curli amyloids, where it modulates biofilm properties96–98. Together, these findings position LL-37 as a cross-reactive regulator of amyloid assemblies across host and microbial contexts. The mechanistic basis and physiological relevance of these interactions remain to be further elucidated.

Parallels with phase-transition biology

The RNA-dependent phase transitions observed for PSMα3 parallel broader principles established in phase-transition biology, where nucleic acids regulate the assembly, material properties, and function of biomolecular condensates. Similar RNA-driven liquid-to-solid transitions have been described for human RNA-binding proteins such as FUS and TDP-43, in which LLPS enables dynamic assembly states that can mature into more solid-like structures associated with pathology76,77,99–110. Antimicrobial peptides have likewise been shown to undergo nucleic-acid-dependent co-assembly and phase separation111, indicating that RNA-mediated control of peptide assembly extends across structurally diverse systems.

Within this framework, the nucleolar accumulation of PSMα3 (Fig. 5) is notable because the nucleolus itself is an RNA-rich phase-separated compartment. The constrained but measurable fluorescence recovery of PSMα3 within nucleoli (Fig. S8) parallels the reduced dynamics observed during aging of RNA–PSMα3 condensates in vitro, although we do not infer a direct mechanistic equivalence between the cellular and in vitro states. Given the rapid membrane disruption induced by PSMα3 at cytotoxic concentrations (Figs. 5–6, Movies S1–S2), we do not interpret nucleolar localization as evidence for a distinct intracellular toxic mechanism. Rather, it reflects the intrinsic nucleic-acid binding capacity of PSMα3 following cellular entry. Whether such interactions become biologically relevant at sub-cytotoxic concentrations, where membrane disruption does not dominate, remains an important question for future investigation.

PSMα3 amyloid-like assembly at the intersection of bacterial virulence and human amyloid biology

The functional relevance of PSMα3 amyloid-like assembly extends beyond bacterial virulence to interactions with human amyloid systems, similarly to LL-37. Its effects are assembly state-dependent: monomeric α-helical PSMα3 inhibits fibrillation of insulin and Aβ40, whereas oligomeric and cross-α fibrillar forms promote Aβ40 aggregation112–114, indicating that supramolecular architecture governs its activity. PSMα3 also binds α-synuclein oligomers with low nanomolar affinity, targeting an oligomer-specific N-terminal motif and inhibiting oligomer-associated toxicity, with no interaction observed for monomeric α-synuclein98,115. This selectivity highlights the emergence of functional surfaces unique to assembled states. Moreover, our data support a broader principle in which supramolecular architecture serves as a dynamic determinant of biological function. Structured fibrillar assemblies are required for LL-37 activity and membrane interactions116, nanoscale organization controls β-defensin immunostimulatory potency70–72, and α-helical assembly states correlate with cytotoxicity across fibril-forming antimicrobial peptides47–51.

Together, these findings argue against PSMα3 assembly as incidental and instead support a model of functionally relevant, assembly-dependent activity. Notably, the RNA-dependent modulation of PSMα3 described here raises the possibility that extracellular nucleic acids in inflammatory environments could regulate its interactions with human amyloids, linking infection to neurodegenerative pathways.

Biological and therapeutic implications

The divergent outcomes of RNA-mediated modulation, namely preservation of PSMα3 virulence versus attenuation of LL-37 cytotoxicity, reflect peptide-specific assembly architectures and suggest that RNA may tilt the host-pathogen balance toward bacterial advantage in infection-relevant environments rich in extracellular nucleic acids. Collectively, the findings position RNA as an environmental regulator of α-helical peptide assemblies, tuning bioactivity by modulating supramolecular organization and reversibility.

Nucleic acids are abundant in infection-relevant settings where PSMs operate, including biofilms enriched in extracellular DNA and RNA, abscesses, neutrophil extracellular traps, and damaged tissues that release host RNA. In such contexts, RNA could control the persistence and timing of membrane-active PSM species by shifting assemblies between condensate-like reservoirs and more inert aggregates, thereby modulating virulence without invoking an intracellular mechanism. This raises the possibility that extracellular RNA can influence the host–pathogen balance in inflammatory milieus by differentially regulating bacterial and host peptide assemblies.

While Poly (AU) served here as a defined, reductionist probe of peptide–RNA interactions, multiple endogenous RNA pools are plausible physiological partners. Within S. aureus biofilms, extracellular RNA is an abundant matrix component that contributes structurally to biofilm organization, raising the possibility that it directly modulates PSM assembly and thereby impacts biofilm stability and virulence-factor availability117,118. Beyond biofilms, amyloid formation by PSMs has been detected in planktonic cultures, where δ-toxin–containing fibrils co-purify with extracellular membrane vesicles from virulent strains21, linking amyloid formation to active secretion pathways. Conversely, the absence of detectable fibrils in certain in vivo infection models21 may reflect the strong concentration dependence and environmental sensitivity of PSM assembly. Host-derived extracellular RNA released from stressed or dying cells can additionally function as a danger-associated molecular pattern119,120 and may further influence the balance between PSMα3-driven cytotoxicity and LL-37-mediated host protection. Defining which endogenous RNA species, concentrations, and structural features govern these assembly-dependent effects in vivo therefore represents an important next step.

More broadly, this work shifts the conceptual focus from amyloid formation as a static endpoint to environmentally regulated phase and assembly transitions. The findings further demonstrate that RNA-mediated regulation extends beyond canonical amyloids to structurally distinct α-helical peptide assemblies. For PSMα3, RNA-sensitive assembly dynamics may enable S. aureus to tune virulence in response to local environmental cues, supporting immune evasion, biofilm-associated persistence, and survival across diverse host niches. For LL-37, RNA-mediated attenuation of cytotoxicity without loss of antimicrobial activity suggests a host-protective mechanism that balances pathogen control with limiting collateral tissue damage. Together, these insights motivate therapeutic strategies aimed at reshaping assembly pathways and material states, rather than simply blocking aggregation.

Methods

Peptide and Poly (AU) RNA preparation

Unlabeled and C-terminal FITC-labeled PSMα3 and LL-37 (PSMα3-FITC and LL-37-FITC) (>98% purity) were purchased from GL Biochem (Shanghai) Ltd. Poly (AU) RNA was purchased from Sigma Aldrich. A stock solution of the peptides was prepared at a concentration of 1 mM in a mixture of 20% dimethyl sulfoxide (DMSO) and 80% ultra-pure water. For light microscopy experiments, we used a mixture of 20% FITC-labeled and 80% unlabeled PSMα3 or LL-37. The Poly (AU) RNA stock solution was prepared at a concentration of 2000 ng/µl in UPW, the stock solution was stored at −80 °C until further use. Labeled Poly (AU) RNA was prepared by introducing 0.02 mg/ml of PI dye to the Poly (AU) RNA stock solution. For the experiments, samples containing 100 µM PSMα3 or LL-37 with Poly (AU) RNA at 10, 50, 100, 200, or 400 ng/µl were prepared in 50 mM HEPES buffer with 150 mM sodium chloride (NaCl), adjusted to pH 7.4, or in 20 μM Tris, 20 μM Bis-Tris, and 20 μM sodium acetate adjusted to pH 4.

Cytotoxicity against HeLa cells tested using the lactate dehydrogenase (LDH) release assay

Human cervical carcinoma HeLa cells (ATCC® CCL-2™) were routinely cultured in Dulbecco’s Modified Eagle’s Medium - high glucose (DMEM) (Sigma, Israel) with L-glutamine and supplemented with penicillin (100 U/ml), streptomycin (0.1 mg/ml), and with 10% foetal calf serum (Sigma, Israel). The cells grown at 37°C and 5% CO2. One day before the experiment, cells were resuspended in growing medium (DMEM supplemented with 10% foetal calf serum) 1 × 105 cells/ml. 50 μL of cell suspension were pipetted into a 96-well plate and grown over night. Thirty minutes before the experiment cells were washed and resuspended in 50 μL of DMEM medium supplemented with L-glutamine (100 U/mL), and with penicillin (100 U/mL), streptomycin (0.1 mg/mL), and with 0.5% heat-inactivated foetal calf serum (assay medium). For the cytotoxicity assay, 100 µM PSMα3 was incubated with or without EGCG at molar ratios of 1:1 or 1:5, or with Poly (AU) RNA at concentrations of 50 ng/µL or 400 ng/µL, for the designated incubation times. Similarly, 100 µM LL-37 was prepared under the same conditions, with or without EGCG at 1:1 or 1:5 molar ratios, or Poly (AU) RNA at 100 ng/µL or 400 ng/µL, following the specified incubation times.

Serial two-fold dilutions in assay medium were performed, and 50 μL of each dilution were pipetted into the three different 96-well with the cells. The plates were incubated for 30 min at 37°C and 5% CO2 and then cell lysis was quantified using the LDH release colorimetric assay according to the manufacturer’s instructions, including all recommended controls (LDH; Cytotoxicity Detection Kit Plus, Roche Applied Science, Germany). Cell-free assay medium was measured as background. Cells subjected to the same experimental conditions apart from peptide addition were used as a control to account for spontaneous LDH release. Cells subjected to the same experimental conditions apart from peptide addition and treated with manufacturer-supplied lysis buffer were used as a control to account for maximum LDH release. Absorbance at 490 and 690 nm was measured in a plate reader (FLUOstar Omega, BMG Labtech, Germany). Absorbance at 690 nm was subtracted from 490 nm readings to correct for background. The mean absorbance of triplicate samples and controls was calculated, followed by background subtraction. Data was obtained from at least three independent biological replicates, with the arithmetic mean used for averaging. Error bars represent the standard deviation (SD). Statistical analysis, including one-way ANOVA, was performed using GraphPad Prism 10.

Assessment of Bacterial Viability Using PrestoBlue™ HS Cell Viability Reagent

Escherichia coli (E. coli) cultures were grown overnight in Luria Broth (LB) medium at 37 °C with shaking at 220 rpm. PSMα3 and LL-37 stock solutions were prepared by dissolving the peptides at a concentration of 1 mM in a solvent mixture of 20% DMSO and 80% ultra-pure water. From these stock solutions, the following sample preparations were made:

For PSMα3: (1)100 µM PSMα3 (2)100 µM PSMα3 with 50 ng/µL Poly (AU) RNA (3) 100 µM PSMα3 with 400 ng/µL Poly (AU) RNA (4) 100 µM PSMα3 with 500 µM EGCG. For LL-37: (1)100 µM LL-37 (2)100 µM LL-37 with 100 ng/µL Poly (AU) RNA (3) 100 µM LL-37 with 400 ng/µL Poly (AU) RNA (4) 100 µM LL-37 with 500 µM EGCG.

All solutions were prepared in 50 mM HEPES buffer pH 7.4 containing 150 mM NaCl. Where applicable, samples were incubated for the specified durations of 2 hours or 24 hours.

On the day of the experiment, dilutions of the PSMα3 solutions to 20 µM and 10 µM were prepared in LB medium and dispensed into a sterile 96-well black flat-bottom plate (Greiner bio-one). E. coli cultures were diluted to an optical density (OD) of 0.2, and bacterial suspensions were added to the wells. The plate was incubated at 37 °C with shaking at 220 rpm for 1 hour to allow the reaction to proceed.

Following incubation, 10X PrestoBlue™ HS Cell Viability Reagent (Invitrogen) was added to the wells. Wells containing only LB medium served as negative controls, while wells containing only E. coli served as positive controls. Plates were sealed with a thermal seal film (EXCEL Scientific) and incubated in a plate reader (CLARIOstar). Bacterial viability was assessed using PrestoBlue fluorescence (excitation: 535–560 nm; emission: 590–615 nm) over time. Each condition was tested in triplicate across three independent experiments. Fluorescence values were averaged, and blank readings were subtracted. Antimicrobial activity was determined at the 2-hour time point, when fluorescence was highest in the positive control. Data were obtained from at least three independent biological replicates, with the arithmetic mean used for averaging. Error bars represent the standard deviation (SD). Statistical analysis, including one-way ANOVA, was performed using GraphPad Prism 10.

Transmission electron microscopy (TEM)

For TEM analysis, 4–5 µL of 100 µM PSMα3, with or without EGCG or Poly (AU) RNA at concentrations of 10 ng/µL, 50 ng/µL, and 400 ng/µL, as well as 100 µM LL-37, with or without Poly (AU) RNA at 100 ng/µL and 400 ng/µL, were directly applied onto glow-discharged 400-mesh copper grids (easiGlow; Pelco, Clovis, CA, USA) with a grid hole size of 42 µm, stabilized with Formvar/carbon (Ted Pella, Inc.). The grids were glow-discharged using a 15-mA current with a negative charge for 25 seconds. Samples were allowed to adhere to the grids for 45 seconds before being stained with a 1% uranyl acetate solution (Electron Microscopy Science, 22400-1) for 45 seconds. The samples were then examined using a ThermoFisher Scientific (FEI) Talos F200C transmission electron microscope, operating at 200 kV and equipped with a Ceta 16M CMOS camera, at the Ilse Katz Institute for Nanoscale Science and Technology, Ben Gurion University of the Negev, Israel.

Turbidity measurements

Turbidity measurements were performed using a FLUOstar Omega plate reader (BMG Labtech) set to a wavelength of 400 nm. For each sample, the protein concentration was maintained at 100 µM, while RNA concentrations were varied across the following levels: 10 ng/µl, 20 ng/µl, 50 ng/µl, 100 ng /µl, 200 ng/µl, and 400 ng/µl. The turbidity of the mixtures was monitored over several hours following resuspension. The maximum absorbance recorded within the first 30 minutes was used as an indicator of the dense phase volume. The experiments were conducted three times with similar observations.

Electrophoretic mobility shift assay (EMSA)

RNA molecules were synthesized by IDT, with oligoA modified by the IRDye 800CW fluorescent dye. Double-stranded RNA was prepared by annealing 10 µM oligoA-IRDye 800CW with 10 µM oligoU in an annealing buffer containing 10 mM Tris-HCl (pH 7.5), 50 mM NaCl, and 1 mM EDTA. The mixture was heated to 95°C for 5 minutes, followed by gradual cooling to 25°C at a rate of 2°C per minute. The final RNA concentration used in the binding reactions was 40 nM (∼400ng/µL).

RNA-protein complexes were incubated at 37°C for 30 minutes before being resolved on a 2.5% agarose gel. Gel electrophoresis was performed in a 0.5× TBE buffer for 7 minutes. The complexes were visualized using a Li-Cor Odyssey FC imaging system in the 800 nm channel with a 30-second exposure. For the EMSA, binding reactions were carried out in a buffer composed of 20 mM HEPES (pH 7.9), 50 mM KCl, 1 mM Dithiothreitol (DTT), 0.1 mM Ethylenediaminetetraacetic acid (EDTA), 5% glycerol, and 0.05% nonyl phenoxypolyethoxylethanol (NP-40). PSMα3 was added to the reactions at final concentrations of 0, 40, 80, 160, and 320 µM. The experiments were conducted three times with similar observations.

Fluorescent microscopy imaging of peptides with RNA in vitro

Prior to imaging, 100 µM PSMα3 or LL-37 (containing 20% FITC-labeled and 80% unlabeled peptide) were combined with PI-labeled Poly (AU) RNA at different concentrations in 50 mM HEPES buffer pH 7.4 containing 150 mM NaCl. Sample were also tested after 65°C heat shock for 15 minutes. In addition, the PSMα3 and RNA mixture was also tested in a different buffer containing 20 μM Tris, 20 μM Bis-Tris, and 20 μM sodium acetate at pH 4 and were subjected to heat shock at 65°C for 15 minutes.

Following preparation, with or without heat shock, a 10 µL aliquot of each sample was transferred to a µ-Slide 8 Well ibidi chamber or to a clean 24 × 60 mm No. 1.5 glass slide for imaging. Fluorescence microscopy was performed using a Leica DMI8 inverted fluorescent microscope equipped with a 63× immersion oil objective (Numerical Aperture, NA = 1.4) at the Life Sciences and Engineering (LS&E) Infrastructure Center, Technion-Israel Institute of Technology, Haifa, Israel. Data were processed and analyzed using Fiji-ImageJ software. The experiment was conducted three times with similar observations.

Fluorescence recovery after photobleaching (FRAP) measurements

FRAP experiments were conducted using a Zeiss LSM 710 laser scanning confocal microscope equipped with a 63× Plan-Apochromat oil immersion objective, NA1.4, at the Life Sciences and Engineering (LS&E) Infrastructure Center, Technion-Israel Institute of Technology, Haifa, Israel. The experiments were performed on condensates from by 100 µM PSMα3 or LL-37 (containing 20% FITC-labeled and 80% unlabeled peptide) mixed with PI-labeled Poly (AU) RNA at various concentrations as indicated in the figures. Photobleaching was performed using 405 nm and 488 nm laser lines. The experiments were conducted three times with similar observations.

For FRAP analysis inside the nucleolus of HeLa cells, PSMα3 was added to the cells at a final concentration of 20 µM containing 20% FITC-labeled and 80% unlabeled peptide. The cells were incubated with the peptide for 20 minutes before conducting the experiment. Data was processed and analyzed using ZEISS ZEN software and Fiji-ImageJ software. The experiments were conducted three times with similar observations.

Total internal reflection fluorescence (TIRF) microscopy

TIRF microscopy was performed using a ZEISS Elyra 7 Super-Resolution Microscope equipped with a 63× Apochromat alpha Plan-Apochromat Oil immersion DIC objective, NA1.46, at the Life Sciences and Engineering (LS&E) Infrastructure Center, Technion-Israel Institute of Technology, Haifa, Israel. Images were acquired using the pco.edge sCMOS cameras, and data were processed and analyzed with ZEISS ZEN software to accurately represent the observed phenomena. The condensates and aggregates analyzed were generated using 100 µM of 20% FITC-PSMα3 (green) and 80% unlabeled PSMα3 with either 50 ng/µL or 400 ng/µL Poly (AU) RNA. AmyTracker 630 (Ebba Biotech) was added at a 1:500 (peptide: AmyTracker) molar ratio to each sample. The experiments were conducted three times with similar observations.

Confocal microscopy visualization of PSMα3 interaction with HeLa cells

HeLa cells were pre-cultured one day before the experiment by preparing a suspension containing 350,000 cells/ml and plating 150 µl of this suspension into each well of a µ-Slide 8 well glass-bottom chamber. The cells were then incubated overnight under standard growth conditions (37°C, 5% CO2) to allow for adherence and growth. On the day of the experiment, the cells were washed three times with phosphate-buffered saline (PBS) to remove any residual media. Hoechst 33342 dye (10 mg/mL stock) was diluted 1:2000 in fresh cell media and added to the cells. The cells were incubated with Hoechst for 10 minutes at 37°C and 5% CO2. After incubation, cells were washed three times with PBS to remove Hoechst residuals. A working solution of PI was prepared by diluting a 1 mg/ml PI stock solution to a final concentration of 0.02 mg/ml in fresh cell growth media. Immediately prior to imaging, PSMα3 was added to the cells at a final concentration of 20 µM containing 20% FITC-labeled and 80% unlabeled peptide. The cells were then imaged using a Ti2-E microscope by Nikon with a CSU-W1 spinning disk confocal unit by Yokogawa, equipped with a 100X CFI SR HP Plan Apochromat Lambda S silicone immersion objective, NA1.35, at the Life Sciences and Engineering (LS&E) Infrastructure Center, Technion-Israel Institute of Technology, Haifa, Israel. Confocal movies and images were captured by Photometrics BSI sCMOS cameras to observe the interaction and localization of PSMα3 within the cells, focusing particularly on its colocalization with nucleic acids in the nucleolar region. The acquired images and movies were subsequently analyzed using Imaris Image Analysis Software (Oxford Instruments). The experiments were conducted three times with similar observations.

Super resolution light microscopy visualization of PSMα3 interaction with E. coli

E. coli lptD4213 cells were a kind gift from Prof. Sima Yaron from the Technion – Israel Institute of Technology. An inoculum was pre-cultured in Luria-Bertani medium (LB) medium at 220 rpm and 37°C for 24 hours prior to the experiment. On the day of the experiment, the optical density (OD) of the culture was measured, and the cells were diluted to an OD of 0.4. The cells were then incubated with 20 μM of 20% FITC-PSMα3 (green) and 80% unlabeled PSMα3, with and without EGCG at 1:5 molar ratio, alongside a bacteria-only control. Samples were incubated for 30 minutes at 220 rpm and 37°C. Following incubation, the samples were centrifuged three times for 3 minutes at 1.5g, with the supernatant discarded and the pellet resuspended in 1X PBS after each spin. The resuspended cells were then stained with 50 μg/mL wheat germ agglutinin (WGA) CF633 for 30 minutes. After staining, the cells were washed three times by centrifugation with PBS to remove excess WGA. During the final wash, the cells were resuspended in 0.1 mg/mL PI solution.

Prior to imaging, the samples were loaded into an ibidi µ-Slide VI 0.4. Imaging was performed using a ZEISS Elyra 7 Super-Resolution Microscope equipped with a 63× Apochromat alpha Plan-Apochromat Oil immersion DIC objective, NA 1.4, and pco.edge sCMOS cameras. Lattice structured illumination images were acquired and processed using the ZEN black software at the Life Sciences and Engineering (LS&E) Infrastructure Center, Technion-Israel Institute of Technology, Haifa, Israel. The experiments were conducted three times with similar observations.

Thioflavin T fluorescence fibrillation kinetics assay

Thioflavin T (ThT) (Sigma Aldrich) is a widely used fluorescent dye for detecting and analyzing amyloid fibril formation kinetics. Fibrillation curves in the presence of ThT typically exhibit an initial lag phase followed by rapid fibril elongation. To accurately capture the fibrillation lag time, PSMα3 was pre-treated before the experiment. The peptide was dissolved in a 1:1 mixture of 1,1,1,3,3,3-Hexafluoroisopropanol (HFIP- Sigma Aldrich) and Trifluoroacetic acid (TFA- Sigma Aldrich) to a final concentration of 1 mg/mL, followed by bath sonication for 10 minutes at room temperature. The organic solvents were then evaporated using a mini-rotational vacuum concentrator (Christ, Germany) at 1,000 rpm for 2 hours at room temperature.

For the experiment, freshly prepared 100 µM PSMα3 peptides, with or without 500 µM Epigallocatechin gallate (EGCG) at a 1:5 molar ratio, were prepared in 50 mM HEPES buffer pH 7.4 containing 150 mM NaCl. ThT was prepared by diluting a stock solution in ultrapure double-distilled water (UPddw) and filtered before use to reach a final concentration of 200 µM Blank control solutions containing all components except for the peptide were prepared for each reaction.

The assay was conducted in black 96-well flat-bottom plates (Greiner Bio-One), which were sealed with a thermal seal film (EXCEL Scientific) to prevent evaporation. Samples were incubated in a plate reader (OMEGA) at 37°C, shaking at 500 rpm for 85 seconds before each reading cycle, for up to 1,000 cycles of 6 minutes each, totaling approximately 100 hours.

Fluorescence was measured in triplicate using an excitation wavelength of 438 ± 20 nm and an emission wavelength of 490 ± 20 nm. All values were averaged, background fluorescence was subtracted using blank controls, and the results were plotted over time. Standard error of the mean is shown as error bars. The entire experiment was independently repeated at least three times on different days to ensure reproducibility.

Solid-State Circular Dichroism Spectroscopy

Solid-state circular dichroism spectroscopy(ss-CD) was conducted with PSMa3 to assess their fibrillar secondary structure components in presence and absence of Poly (AU) RNA. For preparation of peptide and RNA samples, see Peptide and Poly (AU) preparation section. PSMa3 was prepared at a working concentration of 100 µM in 50 mM HEPES, pH 7.4, 150 mM NaCl at 0 ng/µL, 50 ng/µL or 400 ng/µL Poly (AU) RNA. The reaction mixes were incubated for 0 min, 2 h and 72 h at 37 °C in a non-shaking thermocycler. Following incubation, soluble reaction components were removed via dilution washing with 150 µL water and centrifugation at 12.500 rcf for 30 min. 20 µL of the sediment were resuspended in 180 µL water for a second washing centrifugation step at 12.500 rcf for 30 min. 180 µL of the supernatant were discarded and 18 µL of sediment applied to a Chirascan Series fused silica disc (AP/CSSD, Applied Photophysics) in three times 6 µL steps. After every sample application onto to the silica disc, all liquid was evaporated at 37 °C for 5–10 min until only an opaque film remained visible. For secondary structure determination of aggregated/fibrillar peptide components on the silica disc, the disc was inserted into a Chirascan solid sample holder (CS/SSH, Applied Photophysics) and the circular dichroism recorded in the far-UV (180–250 nm, step size 1 nm, time-per-point 1 s) using a CD spectropolarimeter (Applied Photophysics). To counteract sample anisotropy, CD spectra were recorded at least four different disc rotations and the results averaged. Prior to data analysis, silica disc and buffer/water backgrounds were subtracted from recorded sample spectra. Poly (AU) RNA-only background spectra showed complex and high amplitude signals that did not appear for samples in combination with peptides and were therefore not subtracted from sample spectra.

NMR experiments

To prepare the reference samples for NMR analysis, 0.521 mg of PSMα3 was dissolved in 180 µL of 20 mM MES (2-(N-morpholino)ethanesulfonic acid) buffer at pH 6.0, supplemented with 5% (v/v) deuterium oxide (D₂O). From this solution, 170 µL was transferred into a 3.0 mm NMR tube for spectral acquisition. In parallel, a 4 mM stock solution of EGCG was prepared in the same 20 mM MES buffer (pH 6.0). From this, a working solution containing 0.5 mM EGCG in 20 mM MES with 5% (v/v) D₂O was prepared to match the buffer conditions of the peptide sample. A volume of 170 µL was then transferred into a separate 3.0 mm NMR tube for use as the EGCG reference.

All NMR experiments were recorded on a Bruker 700 MHz spectrometer equipped with a triple-resonance cryoprobe (Prodigy) and an AVANCE NEO console. One-dimensional (1D) ¹H-NMR spectra were acquired using the Bruker library pulse sequence zggpw5 for water suppression. The acquisition parameters were: acquisition time (AQ) = 1.90 s, spectral width (SW) = 12.31 ppm, recovery delay (D1) = 1.5 s, and 64 scans. All experiments were conducted at 35 °C, under which optimal spectral quality was obtained.

Two-dimensional (2D) ¹H–¹H TOCSY (Total Correlation Spectroscopy) spectra were collected using the Bruker pulse sequence dipsi2gpph19. Acquisition parameters were: t₂ = 2048 complex points (AQ₂ = 0.118 s), t₁ = 512 complex points (AQ₁ = 0.029 s), SW₁ = SW₂ = 12.31 ppm, with 64 scans and a mixing time (tmix) of 80 ms. Recovery delay (D1) was set to 1.5 s. 2D ¹H–¹H NOESY (Nuclear Overhauser Effect Spectroscopy) spectra were acquired using the noesyfpgpph19 pulse sequence, with acquisition parameters identical to those used for TOCSY. The mixing time was set to 300 ms. Total experimental time for the combined 2D TOCSY and NOESY experiments was approximately 30 hours. NMR data processing was carried out using TopSpin 4.5.0 (Bruker), and spectral assignments for the 2D TOCSY and NOESY spectra were performed using POKY software68. Signal integration was carried out using in-house scripts written in Scilab69.